---
title: "Manuscript_species_level"
output: html_document
date: "2026-08-27"
editor_options: 
  chunk_output_type: console
---

#library
```{r}
library(readr)
 library(here)
#library(tidyverse)   
# library(readxl)
library(ggpubr)
library(funrar)
 library(dplyr)
 library(tidyr)
# library(stringr)
 library(ggplot2)
 library(TNRS) #c'est une application, pas une base de données
 library(stargazer)
library(ggeffects)
library(FactoMineR)
library(factoextra)
library(stringr)
library(tibble)
library(purrr)
library(lme4)
library(emmeans)
library(lmerTest) #for confint
library(betapart) #for turonver
library(patchwork) #to combine plots
library(MuMIn)#Lucie: conseil pour choisir les modèles. le package s'appelle multi-model inference, la fonction dredge va selectionner de manière automatique des subset models depuis le modele le plus general donné en entrée.
library(sjPlot)
library(jtools)#for summ function
library(mice)
library(VIM)
library(gt)
library(kableExtra)
```

#Data
```{r}

data <- read_csv("intermediate data/site_species_abs_abund.csv")
names(data)

#need to pivot in long format
data_long <- data %>% pivot_longer(cols=-c(Region, originSiteID, destSiteID, destPlotID, Year, Treatment), names_to = "Species",values_to = "Abs_Cover")

dim(data_long)
head(data_long)

data_long %>%
  group_by(Region) %>%
  summarise(
    n_years = n_distinct(Year),
    year_min= min(Year),
    year_max= max(Year)
  )

# add variable elevation and plotSize
metadata <- read.csv(here("intermediate data","metadata.csv"))

data_long <- data_long %>% left_join(metadata %>% select(Region, destSiteID, Elevation, PlotSize)) %>%
  relocate(PlotSize, .after= destPlotID)

##add temperature
load("./raw data/TransPlantNetwork.RData")
climdata <- climdata %>% filter(month %in% c("6","7","8")) #average of hottest months 

temperature <- climdata %>%
  filter(year %in% sample(2007:2022)) %>%
  group_by(gradient, elevation, year) %>%
  filter(var == "temperature")%>%
   mutate(elevation = ifelse(gradient == "CH_Calanda" & elevation == 2800, 2600, elevation))%>% # Correction to Calanda experiment
  rename(Region = gradient) 

CH_Calanda_temp <- temperature%>%
  filter(Region == "CH_Calanda")%>%
  group_by(Region, elevation, year)%>%
  filter(year %in% sample(YearMin:YearMax))%>%
  summarise(mean_temperature = mean(value), .groups = "drop")%>%
  pivot_wider(names_from = year , values_from = mean_temperature)%>%
  mutate(row_mean = rowMeans(select(., -c(1:2)), na.rm = TRUE))


survey_years <- data_long %>%
  group_by(Region) %>%
  summarise(year_min = min(Year, na.rm = TRUE),
            year_max = max(Year, na.rm = TRUE),
            .groups = "drop")

region <- unique(temperature$Region) 
region <- region[region!="IN_Kashmir"]
region <- region[region!="US_Arizona"]
region <- region[region!="US_Colorado"]

temp_data <- data.frame()

for (i in region) {
  
  # get the survey year range for this region
  yr_min <- survey_years %>% filter(Region == i) %>% pull(year_min)
  yr_max <- survey_years %>% filter(Region == i) %>% pull(year_max)
  
  temp_results <- temperature%>%
    filter(Region == i)%>%
    filter(year >= yr_min & year <= yr_max) %>%
    group_by(Region,elevation, year)%>%
    summarise(mean_temperature = mean(value),.groups = "drop")%>%
    pivot_wider(names_from = year , values_from = mean_temperature)%>%
    mutate(temp_mean = rowMeans(select(., -c(1:2)), na.rm = TRUE))%>%
    rename(Elevation = elevation)%>%
    select(Region, Elevation, temp_mean) 
  
  temp_data<- rbind(temp_data, temp_results)
  
}


site_elevation <- data_long %>%
  distinct(Region, destSiteID, Elevation) %>%
  left_join(temp_data)%>%
  rename(SiteID = destSiteID)

data<- data_long%>%
  left_join(site_elevation, by = c("Region","Elevation")) 

```


#Colonisatrices & seuil
```{r}

##calculer les transitions annuelles 
#utiliser lead() pour aller au prochain
annual_transition <- 
  data_long %>%
  arrange(Region, originSiteID, destSiteID, destPlotID, Treatment, Species, Year) %>%
  group_by(Region, originSiteID, destSiteID, destPlotID, Treatment, Species) %>% #pas avec Year sinon ca ne marche pas
  mutate(
    Abs_Cover_next = lead(Abs_Cover),
    Year_next = lead(Year),
    Delta_Years = Year_next - Year) %>%
  filter(!is.na(Delta_Years))
  

k <- 5

annual_transition <- annual_transition %>%
  mutate(
    dynamique_statut = case_when(
      Abs_Cover == 0 & Abs_Cover_next > 0 ~ "Colonizer",
      Abs_Cover > 0  & Abs_Cover_next == 0 ~ "Extinct",
      Abs_Cover > 0  & Abs_Cover_next > 0  ~ "Persistent",
      TRUE                                  ~ NA_character_
    )
  )  %>%
  filter(!is.na(dynamique_statut))

# Seuil de détectabilité par site
annual_transition2 <- annual_transition %>%
  group_by(Region, Treatment, originSiteID,
           destSiteID, Year) %>%
  mutate(
    min_cover_nonzero   = min(Abs_Cover_next[Abs_Cover_next > 0],
                              na.rm = TRUE),
    detection_threshold = min_cover_nonzero / k,
    Abs_Cover_adj       = case_when(
      dynamique_statut == "Colonizer" ~ detection_threshold,
      TRUE                            ~ Abs_Cover),
    logN = log10(Abs_Cover_adj)) %>%
  ###
  mutate(
    min_cover_nonzero2 =min(Abs_Cover[Abs_Cover > 0],
                              na.rm = TRUE),
    detection_threshold2 = min_cover_nonzero2 /k, 
    Abs_Cover_next_adj = case_when(
      dynamique_statut == "Extinct" ~detection_threshold2, 
      TRUE ~Abs_Cover_next), 
    `logN+1` = log10(Abs_Cover_next_adj))%>%
  ungroup() %>%
  filter(
    dynamique_statut %in% c("Persistent", "Colonizer", "Extinct"),
    is.finite(logN))

# Calculer r par espèce 
annual_transition3 <- annual_transition2 %>%
  mutate(
    r = log10(Abs_Cover_next_adj / Abs_Cover_adj)
  ) %>%
  filter(is.finite(r)) #removes exctinct otherwise log = -inf, non en fait

# Vérif
dim(annual_transition3)
annual_transition3 %>%
  count(dynamique_statut)

# summary stats
annual_transition3 %>%
  group_by(dynamique_statut)%>%
  summarise(
    mean_r = mean(r),
    sd_r= sd(r, na.rm=T),
    n = n())


```

#Table1 -summary per region
```{r}
table1 <- annual_transition3 %>%
   left_join(temp_data)%>%
  group_by(Region) %>%
  summarise(
    n_sites       = n_distinct(destSiteID),
    n_plots       = n_distinct(destPlotID),
    n_species     = n_distinct(Species),
    year_min      = min(Year,      na.rm = TRUE),
    year_max      = max(Year_next, na.rm = TRUE),
    elev_min      = min(Elevation, na.rm = TRUE),
    elev_max      = max(Elevation, na.rm = TRUE),
    temp_min      = min(temp_mean, na.rm = TRUE),
    temp_max      = max(temp_mean, na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  mutate(
    year_range = paste0(year_min, "–", year_max),
    elev_range = paste0(elev_min, "–", elev_max, " m"),
    temp_range = paste0(round(temp_min, 1), "–",
                        round(temp_max, 1), " °C")
  ) %>%
  select(Region, n_sites, n_plots, n_species,
         year_range, elev_range, temp_range)

table1  %>%
  kable(booktabs = TRUE, align = "lllllc", escape = FALSE, format = "html") %>%
  row_spec(0, bold = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria", font_size = 16) %>%
  save_kable(file = "Figures/Table1.png", zoom = 10) 

#elevation
annual_transition3 %>%
  group_by(Region) %>%
  summarise(elev_min = min(Elevation, na.rm = TRUE),
            elev_max = max(Elevation, na.rm = TRUE),
            .groups  = "drop") %>%
  ggplot(aes(y = reorder(Region, elev_max))) +
  geom_segment(aes(x    = elev_min, xend = elev_max,
                   yend = reorder(Region, elev_max)),
               linewidth = 2, colour = "steelblue") +
  geom_point(aes(x = elev_min), size = 2,
             colour = "dodgerblue") +
  geom_point(aes(x = elev_max), size = 2,
             colour = "firebrick1") +
  labs(x = "Elevation (m)", y = NULL,
       title = "Elevation range per region") +
  theme_minimal(base_size = 11)
```

#Table2 - species info
```{r}

annual_transition3 %>%
  group_by(Region, destSiteID, destPlotID, Year) %>%
  summarise(n_sp_plot = n_distinct(Species), .groups = "drop") %>%
  group_by(Region) %>%
  summarise(
    mean_sp_per_plot = round(mean(n_sp_plot), 1),
    sd_sp_per_plot   = round(sd(n_sp_plot),   1),
    .groups          = "drop"
  )


#nb species per region per treatment
annual_transition3 %>%
  group_by(Region, Treatment) %>%
  summarise(
    n_species = n_distinct(Species),
    .groups   = "drop"
  ) %>%
  pivot_wider(names_from  = Treatment,
              values_from = n_species)

#Dynamic status counts 

table2 <- annual_transition3 %>%
  count(Region, Treatment, dynamique_statut) %>%
  pivot_wider(names_from  = dynamique_statut,
              values_from = n,
              values_fill = 0) %>%
  mutate(Total = Persistent + Colonizer + Extinct)

print(table2)

# Total across all regions
annual_transition3 %>%
  count(dynamique_statut) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

table5 <-annual_transition3 %>%
  filter(Abs_Cover > 0) %>%
  group_by(Species) %>%
  summarise(
    mean_cover = mean(Abs_Cover,   na.rm = TRUE),
    max_cover  = max(Abs_Cover,    na.rm = TRUE),
    n_obs      = n(),
    regions    = paste(unique(Region), collapse = ", "),
    .groups    = "drop"
  ) %>%
  arrange(desc(mean_cover))

cat("Most abundant:\n")
print(head(table5, 5))
cat("\nLeast abundant (present species):\n")
print(tail(table5, 5))


#plant status
annual_transition3 %>%
  count(Region, dynamique_statut) %>%
  group_by(Region) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(dynamique_statut = factor(dynamique_statut,
                                   levels = c("Colonizer",
                                              "Persistent",
                                              "Extinct"))) %>%
  ggplot(aes(x = reorder(Region, pct),
             y = pct, fill = dynamique_statut)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Colonizer"  = "#A8DCAB",
    "Persistent" = "#FFC5D3",
    "Extinct"    = "#4D0135"
  )) +
  coord_flip() +
  labs(x = NULL, y = "% transitions",
       fill = NULL,
       title = "Dynamic status per region") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

#Gompertz par espèce
```{r}
# La condition var(logN) == 0 signifie :
# "est-ce que toutes les valeurs de logN sont identiques ?"

# Exemple :
# logN = c(0.301, 0.301, 0.301, 0.301, 0.301)
# var = 0 → lm() ne peut pas estimer une pente → NA

# Pour estimer r ~ logN, il FAUT que logN varie
# Si logN ne varie pas, on ne peut pas distinguer
# si r change à cause de N ou d'autre chose

species_gompertz <- annual_transition3 %>%
  group_by(Region, Treatment, Species) %>%
  group_modify(~ {
    df <- .x
    if (nrow(df) < 3 | var(df$logN, na.rm = TRUE) == 0) { #au moins 3 points de transition (sinon pas beaucoup de données, mais on peut changer à 5)
      return(tibble(
        a_sp = NA_real_,
        c_sp = NA_real_,
        n    = nrow(df)
      ))
    }

    mod <- lm(r ~ logN, data = df)
    tibble(
      a_sp = coef(mod)["(Intercept)"],
      c_sp = coef(mod)["logN"],
      n    = nrow(df)
    )
  }) %>%
  ungroup()


species_gompertz %>%
  filter(!is.na(a_sp)) %>%
   filter(!is.na(c_sp)) %>%
  group_by(Treatment) %>%
  summarise(
    n_species    = n(),
    mean_n       = mean(n))

# Combien de transitions par groupe en moyenne ?
species_gompertz %>%
  filter(!is.na(a_sp)) %>%
  summarise(
    mean_n   = mean(n),
    median_n = median(n),
    min_n    = min(n),
    pct_5plus = mean(n >= 5) * 100
  )

```

*Colonnes clés* :
- r       : taux de croissance observé (varie par espèce × plot) --> maintenant r pour chaque transition annuelle. donc on a plusieurs r par espèce par traitement
- a       : intercept Gompertz (CONSTANT par site × traitement) --> maintenant a pour chaque espèce par traitement et region
- c       : pente Gompertz (CONSTANT par site × traitement) --> maintenant a pour chaque espèce par traitement et region
- epsilon : résidu (varie par espèce × plot) --> comme r, plusieurs epsilon par espèce, un epsilon a chaque transition annuelle. calcul al reves en utilisant a et c et r. 
- logN    : log(abondance initiale ajustée)

#get mean r and mean epsilon per species per treatment
```{r}

r_data <- # summary stats
annual_transition3 %>%
  group_by(Region, Species, Treatment)%>%
  summarise(
    mean_r = mean(r))

#epsilon
epsilon_gompertz <- annual_transition3 %>%
  left_join(
    species_gompertz %>% select(Region, Treatment, Species, a_sp, c_sp),
    by = c("Region", "Treatment", "Species")
  ) %>%
  filter(!is.na(a_sp)) %>%
  mutate(epsilon = r - (a_sp + c_sp * logN)) %>% 
  filter(is.finite(epsilon))

#group per species per treatment
epsilon_data <- epsilon_gompertz%>%
  group_by(Region, Treatment, Species)%>%
  summarise(
    mean_epsilon = mean(epsilon, na.rm = TRUE)) #parce que en fait après dans les modeles on va pas pouvoir fitter, c'est très moche, car methomatiquement epsilon est centr& autour de zero, qqplot tres moche, donc on veut fitter la dispersion donc var de epsilon. et je prendre log10 car sinon trop asymétrique la distribution.
  

#combine all
#elevation data
elevation_data <- annual_transition3 %>%
  group_by(Region, Treatment, Species) %>%
  summarise(
    mean_elevation = mean(Elevation, na.rm = TRUE),
    .groups = "drop")

demo_data <- species_gompertz %>%
  left_join(r_data, by = c('Region', 'Treatment', 'Species'))%>%
  left_join(epsilon_data, by = c('Region', 'Treatment', 'Species')) %>%
  left_join(elevation_data, by = c("Region", "Treatment", "Species"))


write.csv(demo_data, "demography_data.csv")

names(demo_data)
```

#Models M0

M0 : parametre démographique ~Traitement + (1/region) + (1/espèce) + elevation
Faire dredge et garder que les variables qui restent pour passer à l’étape suivante sur lequel on rajoute les traits.

```{r}
#scale elevation
demo_data_scaled <- demo_data %>% mutate(Elevation_sc = as.numeric(scale(mean_elevation)))
#remove na
demo_data_scaled_clean <- demo_data_scaled %>%
  filter(!is.na(a_sp), !is.na(c_sp),
         !is.na(Elevation_sc), !is.na(Treatment))

demo_data_scaled_clean %>%
  summarise(
    n_total       = n(),
    n_na_mean_r   = sum(is.na(mean_r)),
    n_na_mean_eps = sum(is.na(mean_epsilon)),
    n_na_a        = sum(is.na(a_sp)),
    n_na_c        = sum(is.na(c_sp))
  )

#summary dela distribution des valeurs extrêmes
summary(demo_data_scaled_clean$a_sp)
summary(demo_data_scaled_clean$c_sp)

#remove above 3 SD par ce que sinon les qqplot sont très moches, il y a une structure.
demo_data_scaled_clean <- demo_data_scaled_clean %>%
  mutate(
    a_sp_z = scale(a_sp),
    c_sp_z = scale(c_sp)
  ) %>%
  filter(abs(a_sp_z) < 3, abs(c_sp_z) < 3) %>%
  dplyr::select(-a_sp_z, -c_sp_z)

nrow(demo_data_scaled) - nrow(demo_data_scaled_clean) #299 removed

# Relevel Treatment avec Control_Origin comme référence
demo_data_scaled_clean <- demo_data_scaled_clean %>%
  mutate(Treatment = relevel(factor(Treatment),
                             ref = "Control_Origin"))

levels(demo_data_scaled_clean$Treatment)

#r
M0_r <- lmer(mean_r ~Treatment + (1|Region) + (1|Species) + Elevation_sc, 
             data=demo_data_scaled_clean, 
             na.action ="na.fail", 
             REML = F) #rajouter na.action et REML = F pour dredge)
summary(M0_r)
dd_select_r <- dredge(M0_r,  evaluate = TRUE,  rank = "AICc", fixed    = "Treatment") #on garde tjs treatment dans le model
nrow(dd_select_r[dd_select_r$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M0_r <- get.models(dd_select_r, subset =1) [[1]] 
best_M0_r

#a
M0_a <- lmer(a_sp ~Treatment + (1|Region) + (1|Species) + Elevation_sc, data=demo_data_scaled_clean,na.action ="na.fail", REML = F) 
summary(M0_a)
dd_select_a <- dredge(M0_a,  evaluate = TRUE,  rank = "AICc",, fixed    = "Treatment")
nrow(dd_select_a[dd_select_a$delta<2,]) 
best_M0_a <- get.models(dd_select_a, subset =1) [[1]] 
best_M0_a 

#c
M0_c <- lmer(c_sp ~Treatment + (1|Region) + (1|Species) + Elevation_sc, data=demo_data_scaled_clean, na.action ="na.fail", REML = F) 
summary(M0_c)
dd_select_c <- dredge(M0_c,  evaluate = TRUE,  rank = "AICc", fixed    = "Treatment")
nrow(dd_select_c[dd_select_c$delta<2,]) 
best_M0_c <- get.models(dd_select_c, subset =1) [[1]] 
best_M0_c 

#epsilon
#demo_data_scaled_clean <-demo_data_scaled_clean %>% filter(!is.na(mean_epsilon))

M0_epsilon <- lmer(mean_epsilon ~Treatment + (1|Region) + (1|Species)  + Elevation_sc, data=demo_data_scaled_clean, na.action ="na.fail", REML = F) 
summary(M0_epsilon)
dd_select_epsilon <- dredge(M0_epsilon,  evaluate = TRUE,  rank = "AICc", fixed    = "Treatment")
nrow(dd_select_epsilon[dd_select_epsilon$delta<2,]) 
best_M0_epsilon <- get.models(dd_select_epsilon, subset =1) [[1]] 
best_M0_epsilon

#conditions vérifieée? 
#normalité des résidus?
par(mfrow = c(2, 2))
qqnorm(resid(best_M0_r),       main = "r");       qqline(resid(best_M0_r))
qqnorm(resid(best_M0_a),       main = "a");       qqline(resid(best_M0_a))
qqnorm(resid(best_M0_c),       main = "c");       qqline(resid(best_M0_c))
qqnorm(resid(best_M0_epsilon), main = "epsilon"); qqline(resid(best_M0_epsilon))

#homoscédasticité
par(mfrow=c(2,2))
plot(fitted(best_M0_r), resid(best_M0_r), main = "r")
plot(fitted(best_M0_a), residuals(best_M0_a), main ="a")
plot(fitted(best_M0_c), resid(best_M0_c), main = "c")
plot(fitted(best_M0_epsilon), resid(best_M0_epsilon), main = "epsilon")



```

#Traits

*new TRy request data*
no imputation 

##load trait data

```{r}
# 1. Load TNRS Correspondence Table
# -------------------------- #
tnrs_df <- read.csv("TNRS_Correspondance_marion.csv") %>%
  dplyr::select(-Name_submitted, -name_species) 

# 2. Load and Prepare Trait Matrix
# -------------------------- #
trait_matrix <- read_csv("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/TRY/TRY_post_TNRS_Transplant.csv") %>%
  filter(!(is.na(Species) & is.na(Genus) & is.na(Family))) %>%
  dplyr::select(-AccSpeciesID, -AccSpeciesName)

# Identify trait columns (i.e., non-taxonomic)
trait_cols <- setdiff(names(trait_matrix), c("Species", "Genus", "Family"))

# 3. Add Missing Species from TNRS
# -------------------------- #
# Identify species present in TNRS but missing in the trait matrix
missing_species <- tnrs_df %>%
  filter(!Species %in% trait_matrix$Species) %>%
  select(Species, Genus, Family)

length(unique(missing_species$Species)) #195 species in TNRS that do not have trait data in TRY

# Create rows with NA traits for these missing species
missing_species_filled <- missing_species %>%
  mutate(across(everything(), as.character)) %>%
  bind_cols(as.data.frame(matrix(NA, nrow = nrow(missing_species), ncol = length(trait_cols))))

colnames(missing_species_filled) <- c("Species", "Genus", "Family", trait_cols)

# Append missing species to the trait matrix
trait_matrix_full <- bind_rows(trait_matrix, missing_species_filled)

# Save a copy of the pre-imputation matrix
trait_matrix_pre_impute <- trait_matrix_full %>% #dim41729 (trait_matrix, missing_species_filled)
  distinct(Species, .keep_all = TRUE) %>% #dim 39902 (39587+315)
  filter(Species %in% tnrs_df$Species) #Only keep species that were in our TNRS. 

#same as length(unique(tnrs_df$name_species))

# -------------------------- #
# 4. Impute Missing Trait Values
# -------------------------- #
# Rename trait columns with "trait_" prefix for clarity
names(trait_matrix_full)[4:8] <- paste0("trait_", names(trait_matrix_full)[4:8])

#(marion) find out where are the missing values
summary(trait_matrix_full)

pMiss <- function(x){sum(is.na(x))/length(x)*100}
apply(trait_matrix_full,2,pMiss) #apply function to (2) columns (missing value per column) c'est énorme!!!
apply(trait_matrix_full,1,pMiss) #apply function to (1) rows

md.pattern(trait_matrix_full)
colnames(trait_matrix_full) <- gsub("^trait_", "", colnames(trait_matrix_full)) #remove trait
trait_matrix_full <- trait_matrix_full

# Plot Missing data per variable & pattern
aggr(
  trait_matrix_full,
  col = c("navyblue", "red"),
  numbers = TRUE,
  sortVars = TRUE,
  labels = colnames(trait_matrix_full),
  cex.axis = 0.7,
  gap = 3,
  combined = FALSE,
  ylab = c("Missing values per variable", "Missing data pattern")
)

#Do the same for TNRS dataset 
colnames(trait_matrix_pre_impute) <- gsub("^trait_", "", colnames(trait_matrix_pre_impute)) #remove trait
trait_matrix_pre_impute <- trait_matrix_pre_impute 

apply(trait_matrix_pre_impute,2,pMiss) #apply function to (2) columns (missing value per column) still suuper big!
apply(trait_matrix_pre_impute,1,pMiss) #apply function to (1) rows

# Plot Missing data per variable & pattern on TNRS dataset
aggr(
  trait_matrix_pre_impute,
  col = c("navyblue", "red"),
  numbers = TRUE,
  sortVars = TRUE,
  labels = colnames(trait_matrix_full),
  cex.axis = 0.7,
  gap = 3,
  combined = FALSE,
  ylab = c("Missing values per variable", "Missing data pattern")
)


#Back to Lula's script ---------------------------------

# Impute missing values using Genus mean first, then Family mean
trait_matrix_imputed <- trait_matrix_full %>%
  group_by(Genus) %>%
  mutate(across(starts_with("trait"), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  ungroup() %>%
  group_by(Family) %>%
  mutate(across(starts_with("trait"), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  ungroup()

# -------------------------- #
# 5. Remove Species Duplicates and Filter to Reference List
# -------------------------- #
trait_matrix_imputed <- trait_matrix_imputed %>%
  distinct(Species, .keep_all = TRUE)

trait_matrix_final <- trait_matrix_imputed %>%
  filter(Species %in% tnrs_df$Species)

# -------------------------- #
# 6. Summary Table of Missing Values
# -------------------------- #
# Count NA values before and after imputation
na_counts_pre <- colSums(is.na(trait_matrix_pre_impute[, 4:8]))
na_counts_post <- colSums(is.na(trait_matrix_final[, 4:8]))

# Count rows where *all* traits are NA
n_all_na_pre <- sum(apply(trait_matrix_pre_impute[, 4:8], 1, function(x) all(is.na(x))))
n_all_na_post <- sum(apply(trait_matrix_final[, 4:8], 1, function(x) all(is.na(x))))

# Create summary table
na_summary <- data.frame(
  Trait = names(na_counts_pre),
  NA_pre_imputation = as.integer(na_counts_pre),
  NA_post_imputation = as.integer(na_counts_post)
)

# Add row for "fully empty rows"
na_summary <- rbind(
  na_summary,
  data.frame(
    Trait = "Rows with all traits missing",
    NA_pre_imputation = n_all_na_pre,
    NA_post_imputation = n_all_na_post
  )
)

# Display with gt
na_summary %>%
  gt() %>%
  tab_header(title = "Summary of Missing Trait Values") %>%
  fmt_number(columns = c(NA_pre_imputation, NA_post_imputation), decimals = 0)

# -------------------------- #
# 7. Final Imputation with Column Means (Fallback)
# -------------------------- #
trait_matrix_final <- trait_matrix_final %>%
  mutate(across(4:8, ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# -------------------------- #
# 8. Save Outputs
# -------------------------- #
write.csv(trait_matrix_final, file = "new_TRY_imputed_traits_mean.csv", row.names = FALSE)
write.csv(trait_matrix_pre_impute, file = "new_TRY_raw_traits_with_na.csv", row.names = FALSE)

```

## trait preparation and transformation
```{r}

# Load imputed mean trait dataset (TRY database)
Sp_trait_TRY <- read.csv("new_TRY_imputed_traits_mean.csv", sep = ",", dec = ".") 

#Sp_trait_TRY <- read.csv("new_TRY_raw_traits_with_na.csv", sep = ",", dec = ".") 


# Log-transform continuous traits to normalize distributions
Sp_trait_TRY_trans <- Sp_trait_TRY

Sp_trait_TRY_trans[,4] <- log10(Sp_trait_TRY_trans[,4])    # Maximum height
Sp_trait_TRY_trans[,5] <- log10(Sp_trait_TRY_trans[,5])    # SLA
Sp_trait_TRY_trans[,6] <- log10(Sp_trait_TRY_trans[,6])    # Leaf nitrogen
Sp_trait_TRY_trans[,7] <- log10(Sp_trait_TRY_trans[,7])    # Leaf area
Sp_trait_TRY_trans[,8] <- ifelse(Sp_trait_TRY_trans[,8] == 0, 0.0001, Sp_trait_TRY_trans[,8])  # Avoid log(0) for seed mass

# Center and scale traits
Sp_trait_TRY_trans[,c(4:8)] <- scale(Sp_trait_TRY_trans[,c(4:8)])


```


##join trait x demo data
```{r}
#remove _ 
 demo_data_scaled_clean$Species <- gsub("_", " ", demo_data_scaled_clean$Species)

data <- demo_data_scaled_clean %>%
  left_join(Sp_trait_TRY_trans)

#parque si on ne le fait pas il y a un warning
data_scaled <- data %>%
  mutate(
    VH_sc  = scale(VH),
    SLA_sc = scale(SLA),
    LA_sc  = scale(LA),
    SM_sc  = scale(SM),
    LN_sc  = scale(LN),
    Elevation_sc = scale(mean_elevation)
  ) %>%
  dplyr::select(Region, Treatment, Family, Genus, Species, mean_r, a_sp, c_sp, mean_epsilon, Elevation_sc, VH_sc, SLA_sc, LA_sc, SM_sc, LN_sc)
```

##trait figure raw variation
```{r}

Sp_trait_TRY <- read.csv("new_TRY_raw_traits_with_na.csv", sep = ",", dec = ".") 
# FIGURE 4 — Trait variation per species (raw)

#remove _ 
demo_data_scaled_clean$Species <- gsub("_", " ", demo_data_scaled_clean$Species)
 
data_for_variation <- demo_data_scaled_clean %>%
  left_join(Sp_trait_TRY)

plot_trait_variation <- data_for_variation  %>%
  group_by(Treatment)%>%
   mutate(LA = LA / 100) %>% # mm² → cm²
  pivot_longer(
    cols      = c(VH, SLA, LN, LA, SM),
    names_to  = "trait",
    values_to = "value"
  ) %>%
  unique()%>%
   arrange(Treatment, Region) %>%
              mutate(order_graph = 1:n()) %>%
  filter(!is.na(value)) %>%
  mutate(trait = factor(trait,
                        levels = c("VH", "SLA", "LN", "LA", "SM"),
                        labels= c("VH (m)", "SLA (mm2 mg-1)", "LN (mg.g-1)", "LA (cm2)", "SM (mg)"))) %>%
  ggplot(aes(x = reorder(Region, order_graph), y = value, fill = Treatment)) +
  geom_boxplot(
    outlier.shape = 16,
    outlier.size  = 0.8,
    outlier.alpha = 0.4,
    width         = 0.5
  ) +
  scale_fill_manual(values=c("orange","dodgerblue", "firebrick1")) +
  facet_wrap(~ trait, scales = "free", ncol = 5) +
  labs(
    x     = NULL,
    y     = "Trait value (raw)",
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
   strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )+
  coord_flip()

plot_trait_variation

ggsave(filename = "Trait Distribution.png", plot = plot_trait_variation, bg = "white", width = 10, height = 8, dpi = 600)

data_for_variation %>%
  mutate(LA = LA / 100)  %>%# mm² → cm²
  summarise(
    LA_min    = min(LA,  na.rm = TRUE),
    LA_median = median(LA, na.rm = TRUE),
    LA_max    = max(LA,  na.rm = TRUE),
    LA_q95    = quantile(LA, 0.95, na.rm = TRUE)
  )
# Attendu : median ≈ 5-20 cm², max ≈ 300-350 cm²

#high LN are for fabaceae so OK: 
data_for_variation %>%
     filter(LN > 45) %>%
     select(Species, LN) %>%
     arrange(desc(LN))


data_for_variation%>%
  filter(VH > 5) %>%
  select(Species, VH) %>%
  arrange(desc(VH)) #Humulus scandens: ne plante grimpante annuelle qui peut atteindre 6 m en s'accrochant à d'autres végétaux

annual_transition3 %>%
  filter(Species == "Humulus_scandens") %>%
  count(Region, Treatment)
```

#Models M1

```{r}

#r
summary(best_M0_r)
M1_r <- lmer(mean_r ~Treatment + (1|Region) + (1|Species) + Elevation_sc + VH_sc + SLA_sc + LN_sc + LA_sc + SM_sc, data=data_scaled, na.action ="na.fail", REML = F) #rajouter na.action et REML = F pour dredge)
summary(M1_r)
dd_select_r <- dredge(M1_r,  evaluate = TRUE,  rank = "AICc")
nrow(dd_select_r[dd_select_r$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M1_r <- get.models(dd_select_r, subset =1) [[1]] 
best_M1_r

#a
best_M0_a 
summary(best_M0_a)
M1_a <- lmer(a_sp ~Treatment + (1|Region) + (1|Species) + Elevation_sc + VH_sc + SLA_sc + LN_sc + LA_sc + SM_sc, data=data_scaled, na.action ="na.fail", REML = F) #rajouter na.action et REML = F pour dredge)
summary(M1_a)
dd_select_a <- dredge(M1_a,  evaluate = TRUE,  rank = "AICc")
nrow(dd_select_a[dd_select_a$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M1_a <- get.models(dd_select_a, subset =1) [[1]] 
best_M1_a 

#c
best_M0_c 
summary(best_M0_c)
M1_c <- lmer(c_sp ~Treatment + (1|Region) + (1|Species) + VH_sc + SLA_sc + LN_sc + LA_sc + SM_sc, data=data_scaled, na.action ="na.fail", REML = F) #rajouter na.action et REML = F pour dredge)
summary(M1_c)
dd_select_c <- dredge(M1_c,  evaluate = TRUE,  rank = "AICc")
nrow(dd_select_c[dd_select_c$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M1_c <- get.models(dd_select_c, subset =1) [[1]] 
best_M1_c 

#epsilon
best_M0_epsilon
summary(best_M0_epsilon)
M1_epsilon <- lmer(mean_epsilon ~Treatment + (1|Region) + (1|Species) + VH_sc + SLA_sc + LN_sc + LA_sc + SM_sc, data=data_scaled, na.action ="na.fail", REML = F) 
summary(M1_epsilon)
dd_select_epsilon <- dredge(M1_epsilon,  evaluate = TRUE,  rank = "AICc", fixed    = "Treatment")
nrow(dd_select_epsilon[dd_select_epsilon$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M1_epsilon <- get.models(dd_select_epsilon, subset =1) [[1]] 
best_M1_epsilon
```

#Models M2

with interaction for teh remaining traits and treatment

```{r}
#rajouter les intéraction traits x treatment pour les modeles M2
#r
summary(best_M0_r)
summary(best_M1_r)
M2_r <- lmer(mean_r ~Treatment + Elevation_sc  + (1|Region) + (1|Species)+
               VH_sc * Treatment + 
               SLA_sc * Treatment + 
               LN_sc * Treatment, data=data_scaled, na.action ="na.fail", REML = F) #rajouter na.action et REML = F pour dredge)
summary(M2_r)
dd_select_r <- dredge(M2_r,  evaluate = TRUE,  rank = "AICc")
nrow(dd_select_r[dd_select_r$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M2_r <- get.models(dd_select_r, subset =1) [[1]] 
best_M2_r 

#a
summary(best_M0_a)
summary(best_M1_a)
M2_a <- lmer(a_sp ~Treatment + Elevation_sc + (1|Region) + (1|Species)+
               VH_sc * Treatment + 
               SLA_sc * Treatment + 
               LN_sc * Treatment + 
               LA_sc *Treatment, data=data_scaled, na.action ="na.fail", REML = F) #rajouter na.action et REML = F pour dredge)
summary(M2_a)
dd_select_a <- dredge(M2_a,  evaluate = TRUE,  rank = "AICc")
nrow(dd_select_a[dd_select_a$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M2_a <- get.models(dd_select_a, subset =1) [[1]] 
best_M2_a 

#c
summary(best_M0_c)
summary(best_M1_c)
M2_c <- lmer(c_sp ~Treatment + (1|Region) + (1|Species) +
               VH_sc * Treatment + 
               SLA_sc * Treatment, data=data_scaled, na.action ="na.fail", REML = F) #rajouter na.action et REML = F pour dredge)
summary(M2_c)
dd_select_c <- dredge(M2_c,  evaluate = TRUE,  rank = "AICc")
nrow(dd_select_c[dd_select_c$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M2_c <- get.models(dd_select_c, subset =1) [[1]] 
best_M2_c 

#epsilon
summary(best_M0_epsilon)
summary(best_M1_epsilon)
M2_epsilon <- lmer(mean_epsilon ~Treatment + (1|Species) + (1 |Region) +
                     SLA_sc*Treatment
                     , data=data_scaled, na.action ="na.fail", REML = F) #on avait enlever region et elevation
summary(M2_epsilon)
dd_select_epsilon <- dredge(M2_epsilon,  evaluate = TRUE,  rank = "AICc",  fixed    = "Treatment")
nrow(dd_select_epsilon[dd_select_epsilon$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
best_M2_epsilon <- get.models(dd_select_epsilon, subset =1) [[1]] 
best_M2_epsilon 
```


##emmeans 
```{r}
labels_trt <- c(
  "Control_Destination" = "Control\ndest",
  "Control_Origin"      = "Control\norigin",
  "Warm"                = "Warm"
)

#r
summary(best_M2_r) 
contrast_r <- emmeans(best_M2_r, pairwise ~  ~Treatment )

em_r <- emmeans(best_M2_r, ~ Treatment) %>%
  as.data.frame() %>%
  mutate(Treatment_label = factor(labels_trt[Treatment],
                                  levels = labels_trt))

emtrends(best_M2_r, pairwise ~ Treatment, var="VH_sc" )
emtrends(best_M2_r, pairwise ~ Treatment, var = "SLA_sc")
emtrends(best_M2_r, pairwise ~ Treatment, var = "LN_sc")

#a
summary(best_M2_a) 

contrast_a <- emmeans(best_M2_a, pairwise ~  ~Treatment )

em_a <- emmeans(best_M2_a, ~ Treatment) %>%
  as.data.frame() %>%
  mutate(Treatment_label = factor(labels_trt[Treatment],
                                  levels = labels_trt))

#c
summary(best_M2_c) 

contrast_c <- emmeans(best_M2_c, pairwise ~  ~Treatment )

em_c <- emmeans(best_M2_c, ~ Treatment) %>%
  as.data.frame() %>%
  mutate(Treatment_label = factor(labels_trt[Treatment],
                                  levels = labels_trt))

#epsilon
summary(best_M2_epsilon)

#emmeans(best_M2_epsilon, pairwise ~  ~Treatment )


em_eps <- emmeans(best_M2_epsilon, ~ Treatment) %>%
  as.data.frame() %>%
  mutate(Treatment_label = factor(labels_trt[Treatment],
                                  levels = labels_trt))

```

##Print summaries of best models 
```{r}
#trop bonne fonction tab_model, prints perfect all together model summaries

tab_model(best_M2_r,
          file   = "best_M2_r.html",
          digits = 3,
          show.re.var = TRUE,
          show.icc    = TRUE,
          title        = "M2 — Per-capita growth rate r")

tab_model(best_M2_a,
          file   = "best_M2_a.html",
          digits = 3,
          show.re.var = TRUE,
          title        = "M2 — Intrinsic growth rate a")

tab_model(best_M2_c,
          file   = "best_M2_c.html",
          digits = 3,
          show.re.var = TRUE,
          title        = "M2 — Density dependence c")

tab_model(best_M2_epsilon,
          file   = "best_M2_epsilon.html",
          digits = 3,
          show.re.var = TRUE,
          title        = "M2 — epsilon")

#tous les modèles
tab_model(
  best_M2_r,
  best_M2_a,
  best_M2_c,
  best_M2_epsilon,
  file         = "best_M2_all.html",
  digits       = 3,
  show.re.var  = TRUE,
  show.icc     = TRUE,
  dv.labels    = c("r", "a", "c", "epsilon"),
  title        = "M2 models — demographic parameters"
)
```


#plot models - boxplots

## r - per capita growth rate
```{r}
cols_trt <- c(
  "Control_Origin" = "dodgerblue",
  "Control_Destination"   = "orange",
  "Warm"           = "firebrick1"
)

labels_trt <- c(
  "Control_Origin" = "Control\norigin",
  "Control_Destination"   = "Control\ndest",
  "Warm"           = "Warm"
)

data_scaled <- data_scaled %>%
  mutate(Treatment_label = factor(labels_trt[Treatment],
                                  levels = labels_trt))

# r — taux de croissance observé

pred_r_ln <- ggpredict(best_M2_r,
                        terms = c("LN_sc", "Treatment")) %>%
  as.data.frame() %>%
  rename(Treatment = group) %>%
  mutate(Treatment_label = factor(labels_trt[as.character(Treatment)],
                                  levels = labels_trt))

pred_r_sla <- ggpredict(best_M2_r,
                         terms = c("SLA_sc", "Treatment")) %>%
  as.data.frame() %>%
  rename(Treatment = group) %>%
  mutate(Treatment_label = factor(labels_trt[as.character(Treatment)],
                                  levels = labels_trt))

# boxplot r
p_r_box <- ggplot() +
  geom_boxplot(
    data  = data_scaled,
    aes(x = Treatment_label, y = mean_r, fill = Treatment),
    alpha = 0.35, outlier.shape = NA) +
  geom_pointrange(
    data = em_r,
    aes(x = Treatment_label, y = emmean,
        ymin = lower.CL, ymax = upper.CL),
    colour = "red", size = 0.7, linewidth = 0.9
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  scale_fill_manual(values = cols_trt) +
  labs(x = NULL, y = "r") +
  theme_minimal(base_size = 20) +
  
  #annotation
  annotate("segment", x=2, xend =3, y=0.8, yend=0.8, linewidth =0.5)+
  annotate("text", x=2.5, y=1, label= "****", size = 7)+
  theme(legend.position    = "bottom",
        panel.grid.major.x = element_blank(), axis.text.x = element_blank(), legend.title = element_blank())+
  ylim(-0.8, 1.1)

p_r_box 

#r ~ LN
p_r_ln <- ggplot(pred_r_ln,
                 aes(x = x, y = predicted,
                     colour = Treatment, fill = Treatment)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols_trt, labels = labels_trt) +
  scale_fill_manual(values   = cols_trt, labels = labels_trt) +
   annotate("text", x=1, y= 0.4, label= "CO vs W marg" , hjust=0 )+
  annotate("text", x=1, y= 0.3, label= "CD vs W marg" , hjust=0 )+
  annotate("text", x=1, y= 0.2, label= "CO vs CD ns", hjust=0 )+
  labs(x = "LN (scaled)", y = "r",
       colour = NULL, fill = NULL,
       title = "LN × Treatment") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
p_r_ln

#r ~ SLA
p_r_sla <- ggplot(pred_r_sla,
                  aes(x = x, y = predicted,
                      colour = Treatment,
                      fill   = Treatment)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +

  annotate("text", x=-5, y= 0.4, label= "CO vs W *" , hjust=0 )+
  annotate("text", x=-5, y= 0.3, label= "CD vs W ns" , hjust=0 )+
  annotate("text", x=-5, y= 0.2, label= "CO vs CD ns", hjust=0 )+

  labs(x      = "SLA (scaled)",
       y      = "r",
       colour = NULL,
       fill   = NULL,
       title  = "SLA × Treatment") +
  scale_colour_manual(values = cols_trt, labels = labels_trt) +
  scale_fill_manual(values   = cols_trt, labels = labels_trt) +
  theme_minimal(base_size = 11) +
  theme(legend.position  = "none", 
        panel.grid.minor = element_blank())

p_r_sla

```

## a - average growth rate 

```{r}
# a — taux intrinsèque

p_a_box <- ggplot() +
  geom_boxplot(
    data  = data_scaled,
    aes(x = Treatment_label, y = a_sp, fill = Treatment),
    alpha = 0.35, outlier.shape = NA)+
  geom_pointrange(
    data = em_a,
    aes(x = Treatment_label, y = emmean,
          ymin = lower.CL, ymax = upper.CL),
    colour = "red", size = 0.7, linewidth = 0.9
  ) +
  scale_fill_manual(values = cols_trt) +
  labs(x = NULL, y = "a") +
  
    #annotation
  annotate("segment", x=1, xend =3, y=2.7, yend=2.7, linewidth =0.5)+
  annotate("text", x=2.5, y=2.9, label= "marg", size = 7)+
  annotate("segment", x=2, xend =3, y=2.2, yend=2.2, linewidth =0.5)+
  annotate("text", x=2.5, y=2.3, label= "****", size = 7)+
  
  theme_minimal(base_size = 20) +
theme(legend.position    = "bottom",
        panel.grid.major.x = element_blank(), axis.text.x = element_blank(), legend.title = element_blank())

p_a_box 

```

## c - density dependance

```{r}
# c — densité dépendance

p_c_box <- ggplot() +
  geom_boxplot(
    data  = data_scaled,
    aes(x = Treatment_label, y = c_sp, fill = Treatment),
    alpha = 0.35, outlier.shape = NA) +
  geom_pointrange(
    data = em_c,
    aes(x = Treatment_label, y = emmean,
          ymin = lower.CL, ymax = upper.CL),
    colour = "red", size = 0.7, linewidth = 0.9
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  
      #annotation
  annotate("segment", x=1, xend =3, y=-0.35, yend=-0.35, linewidth =0.5)+
  annotate("text", x=2.5, y=-0.3, label= "marg", size = 7)+

  scale_fill_manual(values = cols_trt) +
  labs(x = NULL, y = "c") +
  theme_minimal(base_size = 20) +
theme(legend.position    = "bottom",
        panel.grid.major.x = element_blank(), axis.text.x = element_blank(), legend.title = element_blank())

p_c_box 

```

## epsilon - stochasticity

```{r}
# epsilon — stochasticité

# Calculer min/max
range_sla <- range(demo_traits_scaled$SLA_sc, na.rm = TRUE)
range_la  <- range(demo_traits_scaled$LA_sc,  na.rm = TRUE)

pred_eps_sla <- ggpredict(
  best_M2_epsilon,
  terms = paste0("SLA_sc [",
                 paste(seq(range_sla[1], range_sla[2],
                           length.out = 100),
                       collapse = ", "), "]")
) %>%
  as.data.frame()

pred_eps_la <- ggpredict(
  best_M2_epsilon,
  terms = paste0("LA_sc [",
                 paste(seq(range_la[1], range_la[2],
                           length.out = 100),
                       collapse = ", "), "]")
) %>%
  as.data.frame()


p_eps_box <- ggplot() +
  geom_boxplot(
    data  = demo_traits_scaled,
    aes(x = Treatment_label, y = epsilon, fill = Treatment),
    alpha = 0.35, outlier.shape = NA ) +
  geom_pointrange(
    data = em_eps,
    aes(x = Treatment_label, y = emmean,
         ymin = asymp.LCL, ymax = asymp.UCL),
    colour = "red", size = 0.7, linewidth = 0.9
  ) +
  scale_fill_manual(values = cols_trt) +
  labs(x = NULL, y = "Epsilon") +
  theme_minimal(base_size = 20) +
  
    #annotation
  annotate("segment", x=1, xend =3, y=3.7, yend=3.7, linewidth =0.5)+
  annotate("text", x=2.5, y=3.9, label= "*", size = 7)+
  annotate("segment", x=1, xend =2, y=2.2, yend=2.2, linewidth =0.5)+
  annotate("text", x=1.5, y=2.5, label= "****", size = 7)+
  theme(legend.position    = "bottom",
        panel.grid.major.x = element_blank(), axis.text.x = element_blank(), legend.title = element_blank())

p_eps_vh <- ggplot(pred_eps_vh,
                   aes(x = x, y = predicted,
                       colour = Treatment, fill = Treatment)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols_trt, labels = labels_trt) +
  scale_fill_manual(values   = cols_trt, labels = labels_trt) +
  
     annotate("text", x=-2, y= 1.3, label= "CO vs W ***" , hjust=0 )+
  annotate("text", x=-2, y= 1, label= "CD vs W ns" , hjust=0 )+
  annotate("text", x=-2, y= 0.7, label= "CO vs CD ***", hjust=0 )+
  
  labs(x = "VH (scaled)", y = "Epsilon",
       colour = NULL, fill = NULL,
       title = "VH × \n Treatment") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
p_eps_vh 

p_eps_sm <- ggplot(pred_eps_sm,
                    aes(x = x, y = predicted,
                        colour = Treatment, fill = Treatment)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = cols_trt, labels = labels_trt) +
  scale_fill_manual(values   = cols_trt, labels = labels_trt) +
  
   annotate("text", x=-2, y= -0.4, label= "CO vs W ns" , hjust=0 )+
  annotate("text", x=-2, y= -0.6, label= "CD vs W ns" , hjust=0 )+
  annotate("text", x=-2, y= -0.8, label= "CO vs CD ns", hjust=0 )+
  
  labs(x = "SM (scaled)", y = "Epsilon",
       colour = NULL, fill = NULL,
       title = "SM × \n Treatment") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
p_eps_sm 
# SLA — effet principal seulement
p_eps_sla <- ggplot(pred_eps_sla,
                    aes(x = x, y = predicted)) +
   geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "grey60") +
  geom_line(linewidth = 1, colour = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  labs(x = "SLA (scaled)", y = "ε",
       title = "SLA") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

# LA — effet principal seulement (marginal)
p_eps_la <- ggplot(pred_eps_la,
                   aes(x = x, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "grey60") +
  geom_line(linewidth = 1, colour = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  labs(x = "LA (scaled)", y = "ε",
       title = "LA") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())


```

##FIGURE 1
```{r}
#All
demo_treatment <- ggarrange(p_r_box, p_a_box, p_c_box, common.legend = T, legend = "bottom", ncol=1)

interaction_r <- ggarrange(p_r_ln, p_r_sla,  common.legend = T, legend = "none", ncol=2)

fig_main <- ggarrange(demo_treatment, interaction_r, ncol=2, widths = c(0.9, 1.5)) 

ggsave("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/Transplantation Alpine/TransPlantAlpin/Figures/Demo_trait.png", fig_main, bg = "white", width =14, height = 10)
```

#PCA
```{r}

dim(data_scaled)
names(data_scaled)
head(data_scaled)

# Run PCA on standardized traits
pca_result <- PCA(data_scaled[11:15], scale.unit = TRUE, graph = TRUE)

# Variance explained by each axis
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 100),
         main = "Variance explained per axis")

# Extract species coordinates in PCA space
eig_values <- pca_result$eig
significant_axes <- 1:ncol(pca_result$ind$coord)
coords <- pca_result$ind$coord[, significant_axes, drop = FALSE]
coord <- data.frame(Species = data_scaled$Species, coords)

# Visualize species and trait loadings
fviz_pca_biplot(pca_result,
                axes = c(1, 2),
                geom.ind = "point",
                col.ind = "cadetblue3",
                alpha.ind = 0.6,
                col.var = "darkblue",
                arrowsize = 1.2,
                label = "var"
) +
  theme_minimal()



pca_demo <- coord %>%
  left_join(data_scaled, by = "Species") %>%
  filter(!is.na(Treatment))


pca_demo <- data.frame(
  Species   = data_scaled$Species,
  Region    = data_scaled$Region,
  Treatment = data_scaled$Treatment,
  mean_r    = data_scaled$mean_r,
  a_sp      = data_scaled$a_sp,
  c_sp      = data_scaled$c_sp,
  Dim.1     = pca_result$ind$coord[, 1],
  Dim.2     = pca_result$ind$coord[, 2]
)

var_coords <- as.data.frame(pca_result$var$coord[, 1:2])
var_coords$varname <- rownames(var_coords)

scale_factor <- max(abs(pca_demo[, c("Dim.1", "Dim.2")])) /
                max(abs(var_coords[, c("Dim.1", "Dim.2")])) * 0.6

pc1_lab <- paste0("PC1 (", round(pca_result$eig[1, 2], 1), "%)")
pc2_lab <- paste0("PC2 (", round(pca_result$eig[2, 2], 1), "%)")

#limites pour le gradient
lim_r <- quantile(pca_demo$mean_r, c(0.02, 0.98), na.rm = TRUE)
lim_a <- quantile(pca_demo$a_sp,   c(0.02, 0.98), na.rm = TRUE)
lim_c <- quantile(pca_demo$c_sp,   c(0.02, 0.98), na.rm = TRUE)

geom_trait_arrows <- function() {
  list(
    geom_segment(
      data = var_coords,
      aes(x = 0, y = 0,
          xend = Dim.1 * scale_factor,
          yend = Dim.2 * scale_factor),
      arrow       = arrow(length = unit(0.2, "cm"), type = "closed"),
      colour      = "black", linewidth = 0.7,
      inherit.aes = FALSE
    ),
    geom_text(
      data = var_coords,
      aes(x     = Dim.1 * scale_factor * 1.25,
          y     = Dim.2 * scale_factor * 1.25,
          label = varname),
      colour      = "black", size = 3.2, fontface = "bold",
      inherit.aes = FALSE
    )
  )
}

a_r <- ggplot(pca_demo %>% filter(!is.na(mean_r)),
                  aes(x = Dim.1, y = Dim.2, colour = mean_r)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_trait_arrows() +
  scale_colour_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick1",
    midpoint = 0,
    limits   = lim_r,
    oob      = scales::squish,
    name     = "r",
    guide    = guide_colorbar(barwidth = 0.5, barheight = 4,
                               title.position = "top")
  ) +
  labs(x = pc1_lab, y = pc2_lab, title = "a) Growth rate r") +
  theme_minimal(base_size = 11) +
  theme(legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 7),
        panel.grid.minor = element_blank())


p_pca_a <- ggplot(pca_demo %>% filter(!is.na(a_sp)),
                  aes(x = Dim.1, y = Dim.2, colour = a_sp)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_trait_arrows() +
  scale_colour_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick1",
    midpoint = 0,
    limits   = lim_a,
    oob      = scales::squish,
    name     = "a",
    guide    = guide_colorbar(barwidth = 0.5, barheight = 4,
                               title.position = "top")
  ) +
  labs(x = pc1_lab, y = pc2_lab, title = "b) Intrinsic rate a") +
  theme_minimal(base_size = 11) +
  theme(legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 7),
        panel.grid.minor = element_blank())


p_pca_c <- ggplot(pca_demo %>% filter(!is.na(c_sp)),
                  aes(x = Dim.1, y = Dim.2, colour = c_sp)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_trait_arrows() +
  scale_colour_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick1",
    midpoint = median(pca_demo$c_sp, na.rm = TRUE),
    limits   = lim_c,
    oob      = scales::squish,
    name     = "c",
    guide    = guide_colorbar(barwidth = 0.5, barheight = 4,
                               title.position = "top")
  ) +
  labs(x = pc1_lab, y = pc2_lab, title = "c) dd c") +
  theme_minimal(base_size = 11) +
  theme(legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 7),
        panel.grid.minor = element_blank())

fig_pca <- p_pca_r | p_pca_a | p_pca_c

fig_pca


######################old
# # Coordonnées des traits pour les flèches
# var_coords <- as.data.frame(pca_result$var$coord[, 1:2])
# var_coords$varname <- rownames(var_coords)
# 

# Limites communes pour chaque paramètre
# lim_r   <- quantile(pca_demo$r_mean,       c(0.02, 0.98), na.rm = TRUE)
# lim_a   <- quantile(pca_demo$a_mean,       c(0.02, 0.98), na.rm = TRUE)
# lim_c   <- quantile(pca_demo$c_mean,       c(0.02, 0.98), na.rm = TRUE)
# lim_eps <- quantile(pca_demo$epsilon_mean, c(0.02, 0.98), na.rm = TRUE)
# 
# # r 
# 
# p_pca_r <- ggplot() +
#   geom_point(
#     data = pca_demo %>% filter(!is.na(r_mean)),
#     aes(x = Dim.1, y = Dim.2, colour = r_mean),
#     size = 2, alpha = 0.7
#   ) +
#   geom_segment(
#     data = var_coords,
#     aes(x = 0, y = 0,
#         xend = Dim.1*5,
#         yend = Dim.2*5),
#     arrow     = arrow(length = unit(0.2, "cm"), type = "closed"),
#     colour    = "black", linewidth = 0.7,
#     inherit.aes = FALSE
#   ) +
#   geom_text(
#     data = var_coords,
#     aes(x     = Dim.1 * 6.2 ,
#         y     = Dim.2 * 6.2 ,
#         label = varname),
#     colour = "black", size = 12,
#     inherit.aes = FALSE
#   ) +
#   scale_colour_gradient2(
#     low      = "steelblue",
#     mid      = "white",
#     high     = "red",
#     midpoint = mean(lim_r),
#     limits   = lim_r,
#     oob      = scales::squish,
#     name     = "r"
#   ) +
#   labs(
#     x     = paste0("PC1 (", round(pca_result$eig[1, 2], 1), "%)"),
#     y     = paste0("PC2 (", round(pca_result$eig[2, 2], 1), "%)"),
#     title = "r"
#   ) +
#   theme_minimal(base_size = 20) +
#  theme(
#   legend.position  = "right",
#   legend.key.size  = unit(0.4, "cm"),   
#   legend.text      = element_text(size = 16),
#   legend.title     = element_text(size = 14),
#   panel.grid.minor = element_blank()
# )
# 
# # a
# 
# p_pca_a <- ggplot() +
#   geom_point(
#     data = pca_demo %>% filter(!is.na(a_mean)),
#     aes(x = Dim.1, y = Dim.2, colour = a_mean),
#     size = 2, alpha = 0.7
#   ) +
#   geom_segment(
#     data = var_coords,
#     aes(x = 0, y = 0,
#         xend = Dim.1 * 5,
#         yend = Dim.2 * 5),
#     arrow       = arrow(length = unit(0.2, "cm"), type = "closed"),
#     colour      = "black", linewidth = 0.7,
#     inherit.aes = FALSE
#   ) +
#   geom_text(
#     data = var_coords,
#     aes(x     = Dim.1 * 5 * 1.2,
#         y     = Dim.2 * 5 * 1.2,
#         label = varname),
#     colour = "black", size = 12,
#     inherit.aes = FALSE
#   ) +
#   scale_colour_gradient2(
#     low      = "steelblue",
#     mid      = "white",
#     high     = "firebrick1",
#     midpoint = mean(lim_a),
#     limits   = lim_a,
#     oob      = scales::squish,
#     name     = "a"
#   ) +
#   labs(
#     x     = paste0("PC1 (", round(pca_result$eig[1, 2], 1), "%)"),
#     y     = paste0("PC2 (", round(pca_result$eig[2, 2], 1), "%)"),
#     title = "a"
#   ) +
#    theme_minimal(base_size = 20) +
#   theme(
#   legend.position  = "right",
#   legend.key.size  = unit(0.4, "cm"),   
#     legend.text      = element_text(size = 16),
#   legend.title     = element_text(size = 14),
#   panel.grid.minor = element_blank()
# )
# 
# # c
# 
# p_pca_c <- ggplot() +
#   geom_point(
#     data = pca_demo %>% filter(!is.na(c_mean)),
#     aes(x = Dim.1, y = Dim.2, colour = c_mean),
#     size = 2, alpha = 0.7
#   ) +
#   geom_segment(
#     data = var_coords,
#     aes(x = 0, y = 0,
#         xend = Dim.1 * 5,
#         yend = Dim.2 * 5),
#     arrow       = arrow(length = unit(0.2, "cm"), type = "closed"),
#     colour      = "black", linewidth = 0.7,
#     inherit.aes = FALSE
#   ) +
#   geom_text(
#     data = var_coords,
#     aes(x     = Dim.1 * 6.2,
#         y     = Dim.2 * 6.2,
#         label = varname),
#     colour = "black", size = 12,
#     inherit.aes = FALSE
#   ) +
#   scale_colour_gradient2(
#     low      = "steelblue",
#     mid      = "white",
#     high     = "firebrick1",
#     midpoint = mean(lim_c),
#     limits   = lim_c,
#     oob      = scales::squish,
#     name     = "c"
#   ) +
#   labs(
#     x     = paste0("PC1 (", round(pca_result$eig[1, 2], 1), "%)"),
#     y     = paste0("PC2 (", round(pca_result$eig[2, 2], 1), "%)"),
#     title = "c"
#   ) +
#   theme_minimal(base_size = 20) +
#  theme(
#   legend.position  = "right",
#   legend.key.size  = unit(0.4, "cm"),  
#    legend.text      = element_text(size = 16),
#   legend.title     = element_text(size = 14),
#   panel.grid.minor = element_blank()
# )
# 
# # epsilon
# 
# p_pca_eps <- ggplot() +
#   geom_point(
#     data = pca_demo %>% filter(!is.na(epsilon_mean)),
#     aes(x = Dim.1, y = Dim.2, colour = epsilon_mean),
#     size = 1.8, alpha = 0.7
#   ) +
#   geom_segment(
#     data = var_coords,
#     aes(x = 0, y = 0,
#         xend = Dim.1 * 5,
#         yend = Dim.2 * 5),
#     arrow       = arrow(length = unit(0.2, "cm"), type = "closed"),
#     colour      = "black", linewidth = 0.7,
#     inherit.aes = FALSE
#   ) +
#   geom_text(
#     data = var_coords,
#     aes(x     = Dim.1 * 6.2,
#         y     = Dim.2 * 6.2,
#         label = varname),
#     colour = "black", size = 12,
#     inherit.aes = FALSE
#   ) +
#   scale_colour_gradient2(
#     low      = "steelblue",
#     mid      = "white",
#     high     = "firebrick1",
#     midpoint = 0,
#     limits   = lim_eps,
#     oob      = scales::squish,
#     name     = "epsilon"
#   ) +
#   labs(
#     x     = paste0("PC1 (", round(pca_result$eig[1, 2], 1), "%)"),
#     y     = paste0("PC2 (", round(pca_result$eig[2, 2], 1), "%)"),
#     title = "Epsilon"
#   ) +
#   theme_minimal(base_size = 20) +
#   theme(
#   legend.position  = "right",
#   legend.key.size  = unit(0.4, "cm"),   # taille de la barre de couleur
#     legend.text      = element_text(size = 16),
#   legend.title     = element_text(size = 14),
#   panel.grid.minor = element_blank()
# )
# 
# # ── Figure finale — 2 × 2 ─────────────────────────────────────
# 
# fig_pca <- ggarrange(p_pca_r, p_pca_a, p_pca_c, p_pca_eps, nrow=2, ncol=2, labels = c("A", "B", "C", "D"), font.label = list(face="bold", size= 30))
# 
# fig_pca


# ggsave("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/Transplantation Alpine/TransPlantAlpin/Figures/PCA_demo.png", fig_pca, bg = "white", width =8, height = 6)
# 
# 
# p_pca_r <- p_pca_r + facet_wrap(~Treatment)
# p_pca_a <- p_pca_a + facet_wrap(~Treatment)
# p_pca_c <- p_pca_c + facet_wrap(~Treatment)
# p_pca_eps <- p_pca_eps + facet_wrap(~Treatment)
# 
# fig2_pca <- ggarrange(p_pca_r, p_pca_a, p_pca_c, p_pca_eps, ncol=1, labels = c("A", "B", "C", "D"), font.label = list(face="bold", size= 30))
# 
# ggsave("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/Transplantation Alpine/TransPlantAlpin/Figures/PCA_demo2.png", fig2_pca, bg = "white", width =6, height =10)
```

##PCA par traitement
```{r}
r_warm <- ggplot(pca_demo %>% filter(!is.na(mean_r)) %>%
                     filter(Treatment =="Warm"),
                  aes(x = Dim.1, y = Dim.2, colour = mean_r)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_trait_arrows() +
  scale_colour_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick1",
    midpoint = 0,
    limits   = lim_r,
    oob      = scales::squish,
    name     = "r",
    guide    = guide_colorbar(barwidth = 0.5, barheight = 4,
                               title.position = "top")
  ) +
  labs(x = pc1_lab, y = pc2_lab, title = "a) Growth rate r warm") +
  theme_minimal(base_size = 11) +
  theme(legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 7),
        panel.grid.minor = element_blank())


r_CD <- ggplot(pca_demo %>% filter(!is.na(mean_r)) %>%
                     filter(Treatment =="Control_Destination"),
                  aes(x = Dim.1, y = Dim.2, colour = mean_r)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_trait_arrows() +
  scale_colour_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick1",
    midpoint = 0,
    limits   = lim_r,
    oob      = scales::squish,
    name     = "r",
    guide    = guide_colorbar(barwidth = 0.5, barheight = 4,
                               title.position = "top")
  ) +
  labs(x = pc1_lab, y = pc2_lab, title = "a) Growth rate r Control_Dest") +
  theme_minimal(base_size = 11) +
  theme(legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 7),
        panel.grid.minor = element_blank())


r_CO <- ggplot(pca_demo %>% filter(!is.na(mean_r)) %>%
                     filter(Treatment =="Control_Origin"),
                  aes(x = Dim.1, y = Dim.2, colour = mean_r)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_trait_arrows() +
  scale_colour_gradient2(
    low      = "steelblue",
    mid      = "white",
    high     = "firebrick1",
    midpoint = 0,
    limits   = lim_r,
    oob      = scales::squish,
    name     = "r",
    guide    = guide_colorbar(barwidth = 0.5, barheight = 4,
                               title.position = "top")
  ) +
  labs(x = pc1_lab, y = pc2_lab, title = "a) Growth rate r Control_Origin") +
  theme_minimal(base_size = 11) +
  theme(legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 7),
        panel.grid.minor = element_blank())


fig_pca_treatment <- r_CO | r_CD | r_warm

fig_pca_treatment
```

