---
title: "Manscript analysis"
output: html_document
date: "2026-07-29"
editor_options: 
  chunk_output_type: console
---

#library
```{r}
library(readr)
 library(here)
#library(tidyverse)   
# library(readxl)
library(ggpubr)
library(funrar)
 library(dplyr)
 library(tidyr)
# library(stringr)
 library(ggplot2)
 library(TNRS) #c'est une application, pas une base de données
 library(stargazer)
library(ggeffects)
library(FactoMineR)
library(factoextra)
library(stringr)
library(tibble)
library(purrr)
library(lme4)
library(emmeans)
library(lmerTest) #for confint
library(betapart) #for turonver
library(patchwork) #to combine plots
library(MuMIn)#Lucie: conseil pour choisir les modèles. le package s'appelle multi-model inference, la fonction dredge va selectionner de manière automatique des subset models depuis le modele le plus general donné en entrée.
library(sjPlot)
library(jtools)#for summ function
library(mice)
library(VIM)
library(broom)
library(rstatix)
library(corrplot)
library(mdatools)
library(paran)
library(ggrepel)
library(ggh4x)
```



#Data


```{r}

#RELATIVE abundance per treatement
site_species_rel_warm <- read.csv("./intermediate data/site_species_warm_end_clean_rel_abund_Cover_initial_final.csv")

site_species_rel_W <-read.csv("./intermediate data/site_species_CO_end_clean_rel_abund_Cover_initial_final.csv")

site_species_rel_CD <- read.csv("./intermediate data/site_species_CD_end_clean_rel_abund_Cover_initial_final.csv")


# Remove 0/0 cases (species absent both times)
data_full_warm <- site_species_rel_warm %>%
  filter(!(Rel_Cover_initial == 0 & Rel_Cover_final == 0)) %>%  # retire seulement les 0/0
  mutate(
    dynamique_statut = case_when(
      Rel_Cover_initial == 0   & Rel_Cover_final > 0  ~ "Colonizer",
      Rel_Cover_final   == 0   & Rel_Cover_initial > 0 ~ "Extinct",
      Rel_Cover_initial > 0    & Rel_Cover_final > 0   ~ "Persistent",
      TRUE ~ NA_character_
    )
  )

data_full_W <- site_species_rel_W %>%
  filter(!(Rel_Cover_initial == 0 & Rel_Cover_final == 0)) %>%
  mutate(
    dynamique_statut = case_when(
      Rel_Cover_initial == 0   & Rel_Cover_final > 0  ~ "Colonizer",
      Rel_Cover_final   == 0   & Rel_Cover_initial > 0 ~ "Extinct",
      Rel_Cover_initial > 0    & Rel_Cover_final > 0   ~ "Persistent",
      TRUE ~ NA_character_
    )
  )

data_full_CD <-site_species_rel_CD %>%
  filter(!(Rel_Cover_initial == 0 & Rel_Cover_final == 0)) %>%
  mutate(
    dynamique_statut = case_when(
      Rel_Cover_initial == 0   & Rel_Cover_final > 0  ~ "Colonizer",
      Rel_Cover_final   == 0   & Rel_Cover_initial > 0 ~ "Extinct",
      Rel_Cover_initial > 0    & Rel_Cover_final > 0   ~ "Persistent",
      TRUE ~ NA_character_
    )
  )

# Combiner les trois traitements
data_full <- bind_rows(
  data_full_warm %>% mutate(Treatment = "warm"),
  data_full_W   %>% mutate(Treatment = "Control_origin"),
  data_full_CD   %>% mutate(Treatment = "Control_dest")
) %>%
  filter(!is.na(dynamique_statut))

# Vérifier les effectifs
data_full %>% count(dynamique_statut, Treatment)


# add variable elevation and plotSize

metadata <- read.csv(here("intermediate data","metadata.csv"))

site_species_rel_abund <-data_full %>% left_join(metadata %>% dplyr::select(Region, destSiteID, Elevation, PlotSize)) %>%
  relocate(PlotSize, .after= destPlotID)

data <-site_species_rel_abund %>% relocate(Elevation, .after= destPlotID)%>%
   mutate(delta_years = Year_final -Year_initial)
names(data)

unique(data$Region)

```


##add temperature
```{r}

# Load original dataset 
load("./raw data/TransPlantNetwork.RData")

climdata <- climdata %>% filter(month %in% c("6","7","8")) #average of hottest months 

temperature <- climdata %>%
  filter(year %in% sample(2007:2022)) %>%
  group_by(gradient, elevation, year) %>%
  filter(var == "temperature")%>%
   mutate(elevation = ifelse(gradient == "CH_Calanda" & elevation == 2800, 2600, elevation))%>% # Correction to Calanda experiment
  rename(Region = gradient) 

CH_Calanda_temp <- temperature%>%
  filter(Region == "CH_Calanda")%>%
  group_by(Region, elevation, year)%>%
  filter(year %in% sample(YearMin:YearMax))%>%
  summarise(mean_temperature = mean(value), .groups = "drop")%>%
  pivot_wider(names_from = year , values_from = mean_temperature)%>%
  mutate(row_mean = rowMeans(dplyr::select(., -c(1:2)), na.rm = TRUE))


survey_years <- data %>%
  group_by(Region) %>%
  summarise(year_min = min(Year_initial, na.rm = TRUE),
            year_max = max(Year_final, na.rm = TRUE),
            .groups = "drop")

region <- unique(temperature$Region) 
region <- region[region!="IN_Kashmir"]
region <- region[region!="US_Arizona"]
region <- region[region!="US_Colorado"]

temp_data <- data.frame()

for (i in region) {
  
  # get the survey year range for this region
  yr_min <- survey_years %>% filter(Region == i) %>% pull(year_min)
  yr_max <- survey_years %>% filter(Region == i) %>% pull(year_max)
  
  temp_results <- temperature%>%
    filter(Region == i)%>%
    filter(year >= yr_min & year <= yr_max) %>%
    group_by(Region,elevation, year)%>%
    summarise(mean_temperature = mean(value),.groups = "drop")%>%
    pivot_wider(names_from = year , values_from = mean_temperature)%>%
    mutate(temp_mean = rowMeans(dplyr::select(., -c(1:2)), na.rm = TRUE))%>%
    rename(Elevation = elevation)%>%
    dplyr::select(Region, Elevation, temp_mean) 
  
  temp_data<- rbind(temp_data, temp_results)
  
}


site_elevation <- data %>%
  distinct(Region, destSiteID, Elevation) %>%
  left_join(temp_data)%>%
  rename(SiteID = destSiteID)

data<- data%>%
  # join origin temp
  left_join(site_elevation %>% dplyr::select(Region, SiteID, temp_mean),
            by = c("Region", "originSiteID" = "SiteID")) %>%
  rename(temp_origin = temp_mean) %>%
  # join dest temp
  left_join(site_elevation %>% dplyr::select(Region, SiteID, temp_mean),
            by = c("Region", "destSiteID" = "SiteID")) %>%
  rename(temp_dest = temp_mean) %>%
  mutate(delta_temp = temp_dest - temp_origin)

```

#Relative abundance check sum to 1




```{r}
A<-  data%>% group_by(Region, Treatment, originSiteID, destSiteID, destPlotID)%>%summarize(sum=sum(Rel_Cover_initial))

#let's remove REgion SE_Abisko and destplotID H32 parce qu'il y a que deux espèces... which can be explained by 
# * species not identified to species level were removed (ex. for SE_Abisko plot H32: Carex sp. removed). 
# * species that were in fact trees (ex. ex. for SE_Abisko plot H32: Salix polarix removed).
# * species that appeared only once in the plot throughout the whole experiment was removed to eliminate potential botanical errors (we don't know if it is really a botanical error or a true presence)

data <- data %>% 
  filter(!destPlotID=="H32")
```


# r plot and site
```{r}

#threshold
k <- 5 

data_threshold <-  data %>%
  group_by(Region, Treatment, originSiteID, destSiteID, destPlotID) %>%
  #mettre un seuil minimum pour avoir une abondance non nulle pour les colonisatrices et les extinctes (respectivement au debut/à la fin)
  mutate(
    min_Cover_final      = min(Rel_Cover_final[Rel_Cover_final > 0],
                               na.rm = TRUE),
    detection_threshold  = min_Cover_final / k,
    Rel_Cover_initial_adj = case_when(
      dynamique_statut == "Colonizer" ~ detection_threshold,
      TRUE                            ~ Rel_Cover_initial
    ),
    min_Cover_initial    = min(Rel_Cover_initial[Rel_Cover_initial > 0],
                               na.rm = TRUE),
    detection_threshold2 = min_Cover_initial / k,
    Rel_Cover_final_adj  = case_when(
      dynamique_statut == "Extinct" ~ detection_threshold2,
      TRUE                          ~ Rel_Cover_final)) %>%
  ungroup() %>%
   filter(
    dynamique_statut %in% c("Persistent", "Colonizer", "Extinct"),
    Rel_Cover_initial_adj > 0,
    is.finite(log10(Rel_Cover_initial_adj)))
  
#calculate r at the plot level
data_plot <- data_threshold %>%
  group_by(Region, Treatment, originSiteID, destSiteID, destPlotID, Species) %>%
  summarise(r_plot  = log10(mean(Rel_Cover_final_adj) / mean(Rel_Cover_initial_adj))) %>%
  filter(is.finite(r_plot))%>%
  ungroup()

#calculate r at the plot level
data_site <- data_threshold%>%
   group_by(Region, Treatment, originSiteID, destSiteID, Species) %>%
  summarise(r_site  = log10(mean(Rel_Cover_final_adj) / mean(Rel_Cover_initial_adj)),
            n_plots= n()) %>%
  filter(is.finite(r_site))


dim(data_plot) # 15723     7
dim(data_site) #3591    7

both <- data_plot %>%
  ungroup() %>%
  left_join(data_site %>% ungroup(),
            by = c("Region", "Treatment", "originSiteID", "destSiteID", "Species"))

dim(both)  # doit avoir autant de lignes que data_plot (15723)

ggplot(both, aes(x = r_site, y = r_plot)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = n_plots), alpha = 0.4, size = 1) +
  scale_colour_viridis_c() +
  labs(x = "r site",
       y = "r plot",
       colour = "Nb de plots") +
 facet_wrap(~ Treatment)+
  theme_minimal()

  
#add metadata
data_plot2 <- data_plot %>%
   left_join(data)

cor.test(data_plot2$r_plot, data_plot2$PlotSize, method = "spearman")
cor.test(data_plot2$r_plot, data_plot2$delta_years, method = "spearman")
cor.test(data_plot2$r_plot, data_plot2$Elevation, method = "spearman")


```


##linear models per treatment 

regarder lmer(r_ij  ~ species (1/Region))

```{r}

data_plot <- transform(data_plot2,
    delta_years_cs=scale(delta_years),
    Elevation_cs=scale(Elevation))

write.csv(data_plot, "./intermediate data/r_plot.csv")


mod_warm <- lmer(r_plot ~ Species  + delta_years_cs + Elevation_cs + (1|Region/ destSiteID), data=data_plot %>% filter(Treatment =="warm"))

mod_CO <- lmer(r_plot ~ Species  + delta_years_cs + Elevation_cs + (1|Region / destSiteID), data=data_plot %>% filter(Treatment =="Control_origin"))

mod_CD <- lmer(r_plot ~ Species  + delta_years_cs + Elevation_cs + (1|Region/ destSiteID), data=data_plot %>% filter(Treatment =="Control_dest"))

# data_plot %>% filter(Treatment == "Control_dest") %>%
#   distinct(Region, destSiteID) %>% count(Region) most destSites = 1
# 
# mod_all <- lmer(r_plot ~ Species * Treatment + delta_years_cs + Elevation_cs +
#                   (1 | Region/destSiteID),
#                 data = data_plot)
#
#emmeans(mod_all, pairwise ~ Treatment | Species)

summary(mod_warm)
summary(mod_CO)
summary(mod_CD)


# dd_select_r <- dredge(M1_r,  evaluate = TRUE,  rank = "AICc")
# nrow(dd_select_r[dd_select_r$delta<2,]) #4 modèles avec AIC avec une diff de moins de 2
# best_M1_r <- get.models(dd_select_r, subset =1) [[1]] 
# best_M1_r 

#rsidual homogeneity -assumption of constant variance
plot(mod_warm)
plot(mod_CO)
plot(mod_CD)

#residuals are normally distributed?
qqnorm(residuals(mod_warm))
qqnorm(residuals(mod_CO))
qqnorm(residuals(mod_CD))
```


## emmeans per model

```{r}

em_species_warm <- emmeans(mod_warm, ~ Species) %>%
  as.data.frame() %>%
  arrange(emmean) %>%
  mutate(
    Treatment = "warm",
    statut_W = case_when(
      emmean > 0 & asymp.LCL>0 ~ "winner", 
      emmean <0 & asymp.UCL <0 ~"loser", 
        TRUE ~"stable"))


em_species_CO <- emmeans(mod_CO, ~ Species) %>%
  as.data.frame() %>%
  arrange(emmean) %>%
  mutate(
     Treatment = "control_origin",
    statut_CO = case_when(
      emmean > 0 & asymp.LCL>0 ~ "winner", 
      emmean <0 & asymp.UCL <0 ~"loser", 
        TRUE ~"stable"))


em_species_CD <- emmeans(mod_CD, ~ Species) %>%
  as.data.frame() %>%
  arrange(emmean) %>%
  mutate(
     Treatment = "control_dest",
    statut_CD = case_when(
      emmean > 0 & asymp.LCL>0 ~ "winner", 
      emmean <0 & asymp.UCL <0 ~"loser", 
        TRUE ~"stable"))

em_CO <- em_species_CO %>%
  dplyr::select(Species,
         Treatment_CO= Treatment,
         emmean_CO  = emmean,
         LCL_CO     = asymp.LCL,
         UCL_CO    = asymp.UCL,
         statut_CO)

em_CD <- em_species_CD %>%
  dplyr::select(Species,
          Treatment_CD= Treatment,
         emmean_CD  = emmean,
         LCL_CD     = asymp.LCL,
         UCL_CD     = asymp.UCL,
         statut_CD)

em_W <- em_species_warm %>%
  dplyr::select(Species,
           Treatment_W = Treatment,
         emmean_W   = emmean,
         LCL_W      = asymp.LCL,
         UCL_W      = asymp.UCL,
         statut_W)

em_species_all <- em_CO %>%
  full_join(em_CD,  by = "Species") %>%
  mutate(statut_CO = str_replace_na(statut_CO), 
         statut_CD = str_replace_na(statut_CD))%>%
  full_join(em_W,   by = "Species")%>%
      mutate(statut_W = str_replace_na(statut_W), 
              statut_CD = str_replace_na(statut_CD), 
             statut_CO= str_replace_na(statut_CO))

table(em_CO$statut_CO)
table(em_CD$statut_CD)
table(em_W$statut_W)

```

##plot winners & losers
```{r}
#inversion
em_species <- em_species_all %>%
  mutate(
    inversion_CO_W = case_when(
      statut_W == "winner" & statut_W == "loser"  ~ "winner → loser",
      statut_W == "loser"  & statut_W == "winner" ~ "loser → winner",
      statut_W == "winner" & statut_W == "stable"  ~ "winner → stable",
      statut_W == "loser"  & statut_W == "stable" ~ "loser → stable",
      statut_W == "winner" & statut_W == "NA" ~ "winner → extinct",
      statut_W == "loser" & statut_W == "NA" ~ "loser → extinct",
      statut_W == "NA" & statut_W == "winner" ~ "colonizer winner",
      statut_W == "NA" & statut_W == "loser" ~ "colonizer loser",
      TRUE                                        ~ "conserved"),
    inversion_CD_W = case_when(
      statut_CD == "winner" & statut_W == "loser"  ~ "winner → loser",
      statut_CD == "loser"  & statut_W == "winner" ~ "loser → winner",
      statut_CD == "winner" & statut_W == "stable"  ~ "winner → stable",
      statut_CD == "loser"  & statut_W == "stable" ~ "loser → stable",
      statut_CD == "winner" & statut_W == "NA" ~ "winner → extinct",
       statut_CD == "loser" & statut_W == "NA" ~ "loser → extinct",
      statut_CD == "NA" & statut_W == "winner" ~ "colonizer winner",
      statut_CD == "NA" & statut_W == "loser" ~ "colonizer loser",
      TRUE                                        ~ "conserved"),
    inversion_CD_CO = case_when(
      statut_CD == "winner" & statut_W == "loser"  ~ "winner → loser",
      statut_CD == "loser"  & statut_W == "winner" ~ "loser → winner",
      statut_CD == "winner" & statut_W == "stable"  ~ "winner → stable",
      statut_CD == "loser"  & statut_W == "stable" ~ "loser → stable",
      statut_CD == "winner" & statut_W == "NA" ~ "winner → extinct",
       statut_CD == "loser" & statut_W == "NA" ~ "loser → extinct",
      statut_CD == "NA" & statut_W == "winner" ~ "colonizer winner",
      statut_CD == "NA" & statut_W == "loser" ~ "colonizer loser",
      TRUE                                        ~ "conserved"))

# conserved         : espèces CO présentes en Warm avec même statut
# winner → loser    : espèces gagnantes en CO qui perdent sous Warm
# loser → winner    : espèces perdantes en CO qui gagnent sous Warm
# winner → stable   : espèces gagnantes en CO qui se stabilisent
# loser → stable    : espèces perdantes en CO qui se stabilisent
# winner → extinct  : espèces gagnantes en CO absentes sous Warm
# loser → extinct   : espèces perdantes en CO absentes sous Warm
# colonizer winner  : espèces absentes en CO, winners sous Warm
# colonizer loser   : espèces absentes en CO, losers sous Warm


# Panel CO
df_CO <- em_species %>%
  filter(!is.na(emmean_CO)) %>%
  dplyr::select(Species, 
                emmean = emmean_CO,
                LCL = LCL_CO,
                UCL = UCL_CO, 
                statut = statut_CO) %>%
  mutate(Treatment = "Control_origin",
         colour_var = statut)

# Panel CD
df_CD <- em_species %>%
  filter(!is.na(emmean_CD)) %>%
  dplyr::select(Species, 
         emmean = emmean_CD,
         LCL = LCL_CD, 
         UCL = UCL_CD,
         statut = statut_CD, 
         inversion_CD_W) %>%
  mutate(Treatment  = "Control_dest",
         colour_var = statut)

# Panel Warm 
df_W <- em_species %>%
  filter(!is.na(emmean_W)) %>%
  dplyr::select(Species, 
                emmean = emmean_W,
                LCL = LCL_W, 
                UCL = UCL_W,
                statut = statut_W, 
                inversion_CO_W, inversion_CD_W) %>%
  mutate(Treatment ="warm", 
         colour_var = statut)

df_all <- bind_rows(df_CO, df_CD, df_W) %>%
  mutate(Treatment = factor(Treatment,
                       levels = c("Control_origin",
                                  "Control_dest",
                                  "warm")))

cols <- c(
  # Statuts pour CO
  "winner"           = "#adf452",
  "stable"           = "grey70",
  "loser"            = "#f452ad",
  # Inversions pour CD et Warm
  "conserved"        = "grey70",
  "winner → loser"   = "red",
  "loser → winner"   = "green",
  "winner → stable"  = "blue",
  "loser → stable"   = "lightblue",
  "winner → extinct" = "darkblue",
  "loser → extinct"  = "black",
  "colonizer winner" = "darkgreen",
  "colonizer loser"  = "orange")


#order before plot
species_in_CO <- df_all %>%
  filter(Treatment == "Control_origin") %>%
  arrange(emmean) %>%
  pull(Species)

species_in_cD_and_CO <- df_all %>%
  filter(Treatment %in% c("Control_origin","Control_dest")) %>%
  arrange(emmean) %>%
  distinct(Species)%>%
  pull(Species)

species_not_in_CO <- df_all %>%
  filter(!Species %in% species_in_CO) %>%
  distinct(Species) %>%
  pull(Species)


all_species_order <- c(species_in_CO, species_not_in_CO)

data_status<- df_all %>%
  arrange(emmean) %>%
  mutate(Species   = factor(Species,
                       levels = all_species_order)) %>%
  mutate(Treatment = dplyr::recode(Treatment, 
                                   "Control_origin" = "Control\n origin",
                                   "Control_dest" = "Control\n dest", 
                                   "warm"= "Warm"))

fig <- ggplot(data_status,
              aes(y      = Species,
                  x      = emmean,
                  xmin   = LCL,
                  xmax   = UCL,
                  colour = statut)) +
  geom_pointrange(linewidth = 0.4, size = 0.3, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey30", linewidth = 0.5) +
  # Ligne de séparation espèces CO vs absentes CO
  geom_hline(
    yintercept = length(species_in_CO) + 0.5,
    linetype   = "dotted",
    colour     = "grey50",
    linewidth  = 0.5
  ) +
  #ligne spearation entre les espèces presentes dans CO et CD mais pas Warm
  geom_hline(
    yintercept = length(species_in_cD_and_CO) +0.5,
    linetype= "dotted",
    colour= "grey50", 
    linewidth = 0.5
  )+
  scale_colour_manual(values = cols, na.value = "grey90") +
  facet_wrap(~ Treatment, ncol = 3) +
  labs(
    x        = "Growth rate",
    y        = "Species",
    colour   = NULL,
  ) +
   theme(
    legend.position  = "bottom",
    legend.text = element_text(size= 20), 
    axis.text.y      = element_blank(),
    axis.text.x      = element_text(size = 20),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size =30),
    strip.background = element_blank(), 
    legend.background = element_blank()
  )

fig

ggsave("Figures/distribution_win_lose.jpeg", plot= fig, height= 12, width = 10)

#data of winners and losers
write.csv(data_status %>% dplyr::select(-inversion_CO_W, -inversion_CD_W), "./intermediate data/winners_losers_r_plot_rel_abund.csv")
```


#Abundance per region

*nice little example*

x <- c(1, NA)
sum(x, na.rm = TRUE)/length(x)
[1] 0.5
sum(x, na.rm = TRUE)/length(x[!is.na(x)])
[1] 1
mean(x, na.rm = TRUE)
mean(x, na.rm = F)
  
##without NA
```{r}

#winners and losers data calculated at plot level for each region 
#[pending]

#relative abundance of species at plot level
data_abund <- read.csv("./intermediate data/r_plot.csv")%>% 
  dplyr::select(-X) %>%
  mutate(Treatment= dplyr::recode(Treatment,  "Control_origin"= "Control\n origin", "Control_dest"="Control\n dest", "warm"="Warm"))  %>%filter(dynamique_statut %in% c("Extinct","Persistent"))

#matrix - mais ca revient au meme que le code en dessous
# ab_matrix   <- data_abund%>%
#   filter(Region =="CH_Calanda")%>%
#   dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
#   mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
#   dplyr::select(-Region, -destSiteID, -destPlotID)%>%
#   tidyr::pivot_wider(id_cols = Species, 
#               names_from = Region_Site_Plot, 
#               values_from = Rel_Cover_initial,
#               values_fill = NA)%>%
#   left_join(data_status %>% dplyr::select(Species, statut, Treatment))%>%
#   relocate(statut, .after = Species)%>%
#   relocate(Treatment, .after=Species)
#   
# plot_names <- colnames(ab_matrix %>% dplyr::select(-Species, -statut, -Treatment))
# 
# ab_matrix   <- ab_matrix  %>%
#   mutate(mean_abund = rowMeans((across(all_of(plot_names))), na.rm=T))%>%
#   relocate(mean_abund, .after=statut)


mean_abund_region <- data_abund %>%
  group_by(Region, Species) %>%
  summarise(
    mean_abund = mean(Rel_Cover_initial, na.rm = TRUE),
    n_plots    = n_distinct(paste(destSiteID, destPlotID)),
    .groups = "drop"
  ) %>%
  left_join([pending] %>% distinct(Species, statut, Treatment), by = "Species")

cols_statut <- c(loser = "#f452ad", stable = "grey70", winner = "#adf452")

df <- mean_abund_region %>%
  filter(!is.na(statut)) %>%
  mutate(statut = factor(statut, levels = c("loser", "stable", "winner")))

#for plot position
top_whisk <- df %>%
  group_by(Region, Treatment, statut) %>%
  summarise(w = boxplot.stats(mean_abund)$stats[5], .groups = "drop") %>%
  group_by(Region, Treatment) %>%
  summarise(top = max(w), .groups = "drop")

region_rng <- top_whisk %>%
  group_by(Region) %>%
  summarise(rng = max(top), .groups = "drop")

# Dunn + positions x (dodge) et y (empilées au-dessus des moustaches)
stat_test <- df %>%
  group_by(Region, Treatment) %>%
  dunn_test(mean_abund ~ statut, p.adjust.method = "BH") %>%
  add_x_position(x = "Treatment", dodge = 0.8) %>%
  filter(p.adj.signif != "ns") %>%                 # on n'empile que les crochets affichés
  left_join(top_whisk,  by = c("Region", "Treatment")) %>%
  left_join(region_rng, by = "Region") %>%
  group_by(Region, Treatment) %>%
  mutate(y.position = top + rng * 0.09 * row_number()) %>%
  ungroup()

n_reg <- n_distinct(df$Region)

plot_abund <- ggplot(df, aes(x = Treatment, y = mean_abund)) +
  geom_boxplot(aes(fill = statut),
               outliers = FALSE, width = 0.7, linewidth = 0.3,
               position = position_dodge(0.8)) +
  stat_pvalue_manual(stat_test, label = "p.adj.signif",
                     tip.length = 0.01, size = 3.2, bracket.size = 0.3) +
  scale_fill_manual(values = cols_statut, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.08))) +
  facet_wrap(~ Region, scales = "free_y", ncol = 4) +
  labs(x = NULL, y = "Mean relative abundance") +
  theme_minimal(base_size = 10) +
  theme(legend.position    = "top",
        strip.text         = element_text(face = "bold", hjust = 0),
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        panel.spacing      = unit(0.8, "lines"))

plot_abund

# #CO
# ab_matrix_CO   <- data_abund%>%
#   filter(Treatment =="Control\n origin") %>%
#   dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
#   mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
#   dplyr::select(-Region, -destSiteID, -destPlotID)%>%
#   tidyr::pivot_wider(id_cols = Species, 
#               names_from = Region_Site_Plot, 
#               values_from = Rel_Cover_initial,
#               values_fill = NA)%>%
#   left_join(data_status %>% filter(Treatment=="Control\n origin") %>% dplyr::select(Species, statut))%>%
#   relocate(statut, .after = Species)
#   
# plot_names <- colnames(ab_matrix_CO %>% dplyr::select(-Species, -statut))
# 
# ab_matrix_CO   <- ab_matrix_CO  %>%
#   mutate(mean_abund = rowMeans((across(all_of(plot_names))), na.rm=T))%>%
#   relocate(mean_abund, .after=statut)
# 
# #CD
# ab_matrix_CD   <- data_abund%>%
#   filter(Treatment =="Control\n dest") %>%
#   dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
#   mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
#   dplyr::select(-Region, -destSiteID, -destPlotID)%>%
#   tidyr::pivot_wider(id_cols = Species, 
#               names_from = Region_Site_Plot, 
#               values_from = Rel_Cover_initial,
#               values_fill = NA)%>%
#   left_join(data_status %>% filter(Treatment=="Control\n dest")%>% dplyr::select(Species, statut))%>%
#   relocate(statut, .after = Species)
#   
# plot_names_CD <- colnames(ab_matrix_CD %>% dplyr::select(-Species, -statut))
# 
# ab_matrix_CD   <- ab_matrix_CD  %>%
#     mutate(mean_abund = rowMeans((across(all_of(plot_names_CD))), na.rm=T))%>%
#   relocate(mean_abund, .after=statut)
# 
# #Warm
# ab_matrix_warm   <- data_abund%>%
#   filter(Treatment =="Warm") %>%
#   dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
#   mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
#   dplyr::select(-Region, -destSiteID, -destPlotID)%>%
#   tidyr::pivot_wider(id_cols = Species, 
#               names_from = Region_Site_Plot, 
#               values_from = Rel_Cover_initial,
#               values_fill = NA)%>%
#   left_join(data_status%>% filter(Treatment=="Warm") %>% dplyr::select(Species, statut))%>%
#   relocate(statut, .after = Species)
#   
# plot_names_warm <- colnames(ab_matrix_warm %>% dplyr::select(-Species, -statut))
# 
# ab_matrix_warm   <- ab_matrix_warm  %>%
#     mutate(mean_abund = rowMeans((across(all_of(plot_names_warm))), na.rm=T))%>%
#   relocate(mean_abund, .after=statut)
# 
# #merge
# Abund<- rbind(ab_matrix_CO %>% dplyr::select(Species, statut, mean_abund)%>% mutate(Treatment ="Control\n origin"), 
#           ab_matrix_CD  %>% dplyr::select(Species, statut, mean_abund)%>% mutate(Treatment ="Control\n dest"),
#           ab_matrix_warm  %>% dplyr::select(Species, statut, mean_abund)%>% mutate(Treatment ="Warm"))


# kruskal_test(Abund %>% filter(Treatment=="Control\n origin"), mean_abund ~statut)
# kruskal_test(Abund %>% filter(Treatment=="Control\n dest"), mean_abund ~statut)
# kruskal_test(Abund %>% filter(Treatment=="Warm"), mean_abund ~statut)
# 
# dunn_CO <-dunn_test(Abund %>% filter(Treatment=="Control\n origin"), formula = mean_abund ~statut, p.adjust.method = "BH") %>%
#   add_xy_position(x = "statut")%>%
#   dplyr::select(-y.position)%>%
#   mutate(y.position = c(0.15,0.2,0.85))%>% mutate(Treatment ="Control\n origin")
# 
# dunn_CD <- dunn_test(Abund %>% filter(Treatment=="Control\n dest"), formula = mean_abund ~statut, p.adjust.method = "BH")%>%
#   add_xy_position(x = "statut")%>%
#   dplyr::select(-y.position)%>%
#   mutate(y.position = c(0.15,0.2,0.175))%>% mutate(Treatment ="Control\n dest")
# 
# dunn_test(Abund %>% filter(Treatment=="Warm"), formula = mean_abund ~statut, p.adjust.method = "BH")
# 
# dunn_abund <- rbind(dunn_CO, dunn_CD)
# 
# KW_warm <- data.frame(
#   Treatment = "Warm",  
#   x = 3,
#   y = 0.2,
#   label = "ns"
# )
# 
# 
# 
# 
# #individual plots
# (plot_abund_CO <- Abund %>%
#     filter(Treatment =="Control\n origin")%>%
#  ggplot() +
#   aes(x = statut, y = mean_abund, fill = statut) +
#   geom_boxplot(outliers = F) +
#   scale_fill_manual(values = c( 
#   "winner" = "#adf452",
#   "stable"           = "grey70",
#   "loser"            = "#f452ad")) +
#     labs(x= NULL, y= "Mean abundance")+
#       stat_pvalue_manual(dunn_CO, 
#                        label="p.adj.signif",
#                        tip.length = 0,
#                        hide.ns = T)+
#   theme_minimal())
# 
# (plot_abund_CD <- Abund %>%
#     filter(Treatment =="Control\n dest")%>%
#  ggplot() +
#   aes(x = statut, y = mean_abund, fill = statut) +
#   geom_boxplot(outliers = F) +
#   scale_fill_manual(values = c( 
#   "winner" = "#adf452",
#   "stable"           = "grey70",
#   "loser"            = "#f452ad")) +
#     labs(x= NULL, y= "Mean abundance")+
#     stat_pvalue_manual(dunn_CD, 
#                        label="p.adj.signif",
#                        tip.length = 0,
#                        hide.ns = T)+
#   theme_minimal())
# 
# (plot_abund_W <- Abund %>%
#     filter(Treatment =="Warm")%>%
#  ggplot() +
#   aes(x = statut, y = mean_abund, fill = statut) +
#   geom_boxplot(outliers = F) +
#   scale_fill_manual(values = c( 
#   "winner" = "#adf452",
#   "stable"           = "grey70",
#   "loser"            = "#f452ad")) +
#     labs(x= NULL, y= "Mean abundance")+
#      geom_text(data = KW_warm,
#             aes(x = x, y = y, label = label),
#             inherit.aes = FALSE)+
#   theme_minimal())


```

##with NA
```{r eval=FALSE, include=FALSE}

#winners and losers data calculated at plot level
data_status <- read.csv("./intermediate data/winners_losers_r_plot_rel_abund.csv") %>% dplyr::select(-X) 

#relative abundance of species at plot level
data_abund <- read.csv("./intermediate data/r_plot.csv")%>% 
  dplyr::select(-X) %>%
  mutate(Treatment= dplyr::recode(Treatment,  "Control_origin"= "Control\n origin", "Control_dest"="Control\n dest", "warm"="Warm"))  %>%filter(dynamique_statut %in% c("Extinct","Persistent"))

#CO
ab_matrix_CO   <- data_abund%>%
  filter(Treatment =="Control\n origin") %>%
  dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
  mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
  dplyr::select(-Region, -destSiteID, -destPlotID)%>%
  tidyr::pivot_wider(id_cols = Species, 
              names_from = Region_Site_Plot, 
              values_from = Rel_Cover_initial,
              values_fill = NA)%>%
  left_join(data_status %>% filter(Treatment=="Control\n origin") %>% dplyr::select(Species, statut))%>%
  relocate(statut, .after = Species)
  
plot_names <- colnames(ab_matrix_CO %>% dplyr::select(-Species, -statut))
ncol_CO <- ncol(ab_matrix_CO) -2
ab_matrix_CO   <- ab_matrix_CO  %>%
  mutate(mean_abund = rowSums((across(all_of(plot_names))), na.rm=T)/ncol_CO)%>%
  relocate(mean_abund, .after=statut)

#CD
ab_matrix_CD   <- data_abund%>%
  filter(Treatment =="Control\n dest") %>%
  dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
  mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
  dplyr::select(-Region, -destSiteID, -destPlotID)%>%
  tidyr::pivot_wider(id_cols = Species, 
              names_from = Region_Site_Plot, 
              values_from = Rel_Cover_initial,
              values_fill = NA)%>%
  left_join(data_status %>% filter(Treatment=="Control\n dest")%>% dplyr::select(Species, statut))%>%
  relocate(statut, .after = Species)
  
plot_names_CD <- colnames(ab_matrix_CD %>% dplyr::select(-Species, -statut))
ncol_CD <- ncol(ab_matrix_CD) -2
ab_matrix_CD   <- ab_matrix_CD  %>%
    mutate(mean_abund = rowSums((across(all_of(plot_names_CD))), na.rm=T)/ncol_CD)%>%
  relocate(mean_abund, .after=statut)

#Warm
ab_matrix_warm   <- data_abund%>%
  filter(Treatment =="Warm") %>%
  dplyr::select(Region, destSiteID, destPlotID, Species, Rel_Cover_initial) %>%
  mutate(Region_Site_Plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
  dplyr::select(-Region, -destSiteID, -destPlotID)%>%
  tidyr::pivot_wider(id_cols = Species, 
              names_from = Region_Site_Plot, 
              values_from = Rel_Cover_initial,
              values_fill = NA)%>%
  left_join(data_status%>% filter(Treatment=="Warm") %>% dplyr::select(Species, statut))%>%
  relocate(statut, .after = Species)
  
plot_names_warm <- colnames(ab_matrix_warm %>% dplyr::select(-Species, -statut))
ncol_W <- ncol(ab_matrix_warm) -2
ab_matrix_warm   <- ab_matrix_warm  %>%
    mutate(mean_abund = rowSums((across(all_of(plot_names_warm))), na.rm=T)/ncol_W)%>%
  relocate(mean_abund, .after=statut)

#merge
Abund<- rbind(ab_matrix_CO %>% dplyr::select(Species, statut, mean_abund)%>% mutate(Treatment ="Control\n origin"), 
          ab_matrix_CD  %>% dplyr::select(Species, statut, mean_abund)%>% mutate(Treatment ="Control\n dest"),
          ab_matrix_warm  %>% dplyr::select(Species, statut, mean_abund)%>% mutate(Treatment ="Warm"))

#test
kruskal_test(Abund %>% filter(Treatment=="Control\n origin"), mean_abund ~statut) #ns
kruskal_test(Abund %>% filter(Treatment=="Control\n dest"), mean_abund ~statut)
kruskal_test(Abund %>% filter(Treatment=="Warm"), mean_abund ~statut)

dunn_W <-dunn_test(Abund %>% filter(Treatment=="Warm"), formula = mean_abund ~statut, p.adjust.method = "BH") %>%
  add_xy_position(x = "statut")%>%
  dplyr::select(-y.position)%>%
  mutate(y.position = c(0.015,0.02,0.085))%>% mutate(Treatment ="Warm")

dunn_CD <- dunn_test(Abund %>% filter(Treatment=="Control\n dest"), formula = mean_abund ~statut, p.adjust.method = "BH")%>%
  add_xy_position(x = "statut")%>%
  dplyr::select(-y.position)%>%
  mutate(y.position = c(0.015,0.02,0.0175))%>% mutate(Treatment ="Control\n dest")

dunn_abund <- rbind(dunn_W, dunn_CD)

KW_CO <- data.frame(
  Treatment = "Control\n origin",  
  x = 3,
  y = 0.02,
  label = "ns"
)


#plot
(plot_abund <- Abund %>%
 ggplot() +
  aes(x = statut, y = mean_abund, fill = statut) +
  geom_boxplot(outliers = F) +
  scale_fill_manual(values = c( 
  "winner" = "#adf452",
  "stable"           = "grey70",
  "loser"            = "#f452ad")) +
    labs(x= NULL, y= "Mean abundance")+
    facet_wrap(~Treatment)+
     geom_text(data = KW_CO,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE) +
      stat_pvalue_manual(dunn_abund, 
                       label="p.adj.signif",
                       tip.length = 0,
                       hide.ns = T)+
  theme_minimal())

```


#Restrictiveness
```{r}

#relative abundance of species at plot level
data_abund <- read.csv("./intermediate data/r_plot.csv")%>% 
  dplyr::select(-X) %>%
  mutate(Treatment= dplyr::recode(Treatment,  "Control_origin"= "Control\n origin", "Control_dest"="Control\n dest", "warm"="Warm"))   %>%filter(dynamique_statut %in% c("Extinct","Persistent"))

############################################### ~ REGION LEVEL ~ ###################################################

#warm region level
pres_matrix_region <- data_abund %>%
  mutate(
    Species    = gsub("_", " ", Species)) %>%
  distinct(Region, Species) %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols     = Region,
    names_from  = Species,
    values_from = present,
    values_fill = 0,
    values_fn   = max
  ) %>%
  tibble::column_to_rownames("Region") %>%
  as.matrix()

dim(pres_matrix_region_warm)

# restrictedness region level
restrict_region <- funrar::restrictedness(
  pres_matrix = pres_matrix_region,
  relative    = TRUE
) %>%
  rename(Species = species)%>%
  rename(Ri_region = Ri)

############################################### ~ PLOT LEVEL ~ ###################################################
pres_matrix_plot <- data_abund %>%
  mutate(
    Species    = gsub("_", " ", Species)) %>%
  distinct(Region, destSiteID, destPlotID, Species) %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols     = c("Region","destSiteID", "destPlotID"),
    names_from  = Species,
    values_from = present,
    values_fill = 0,
    values_fn   = max
  ) %>%
  mutate(
    Region_site_plot = paste0(Region, "_", destSiteID, "_", destPlotID))%>%
  dplyr::select(-Region, -destSiteID, -destPlotID)%>%
  tibble::column_to_rownames("Region_site_plot") %>%
  as.matrix()


#restrictedness plot level
restrict_plot <- funrar::restrictedness(
  pres_matrix = pres_matrix_plot,
  relative    = TRUE
) %>%
  rename(Species = species)%>%
  rename(Ri_plot = Ri)

# Compare with restrictiveness
freq_comparison <- restrict_region%>%
  left_join(restrict_plot, by="Species")

# Corrélations entre les 2 mesures
cor(freq_comparison[, c("Ri_region", "Ri_plot")])

#add winner / loser / stable info
#winners and losers data calculated at plot level
data_status <- read.csv("./intermediate data/winners_losers_r_plot_rel_abund.csv") %>% dplyr::select(-X) 

#plot level
restrict_total <- restrict_plot %>%
  left_join(data_status %>% mutate(Species = gsub("_", " ", Species)), by= "Species")

#stat test : Kruskall-Wallis follwoed by Dunn Test if KW significant
kw_trt <- restrict_total %>%
  group_by(Treatment) %>%
  kruskal_test(Ri_plot ~ statut)
kw_trt

dunn_trt <- restrict_total %>%
  group_by(Treatment) %>%
  dunn_test(Ri_plot ~ statut, p.adjust.method = "BH") %>%
  add_x_position(x = "statut")%>%
   mutate(y.position = rep(c(0.90, 0.80, 0.85), 3))
dunn_trt

#plots
(R_all <- (ggplot(restrict_total)+
    aes(x=statut, y = Ri_plot, fill= statut)+
    geom_boxplot(outliers = F)+
    scale_fill_manual(values= c("#f452ad", "grey", "#adf452"))+
    theme_minimal(base_size = 15)+
      facet_wrap(~Treatment)+
       stat_pvalue_manual(dunn_trt, tip.length = 0,
                     label = "p.adj.signif", hide.ns = TRUE) +
    labs(
      x=NULL,
      y= "Restrictedness",
      title= "plot level")+
    theme(
      axis.title.x = element_blank(),
      legend.title = element_blank()
    )))

#calculate  quantiles pour définir rare / intermédiaire / commun
q_R <- quantile(restrict_total$Ri_plot, probs = c(0.25, 0.75))

Ri_df <- restrict_total %>%
  mutate(
    Restrict= case_when(
      Ri_plot >= q[2] ~ "Rare",
      Ri_plot <=q[1] ~"Common",
      T ~ "average"))

n_rare <- Ri_df%>%
  filter(Restrict == "Rare")%>%
  summarise(
    n_rare = n()) 

n_common <- Ri_df%>%
  filter(Restrict == "Common")%>%
  summarise(
    n_common = n()) 
```

#Traits

##add the traits

```{r}
# Load imputed mean trait dataset (TRY database)
Sp_trait_TRY <- read.csv("new_TRY_imputed_traits_mean.csv", sep = ",", dec = ".") 

# Log-transform continuous traits to normalize distributions
Sp_trait_TRY_trans <- Sp_trait_TRY

Sp_trait_TRY_trans[,4] <- log10(Sp_trait_TRY_trans[,4])    # Maximum height
Sp_trait_TRY_trans[,5] <- log10(Sp_trait_TRY_trans[,5])    # SLA
Sp_trait_TRY_trans[,6] <- log10(Sp_trait_TRY_trans[,6])    # Leaf nitrogen
Sp_trait_TRY_trans[,7] <- log10(Sp_trait_TRY_trans[,7])    # Leaf area
Sp_trait_TRY_trans[,8] <- ifelse(Sp_trait_TRY_trans[,8] == 0, 0.0001, Sp_trait_TRY_trans[,8])  # Avoid log(0) for seed mass

# Center and scale traits
Sp_trait_TRY_trans[,c(4:8)] <- scale(Sp_trait_TRY_trans[,c(4:8)])
```


##join
```{r}
#remove _ 
data_status$Species <- gsub("_", " ",data_status$Species)

data <- data_status %>%
  left_join(Sp_trait_TRY_trans)

#parque si on ne le fait pas il y a un warning
data_scaled <- data %>%
  mutate(
    VH_sc  = scale(VH),
    SLA_sc = scale(SLA),
    LA_sc  = scale(LA),
    SM_sc  = scale(SM),
    LN_sc  = scale(LN)
  ) %>%
  dplyr::select(Treatment, Family, Genus, Species, emmean, VH_sc, SLA_sc, LA_sc, SM_sc, LN_sc)%>%
  filter(!is.na(Family))

##which family is winner / loser? 
# Family et statut sont dans data (avant data_scaled qui retire statut)
family_status <- data %>%
  distinct(Species, Treatment, statut, Family) %>%
  filter(!is.na(Family), !is.na(statut))


family_count <- family_status %>%
  count(Family, Treatment, statut) %>%
  group_by(Family, Treatment) %>%
  mutate(
    n_total = sum(n),
    pct     = n / n_total * 100
  ) %>%
  ungroup()

# Familles avec le plus de winners sous Warm 
family_count %>%
  filter(Treatment == "Warm", statut == "winner") %>%
  arrange(desc(pct)) %>%
   filter(n >= 5) %>%
  dplyr::select(Family, n, n_total, pct) %>%
  head(10)

# Familles avec le plus de losers sous Warm 
family_count %>%
  filter(Treatment == "Warm", statut == "loser") %>%
  arrange(desc(pct)) %>%
   filter(n >= 5) %>%
  dplyr::select(Family, n, n_total, pct) %>%
  head(10)

# ── Figure ────────────────────────────────────────────────────
# Familles avec >= 3 espèces
families_keep <- family_count %>%
  group_by(Family) %>%
  summarise(n_sp = sum(n), .groups = "drop") %>%
  filter(n_sp >= 3) %>%
  pull(Family)

# Compléter les combinaisons manquantes
all_combos_fam <- expand.grid(
  Family    = families_keep,
  Treatment = c("Control\n origin", "Control\n dest", "Warm"),
  statut    = c("winner", "stable", "loser"),
  stringsAsFactors = FALSE
)

family_count_complete <- all_combos_fam %>%
  left_join(
    family_count %>% filter(Family %in% families_keep),
    by = c("Family", "Treatment", "statut")
  ) %>%
  mutate(
    n   = ifelse(is.na(n),   0, n),
    pct = ifelse(is.na(pct), 0, pct),
    statut    = factor(statut,
                       levels = c("winner", "stable", "loser")),
    Treatment = factor(Treatment,
                       levels =  c("Control\n origin", "Control\n dest", "Warm"))
  )

# Ordonner par % winners sous Warm
family_order <- family_count_complete%>%
  filter(Treatment == "Warm", statut == "winner") %>%
  arrange(pct) %>%
  pull(Family)
 family_count_complete <-  family_count_complete %>%
  mutate(Family = factor(Family, levels = family_order))

# Figure
fig_family <- family_count_complete %>%
  filter(n>=3)%>%
  ggplot(aes(x = pct, y = Family, fill = statut)) +
  geom_col(position = "stack") +
  geom_text(
    data =  family_count_complete %>%
      filter(n>= 3)%>%
      group_by(Family, Treatment) %>%
      summarise(n_total = sum(n), .groups = "drop"),
    aes(x = 102, y = Family,
        label = paste0("n=", n_total)),
    inherit.aes = FALSE,
    size = 2.5, hjust = 0, colour = "grey40"
  ) +
  scale_x_continuous(
    limits = c(0, 115),
    labels = function(x) ifelse(x <= 100, paste0(x, "%"), "")
  ) +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  facet_wrap(~ Treatment, ncol = 3,
             labeller = labeller(Treatment = c(
               "Control_origin" = "Control origin",
               "Control_dest"   = "Control dest",
               "warm"           = "Warm"
             ))) +
  labs(
    x        = "% species",
    y        = NULL,
    fill     = NULL  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    axis.text.y      = element_text(size = 12),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )

fig_family
```


##distinctiveness 

cf script Lula

###globale

1 ACP pour les 642 espèces 

####data for PCA
```{r}
#remove _ 
data_status$Species <- gsub("_", " ",data_status$Species) 

#data per treatment for each PCA
data_warm <- data_status %>% filter(Treatment =="warm")
data_warm <- data_warm %>%
  left_join(Sp_trait_TRY_trans) 

length(unique(data_warm$Species)) #503 unique species in warm

data_CO <- data_status %>% filter(Treatment =="Control\n origin")
data_CO<- data_CO %>%
  left_join(Sp_trait_TRY_trans) 

length(unique(data_W$Species)) #465 unique species in CO

data_CD <- data_status %>% filter(Treatment =="Control\n dest")
data_CD <- data_CD %>%
  left_join(Sp_trait_TRY_trans) 

length(unique(data_CD$Species)) #348 nique species in warm

#parque si on ne le fait pas il y a un warning
data_w_scaled <- data_warm %>%
  mutate(
    VH_sc  = scale(VH),
    SLA_sc = scale(SLA),
    LA_sc  = scale(LA),
    SM_sc  = scale(SM),
    LN_sc  = scale(LN)
  ) %>%
  dplyr::select(Treatment, Family, Genus, Species, emmean, VH_sc, SLA_sc, LA_sc, SM_sc, LN_sc)%>%
  filter(!is.na(Family))

data_CO_scaled <- data_CO %>%
  mutate(
    VH_sc  = scale(VH),
    SLA_sc = scale(SLA),
    LA_sc  = scale(LA),
    SM_sc  = scale(SM),
    LN_sc  = scale(LN)
  ) %>%
  dplyr::select(Treatment, Family, Genus, Species, emmean, VH_sc, SLA_sc, LA_sc, SM_sc, LN_sc)%>%
  filter(!is.na(Family))

data_CD_scaled <- data_CD %>%
  mutate(
    VH_sc  = scale(VH),
    SLA_sc = scale(SLA),
    LA_sc  = scale(LA),
    SM_sc  = scale(SM),
    LN_sc  = scale(LN)
  ) %>%
  dplyr::select(Treatment, Family, Genus, Species, emmean, VH_sc, SLA_sc, LA_sc, SM_sc, LN_sc)%>%
  filter(!is.na(Family))

##data global only species and traits les 642 species
data_status$Species<- str_replace(data_status$Species, "[.]", "-")

data_global <- data_status %>%
  distinct(Species) %>%
  left_join(Sp_trait_TRY_trans) 

length(unique(data_global$Species)) #642

data_global_scaled <- data_global %>%
  mutate(
    VH_sc  = scale(VH),
    SLA_sc = scale(SLA),
    LA_sc  = scale(LA),
    SM_sc  = scale(SM),
    LN_sc  = scale(LN)
  ) %>%
  dplyr::select(Family, Genus, Species, VH_sc, SLA_sc, LA_sc, SM_sc, LN_sc)%>%
  filter(!is.na(Family))

length(unique(data_global_scaled$Species)) #639 unique species in warm

```

####PCA
```{r}

#data is data_global_scaled

pca_global <- PCA(X = data_global_scaled[4:8], scale.unit = T, graph = F)
pca_global$eig
pca_global$var$contrib
fviz_eig(pca_global)
fviz_pca(pca_global)
fviz_contrib(pca_global, choice="var", axes=1)
fviz_contrib(pca_global, choice="var", axes=2)
fviz_contrib(pca_global, choice="var", axes=3)
fviz_contrib(pca_global, choice="var", axes=4, size= 5)
fviz_contrib(pca_global, choice="var", axes=5)
loading_df <- as.data.frame(pca_global$var$coord)
loading_df$Trait <- rownames(pca_global$var$coord)
loading_df <- separate(data = loading_df,col = Trait, into = c("Trait", "scale"), sep="_")
loading_df <- loading_df %>% dplyr::select(-scale) %>% rename(
  Dim1 = Dim.1,
  Dim2 = Dim.2,
  Dim3 = Dim.3,
  Dim4 = Dim.4,
  Dim5 = Dim.5
)

#calcul de distinctiveness sur cette ACP
scores_df <- as.data.frame(pca_global$ind$coord)
rownames(scores_df) <- data_global_scaled$Species
species_names    <- rownames(scores_df)
Distance <- compute_dist_matrix(scores_df,
                                metric = "euclidean",
                                center = FALSE,
                                scale  = FALSE)
rownames(Distance) <- colnames(Distance) <- species_names
# rescale distances to [0, 1]
Distance_scaled <- Distance / max(Distance, na.rm = TRUE)

dim(Distance_scaled)
di_df <- funrar::distinctiveness_global(dist_obj = Distance_scaled, di_name = "global_di") 

# quantiles pour définir rare / intermédiaire / commun
q <- quantile(di_df$global_di, probs = c(0.25, 0.75))

di_df <- di_df %>%
  mutate(
    Rarity= case_when(
      global_di >= q[2] ~ "Frare",
      global_di <=q[1] ~"Fcommon",
      T ~ "average"))

n_rare <- di_df%>%
  filter(Rarity == "Frare")%>%
  summarise(
    n_rare = n()) 

n_Wmmon <- di_df%>%
  filter(Rarity == "Fcommon")%>%
  summarise(
    n_Wmmon = n()) 

scores_df$Species <- rownames(scores_df)
di_df <- di_df %>% 
  rename(Species=species)

rarity_df <- scores_df %>% left_join(di_df, by="Species")
rarity_df <- rarity_df %>% rename(
  Dim1 = Dim.1,
  Dim2 = Dim.2,
  Dim3 = Dim.3,
  Dim4 = Dim.4,
  Dim5 = Dim.5
)

(tab <- table(rarity_df$Rarity))

scale_factor <- 5
eig <- round(pca_global$eig[1:2, 2], 1)

Fig1 <- ggplot() +
  geom_point(data = rarity_df, aes(x = Dim1, y = Dim2, color= global_di), alpha = 0.7, size = 1.5) +
   scale_colour_gradient(
    low = "#ffffbf",  high = "red",
    name = "Global\n distinctiveness"
  )+
  geom_segment(data = loading_df,
               aes(x = 0, y = 0, xend = Dim1 * scale_factor, yend = Dim2 * scale_factor),
               arrow = arrow(length = unit(0.25, "cm")), color = "black") +
  geom_text(data = loading_df,
            aes(x = Dim1 * scale_factor * 1.1, y = Dim2 * scale_factor * 1.1, label = Trait),
            color = "black", size = 6) +
  theme_minimal() +
  theme(panel.border = element_rect(fill=NULL),
        axis.title.x = element_text(size = 20), 
        axis.title.y = element_text(size=20), 
        axis.text.x = element_text(size = 15),
         axis.text.y = element_text(size = 15),
        legend.title = element_text(size=20),
        legend.text = element_text(size = 15))+
 labs(x = paste0("Dim1 (", eig[1], " %)"), y = paste0("Dim2 (", eig[2], " %)"))

Fig1 
ggsave("Figures/PCA/PCA_rotated_global_distinctiveness_unrotated.png", plot = Fig1, width=6, height = 4)
```



##rotated PCA
```{r eval=FALSE, include=FALSE}
nfactors <- paran(data_global_scaled[4:8])$Retained #Horn analysis #nfactors=2
PCA_globale <- psych::principal(scale(data_global_scaled[4:8]), nfactors=nfactors, #nfactors=2
                                 rotate="varimax", covar = T)
print(PCA_globale, digits = 3, sort = TRUE)  # loadings, communalités, variance expliquée
PCA_globale$loadings   # matrice traits x composantes rotées (RC1, RC2)
head(PCA_globale$scores)  # scores des 642 espèces sur RC1, RC2

#scores of species
scores_df <- as.data.frame(PCA_globale$scores)
colnames(scores_df) <- c("Dim1", "Dim2") 
scores_df$Species <- data_global_scaled$Species

#loadings
loading_df <- as.data.frame(unclass(PCA_globale$loadings))
colnames(loading_df) <-c("Dim1", "Dim2") 
loading_df$Trait <- rownames(PCA_globale$loadings)
loading_df <- separate(data = loading_df,col = Trait, into = c("Trait", "scale"), sep="_")
loading_df <- loading_df %>% dplyr::select(-scale)

# Facteur d'échelle pour que les flèches soient visibles à côté du nuage de points
scale_factor <- 4

psych::fa.diagram(PCA_globale)
psych::biplot.psych(PCA_globale, choose = c(1,2))

PCA_globale$Vaccounted  #variance total =5 (5 traits): donc 1.84/8=0.37

#plot PCA global
Fig1 <- ggplot() +
  geom_point(data = rarity_df, aes(x = Dim1, y = Dim2, color= global_di), alpha = 0.7, size = 1.5) +
   scale_Wlour_gradient(
    low = "#ffffbf",  high = "red",
    name = "Global\n distinctiveness"
  )+
  geom_segment(data = loading_df,
               aes(x = 0, y = 0, xend = Dim1 * scale_factor, yend = Dim2 * scale_factor),
               arrow = arrow(length = unit(0.25, "cm")), color = "black") +
  geom_text(data = loading_df,
            aes(x = Dim1 * scale_factor * 1.1, y = Dim2 * scale_factor * 1.1, label = Trait),
            color = "black", size = 6) +
  theme_minimal() +
  theme(panel.border = element_rect(fill=NULL),
        axis.title.x = element_text(size = 20), 
        axis.title.y = element_text(size=20), 
        axis.text.x = element_text(size = 15),
         axis.text.y = element_text(size = 15),
        legend.title = element_text(size=20),
        legend.text = element_text(size = 15))+
  labs(x = "Dim1 (37.0 %)", y = "Dim2 (27.4 %)")

Fig1 
ggsave("Figures/PCA/PCA_rotated_global_distinctiveness.png", plot = Fig1, width=6, height = 4)

#calcul de distinctiveness sur cette ACP rotated
scores_df <- as.data.frame(PCA_globale$scores)
colnames(scores_df) <- c("Dim1", "Dim2")
rownames(scores_df) <- data_global_scaled$Species
species_names    <- rownames(scores_df)

#data<- data_global_scaled[4:8]
#rownames(data) <- data_global_scaled$Species
Distance <- compute_dist_matrix(scores_df, #(si on reste sur les deux dims de la rotated PCA)
                                metric = "euclidean",
                                center = FALSE,
                                scale  = FALSE)
rownames(Distance) <- colnames(Distance) <- species_names
dim(Distance)
di_df <- funrar::distinctiveness_global(dist_obj = Distance, di_name = "global_di") 

# quantiles pour définir rare / intermédiaire / commun
q <- quantile(di_df$global_di, probs = c(0.25, 0.75))

di_df <- di_df %>%
  mutate(
    Rarity= case_when(
      global_di >= q[2] ~ "Frare",
      global_di <=q[1] ~"Fcommon",
      T ~ "average"))

n_rare <- di_df%>%
  filter(Rarity == "Frare")%>%
  summarise(
    n_rare = n()) 

n_Wmmon <- di_df%>%
  filter(Rarity == "Fcommon")%>%
  summarise(
    n_Wmmon = n()) 

scores_df$Species <- rownames(scores_df)
di_df <- di_df %>% 
  rename(Species=species)

rarity_df <- scores_df %>% left_join(di_df, by="Species")

(tab <- table(rarity_df$Rarity))
```




#Distinctiveness & Resrtictiveness - Region
```{r}
#distinctiveness
names(rarity_df)
dim(rarity_df)

#restrictiveness 
names(Ri_df)
dim(Ri_df)

#merge
Functional_rarity <- 
  Ri_df %>% dplyr::select(Species, Ri_plot, emmean, statut, Treatment, Restrict)%>%
  left_join(rarity_df, by= "Species")%>%
  mutate(Restrict2 = log10(1-Ri_plot))
dim(Functional_rarity)

#plot
ggplot(Functional_rarity) +
  aes(x = Restrict2, y = global_di, colour = statut) +
  geom_point() +
  scale_colour_manual(values= c("#f452ad", "grey", "#adf452"))+
  theme_minimal() +
  facet_wrap(vars(Treatment))+
#  xlim(0,1)+
  ylim(0,1)



############
library(dplyr)
library(ggplot2)

# 1. Quantile thresholds per Treatment (on the ORIGINAL Ri scale)
#    distinct() so that species duplicated within a treatment don't weight the quantiles
thresholds <- Functional_rarity %>%
  filter(!is.na(global_di), !is.na(Ri_plot)) %>%
  distinct(Treatment, Species, .keep_all = TRUE) %>%
  group_by(Treatment) %>%
  summarise(
    Di_q25 = quantile(global_di, 0.25),
    Di_q75 = quantile(global_di, 0.75),
    Ri_q25 = quantile(Ri_plot, 0.25),
    Ri_q75 = quantile(Ri_plot, 0.75),
    .groups = "drop"
  ) %>%
  # translate to the plotted x scale: high Ri -> LOW Restrict2
  mutate(x_rare   = log10(1 - Ri_q75),   # left line
         x_common = log10(1 - Ri_q25))   # right line

# 2. Classification
Functional_rarity <- Functional_rarity %>%
  left_join(thresholds, by = "Treatment") %>%
  mutate(eco_class = case_when(
    global_di >= Di_q75 & Ri_plot >= Ri_q75 ~ "Ecologically rare",
    global_di <= Di_q25 & Ri_plot <= Ri_q25 ~ "Ecologically common",
    global_di <  Di_q75 & global_di > Di_q25 &
      Ri_plot < Ri_q75 & Ri_plot > Ri_q25   ~ "Ecologically average",
    TRUE                                    ~ "Intermediate"
  ))

#Plot
ggplot(Functional_rarity, aes(x = Restrict2, y = global_di)) +
    scale_colour_manual(values= c("#f452ad", "grey", "#adf452"))+
  # shaded corners (fill mapped to a label -> appears in legend)
  geom_rect(data = thresholds, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = x_rare, ymin = Di_q75, ymax = Inf,
                fill = "Ecologically rare"),
            alpha = 0.15) +
  # geom_rect(data = thresholds, inherit.aes = FALSE,
  #           aes(xmin = x_common, xmax = Inf, ymin = -Inf, ymax = Di_q25,
  #               fill = "Ecologically common"),
  #           alpha = 0.15) +
  scale_fill_manual(
    name   = NULL,
    values = c("Ecologically rare" = "firebrick",
               "Ecologically common" = "steelblue"),
    breaks = c("Ecologically rare", "Ecologically common")
  ) +
  # threshold lines
  geom_vline(data = thresholds, aes(xintercept = x_rare),   linetype = "dashed") +
  geom_vline(data = thresholds, aes(xintercept = x_common), linetype = "dotted") +
  geom_hline(data = thresholds, aes(yintercept = Di_q75),   linetype = "dashed") +
  geom_hline(data = thresholds, aes(yintercept = Di_q25),   linetype = "dotted") +
  geom_point(aes(colour = statut)) +
  facet_wrap(vars(Treatment)) +
  coord_cartesian(ylim = c(0.1, 0.65)) +
  labs(x = "log10(1 - Ri)  (← more restricted)",
       y = "Functional distinctiveness",
       colour = "Statut") +
  guides(fill = guide_legend(override.aes = list(alpha = 0.3))) +
  theme_minimal()
 


#"~5% rare" claim of Loiseau
Functional_rarity %>%
  distinct(Treatment, Species, eco_class) %>%
  count(Treatment, eco_class) %>%
  group_by(Treatment) %>%
  mutate(prop = round(100 * n / sum(n), 1))
```



##PCA per treatment 
```{r}

###warm -----------
# Espèces présentes sous Warm avec leurs coordonnées dans l'ACP GLOBALE
species_warm <- data_plot %>%
  mutate(Species = gsub("_", " ", Species)) 

species_warm$Species <- str_replace(species_warm$Species, "[.]", "-") 

species_warm <- species_warm %>%
  filter(Treatment =="warm")%>%
  pull(Species) %>%
  unique() 

df_W<- df_W %>% #see df_W calculatedd au début
  mutate(Species = gsub("_", " ", Species)) 

df_W$Species <- str_replace(df_W$Species, "[.]", "-") 

rarity_df_warm <- rarity_df%>%
    filter(Species %in% species_warm) %>%
  # Joindre emmean et statut de Warm
  left_join(
    df_W %>% dplyr::select(Species, emmean, statut),
    by = "Species"
  ) 

scale_factor_global <- 5

lim_emmean <- quantile(rarity_df_warm$emmean,
                       c(0.02, 0.98), na.rm = TRUE)

# PCA_warm <- ggplot(
#   rarity_df_warm,
#   aes(x = Dim1, y = Dim2)
# ) +
#   # Points colorés par emmean
#   # Shape = rareté fonctionnelle
#   geom_point(
#     aes(colour = emmean,
#         shape  = Rarity,
#         size   = Rarity),
#     alpha = 0.8, size=3) +
#   # Flèches des traits
#   geom_segment(data = pca_global$eig,
#                aes(x = 0, y = 0, xend = Dim1 * scale_factor, yend = Dim2 * scale_factor),
#                arrow = arrow(length = unit(0.25, "cm")), color = "black") +
#   geom_text(data = loading_df,
#             aes(x = Dim1 * scale_factor * 1.1, y = Dim2 * scale_factor * 1.1, label = Trait),
#             color = "black", size = 5) +
#   scale_Wlour_gradient2(
#     low      = "red",
#     mid      = "grey70",
#     high     = "green",
#     midpoint = 0,
# #    limits   = lim_emmean,
#     oob      = scales::squish,
#     name     = "r",
#     guide    = guide_Wlorbar(barwidth      = 4,
#                               barheight     = 1,
#                               title.position = "top", 
#                               direction="horizontal")
#   ) +
#   scale_shape_manual(
#     values = c(
#       "Frare"   = 8,  
#       "average" = 20,  
#       "Fcommon" = 16  
#     ),
#     labels = c(
#       "Frare"   = "Functionally rare",
#       "average" = "Average",
#       "Fcommon" = "Functionally common"
#     ),
#     name = "Distinctiveness (Global)",
#       guide = guide_legend(direction="vertical")
#   ) +
#   scale_size_manual(
#     values = c(
#       "Frare"   = 2,
#       "average" = 1,
#       "Fcommon" = 2
#     ),
#     guide = "none"
#   ) +
#   labs(
#     x        = "Dim1",
#     y        = "Dim2",
#     title    = "Warm (n=503)" ) +
#   theme_minimal(base_size = 11) +
#   theme(
#     legend.position  = "right",
#     panel.grid.minor = element_blank())
# 
# PCA_warm


#test r ~PC1 and PC2
mod_warm <- lm(emmean ~ Dim1, data=rarity_df_warm %>% filter(Rarity=="Frare"))
summary(mod_warm)

###------------------
#control_origin----------------------------

species_W <- data_plot %>%
  mutate(Species = gsub("_", " ", Species))

species_W$Species <- str_replace(species_W$Species, "[.]", "-")

species_W <- species_W %>%
  filter(Treatment =="Control_origin")%>%
  pull(Species) %>%
  unique()

df_W<- df_W %>% #see df_W calculatedd au début
  mutate(Species = gsub("_", " ", Species))

df_W$Species <- str_replace(df_W$Species, "[.]", "-")

rarity_df_W <- rarity_df%>%
    filter(Species %in% species_W) %>%
  # Joindre emmean et statut de Warm
  left_join(
    df_W %>% dplyr::select(Species, emmean, statut),
    by = "Species"
  ) 

# PCA_W <- ggplot(
#   rarity_df_W,
#   aes(x = Dim1, y = Dim2)
# ) +
#   # Points colorés par emmean
#   # Shape = rareté fonctionnelle
#   geom_point(
#     aes(colour = emmean,
#         shape  = Rarity,
#         size   = Rarity),
#     alpha = 0.8, size=3) +
#   # Flèches des traits
#   geom_segment(data = loading_df,
#                aes(x = 0, y = 0, xend = Dim1 * scale_factor, yend = Dim2 * scale_factor),
#                arrow = arrow(length = unit(0.25, "cm")), color = "black") +
#   geom_text(data = loading_df,
#             aes(x = Dim1 * scale_factor * 1.1, y = Dim2 * scale_factor * 1.1, label = Trait),
#             color = "black", size = 5) +
#   scale_Wlour_gradient2(
#     low      = "red",
#     mid      = "grey70",
#     high     = "green",
#     midpoint = 0,
# #    limits   = lim_emmean,
#     oob      = scales::squish,
#     name     = "r",
#     guide    = guide_Wlorbar(barwidth      = 4,
#                               barheight     = 1,
#                               title.position = "top", 
#                               direction="horizontal")
#   ) +
#   scale_shape_manual(
#     values = c(
#       "Frare"   = 8,  
#       "average" = 20,  
#       "Fcommon" = 16  
#     ),
#     labels = c(
#       "Frare"   = "Functionally rare",
#       "average" = "Average",
#       "Fcommon" = "Functionally common"
#     ),
#     name = "Distinctiveness (Global)",
#       guide = guide_legend(direction="vertical")
#   ) +
#   labs(
#     x        = "Dim1",
#     y        = "Dim2",
#     title    = "CO (n=465)" ) +
#   theme_minimal(base_size = 11) +
#   theme(
#     legend.position  = "right",
#     panel.grid.minor = element_blank())
# 
# PCA_W 

#test r ~PC1 and PC2
mod_W <- lm(emmean ~ Dim1, data= rarity_df_W%>% filter(Rarity=="Frare"))
summary(mod_W)

#control_dest------------------------------

species_CD <- data_plot %>%
  mutate(Species = gsub("_", " ", Species))

species_CD$Species <- str_replace(species_CD$Species, "[.]", "-")

species_CD <- species_CD %>%
  filter(Treatment =="Control_dest")%>%
  pull(Species) %>%
  unique()

df_CD<- df_CD %>% #see df_W calculatedd au début
  mutate(Species = gsub("_", " ", Species))

df_CD$Species <- str_replace(df_CD$Species, "[.]", "-")

rarity_df_CD <- rarity_df%>%
    filter(Species %in% species_CD) %>%
  # Joindre emmean et statut de Warm
  left_join(
    df_CD %>% dplyr::select(Species, emmean, statut),
    by = "Species"
  ) 

# PCA_CD <- ggplot(
#   rarity_df_CD,
#   aes(x = Dim1, y = Dim2)
# ) +
#   # Points colorés par emmean
#   # Shape = rareté fonctionnelle
#   geom_point(
#     aes(colour = emmean,
#         shape  = Rarity,
#         size   = Rarity),
#     alpha = 0.8, size=3) +
#   # Flèches des traits
#   geom_segment(data = loading_df,
#                aes(x = 0, y = 0, xend = Dim1 * scale_factor, yend = Dim2 * scale_factor),
#                arrow = arrow(length = unit(0.25, "cm")), color = "black") +
#   geom_text(data = loading_df,
#             aes(x = Dim1 * scale_factor * 1.1, y = Dim2 * scale_factor * 1.1, label = Trait),
#             color = "black", size = 5) +
#   scale_Wlour_gradient2(
#     low      = "red",
#     mid      = "grey70",
#     high     = "green",
#     midpoint = 0,
# #    limits   = lim_emmean,
#     oob      = scales::squish,
#     name     = "r",
#     guide    = guide_Wlorbar(barwidth      = 4,
#                               barheight     = 1,
#                               title.position = "top", 
#                               direction="horizontal")
#   ) +
#   scale_shape_manual(
#     values = c(
#       "Frare"   = 8,  
#       "average" = 20,  
#       "Fcommon" = 16  
#     ),
#     labels = c(
#       "Frare"   = "Functionally rare",
#       "average" = "Average",
#       "Fcommon" = "Functionally common"
#     ),
#     name = "Distinctiveness (Global)",
#       guide = guide_legend(direction="vertical")
#   ) +
#   labs(
#     x        = "Dim1",
#     y        = "Dim2",
#     title    = "CD (n=358)" ) +
#   theme_minimal(base_size = 11) +
#   theme(
#     legend.position  = "right",
#     panel.grid.minor = element_blank())
# 
#PCA_W 

#test r ~PC1 and PC2
mod_CD <- lm(emmean ~ Dim1, data= rarity_df_CD%>% filter(Rarity=="Frare"))
summary(mod_CD)

# pca_treatment <- ggarrange(PCA_W, PCA_CD, PCA_warm, common.legend = T, legend="bottom", nrow=1)
# pca_treatment
# 
# ggsave("Figures/PCA/pca_treatment_statu_distinctiveness.jpeg", pca_treatment, height= 6, width=12)
```

## Chi2 proprotions
```{r}
#does functional distinctiveness predict being a loser or a winner? 
#data combine
pca_all_treatments <- bind_rows(
  rarity_df_warm %>% mutate(Treatment = "warm"),
  rarity_df_W   %>% mutate(Treatment = "Control_origin"),
  rarity_df_CD   %>% mutate(Treatment = "Control_dest")
) %>%
  filter(!is.na(statut), !is.na(global_di)) %>%
  mutate(
    statut    = factor(statut,
                       levels = c("winner", "stable", "loser")),
    Treatment = factor(Treatment,
                       levels = c("Control_origin",
                                  "Control_dest",
                                  "warm"))
  )

tab2 <- table(pca_all_treatments$Rarity, pca_all_treatments$statut)

tab3<-pca_all_treatments %>%
  count(Treatment, Rarity, statut) %>%
  group_by(Treatment, Rarity)


# chi2 tests : a small p-value means the variables are not independent (they are associated).
#All expected counts here are well above 5
#https://www.datanovia.com/learn/biostatistics/categorical/chi-square-test-of-independence-in-r

chisq_warm <- chisq.test(xtabs(~ Rarity + statut, data = pca_all_treatments %>%
    filter(Treatment == "warm")))

chisq_W <- chisq.test(xtabs(~ Rarity + statut, data = pca_all_treatments %>%
    filter(Treatment == "Control_origin")))

chisq_CD <- chisq.test(xtabs(~ Rarity + statut, data = pca_all_treatments %>%
    filter(Treatment == "Control_dest")))

#Les résidus de Pearson indiquent l'écart à l'indépendance: Large positive (blue) = more than expected (attraction); large negative (red) = fewer than expected (repulsion):
par(mfrow=c(1,3))
corrplot(chisq_warm$residuals, is.cor = FALSE, main="warm", mar=c(5,4,4,2)) 
corrplot(chisq_W$residuals, is.cor = FALSE, main="Control_origin", mar=c(5,4,4,2))
corrplot(chisq_CD$residuals, is.cor = FALSE, main="Control_dest", mar=c(5,4,4,2))

# Cramer's V: (effect size) V proche de 0 = association faible
cramer_v <- function(chisq_obj) {
     n   <- sum(chisq_obj$observed)
     k   <- min(dim(chisq_obj$observed))
     sqrt(chisq_obj$statistic / (n * (k - 1)))
 }
 
cramer_v(chisq_warm)
cramer_v(chisq_W)
cramer_v(chisq_CD)

####
# Résidus Pearson pour comprendre le sens
round(chisq_warm$residuals, 2)
round(chisq_W$residuals, 2)

# Observé vs attendu sous Warm
chisq_warm$observed
round(chisq_warm$expected, 1)
#####################

# fig % statuts par Rarity_global × traitement ───────
fig_chi2 <- pca_all_treatments %>%
  count(Treatment, Rarity, statut) %>%
  group_by(Treatment, Rarity) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(
    Rarity_global = factor(Rarity,
                           levels = c("Frare", "average", "Fcommon"),
                           labels = c("Functionally\nrare",
                                      "Average",
                                      "Functionally\ncommon")),
    statut = factor(statut,
                    levels = c("winner", "stable", "loser"))
  ) %>%
  ggplot(aes(x = Rarity, y = pct, fill = statut)) +
  geom_Wl(position = "stack") +
  geom_text(aes(label = ifelse(pct > 5,
                               paste0(round(pct, 0), "%"), "")),
            position = position_stack(vjust = 0.5),
            colour = "white", fontface = "bold", size = 3) +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  facet_wrap(~ Treatment, ncol = 3,
             labeller = labeller(Treatment = c(
               "Control_origin" = "Control origin",
               "Control_dest"   = "Control dest",
               "warm"           = "Warm"
             ))) +
  labs(x     = " ",
       y     = "% species",
       fill  = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

fig_chi2


```


##Linear models
```{r}
#function to test the linear models of emmean en fonction des differentes dimensions et du status de rareté

mod_lin_emmean <- function(data, dim, rarity_filter=NULL){
  if (!is.null(rarity_filter)) {
    data <- data %>% filter(Rarity == rarity_filter)
  }
  formula <- as.formula(paste("emmean ~", dim))
  mod <- lm(formula, data=data)
}

a<- mod_lin_emmean(rarity_df_W, dim = "Dim1", rarity_filter = "Frare")
summary(a)

#test de comparaison avec 
mod_W_all_1 <- lm(emmean ~ Dim1, data=rarity_df_W)
summary(mod_W_all_1)
mod_W_rare_1 <- lm(emmean ~ Dim1, data=rarity_df_W %>%filter(Rarity=="Frare"))
summary(mod_W_rare_1)

#ok it works



datasets <- list(CO = rarity_df_W, warm = rarity_df_warm, CD = rarity_df_CD)
dims <- c("Dim1", "Dim2","Dim3", "Dim4", "Dim5")
rarities <- c(NULL, "Frare", "Fcommon", "average")  # "all" = pas de filtre

grid <- expand.grid(
  dataset_name = names(datasets),
  dim_var = dims,
  rarity = rarities,
  stringsAsFactors = FALSE
)

nrow(grid)  

#boucle
results <- list()

for (i in 1:nrow(grid)){
  row <- grid[i,]
  data_i <- datasets[[row$dataset_name]]
  
  mod <- mod_lin_emmean(data = data_i, dim = row$dim_var, rarity_filter = row$rarity)
  
  # nom du modèle, ex: "mod_W_Frare_Dim1"
  mod_name <- paste0("mod_", row$dataset_name, "_", row$rarity, "_", row$dim_var)
  
  results[[mod_name]] <- mod
}


str(summary(results$mod_W_Frare_Dim1))
#R²
summary(mod_W_rare_1)$adj.r.squared
#pente
summary(mod_W_rare_1)$coefficient[2,1]
#df
summary(mod_W_rare_1)$fstatistic[[2]] #F(X, - )
summary(mod_W_rare_1)$fstatistic[[3]] #F(-,X)
summary(mod_W_rare_1)$fstatistic[[1]] #F
#pvalue
summary(mod_W_rare_1)$coefficient[2,4]

extract_lm_stats <- function(mod, dataset_name, rarity, dim_var) {
  s <- summary(mod)
  data.frame(
    mod_name  = paste0("mod_", dataset_name, "_", rarity, "_", dim_var),
    Treatment = dataset_name,
    Rarity    = rarity,
    Dim       = dim_var,
    R_adj     = s$adj.r.squared,
    df1       = unname(s$fstatistic[[2]]),
    df2       = unname(s$fstatistic[[3]]),
    F_stat    = unname(s$fstatistic[[1]]),
    pvalue    = s$coefficients[2, 4],
    slope     = s$coefficients[2, 1],
    stringsAsFactors = FALSE
  )
}

results <- list()
summary_list <- list()

for (i in 1:nrow(grid)) {
  row <- grid[i, ]
  data_i <- datasets[[row$dataset_name]]
  
  mod <- mod_lin_emmean(data = data_i, dim = row$dim_var, rarity_filter = row$rarity)
  
  mod_name <- paste0("mod_", row$dataset_name, "_", row$rarity, "_", row$dim_var)
  results[[mod_name]] <- mod
  
  summary_list[[i]] <- extract_lm_stats(mod, row$dataset_name, row$rarity, row$dim_var)
}

summary_df <- do.call(rbind, summary_list)

summary_df_filtered <- summary_df %>% filter(pvalue < 0.05)


###plots
summary_df <- summary_df %>%mutate(Rarity = recode(Rarity, "Frare" = "Functionally\n rare", "Fcommon" = "Functionally\n common", "average" = "Average"))

summary_df <- summary_df %>%mutate(Treatment = recode(Treatment, "CO" = "Control\n origin", "CD" = "Control\n destination", "warm" = "Warm"))

ggplot(summary_df) +
  aes(x = R_adj, y = Rarity, color = Dim) +
  geom_point(size =5) +
  geom_segment(aes(x = 0, xend=R_adj),
 size = 1)+
    scale_Wlor_manual(values=c('#eb7267','#67eb72','#ceeb67', '#67bdeb', 'grey'))+
  theme_minimal(base_size = 12) +
  facet_wrap2(vars(Treatment),
              strip = strip_themed(
                text_x = elem_list_text(color = c('orange', 'dodgerblue', 'firebrick1'))
              )) +
  labs(x="adjusted R²", 
       y=NULL,
       color ="PCA dimensions")+
  theme(axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        legend.title = element_text(size =12),
        strip.text = element_text(size=20))

ggplot(summary_df) +
  aes(x = R_adj, y = Treatment, color = Dim) +
  geom_point(size =5) +
  geom_segment(aes(x = 0, xend=R_adj),
 size = 1)+
  scale_Wlor_manual(values=c('#eb7267','#67eb72','#ceeb67', '#67bdeb', 'grey'))+
  theme_minimal(base_size = 12) +
  facet_wrap2(vars(Rarity),
              strip = strip_themed(
                text_x = elem_list_text(color = c('black', 'black', 'violet'))
              )) +
  labs(x="adjusted R²", 
       y=NULL,
       color ="PCA dimensions")+
  theme(axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        legend.title = element_text(size =12),
        strip.text = element_text(size=20))
  

 
```

j'ai besoin de R², pente, F, p value

*The overall regression was statistically significant (R2 = [R2 value], F(df regression, df residual) = [F-value], p = [p-value]).
  --> *The overall regression was statistically significant (R2 = 0.73, F(2, 17) = 23.46, p = < .000).*
  
It was found that [predictor variable] significantly predicted [response variable] (β = [β-value], p = [p-value]).
 --> *It was found that hours studied significantly predicted exam score (β = 5.56, p = < .000).*


#functionally rare species in CO and warm 
```{r}

##functionally rare species in CO and warm 
Frare_W <- rarity_df_W %>%
  filter(Rarity=="Frare")%>%
  pull(Species)

Frare_CD <- rarity_df_CD %>%
  filter(Rarity=="Frare")%>%
  pull(Species)

Species_W <- rarity_df_warm %>%
  filter(Rarity=="Frare")%>%
  pull(Species)

length(Frare_W) #119
length(Frare_CD) #76
length(Species_W) #124

Frare_W_W <- intersect(Frare_W, Species_W) #show whether rare species in CO stay rare in warm and whether they lose or win
Frare_CD_W <- intersect(Frare_CD, Species_W) #show whether rare species in CD stay rare in warm and whether they lose or win
  

common_W_W_rare <- pca_all_treatments%>%
  filter(Treatment =="warm")%>%
  filter(Species %in% Species_W)

mod4<- lm(emmean ~ global_di, data= common_W_W_rare)
summary(mod4)

mod6<- lm(emmean ~ Dim1, data=common_W_W_rare)
summary(mod6)

plot(mod6)

#modele 
#mod_5PCA<- lm(emmean ~ Dim1, data=common_W_W_rare)
summary(mod_5PCA)

#mod_4PCA<- lm(emmean ~ Dim1, data=common_W_W_rare)
summary(mod_4PCA)

#mod_3PCA<- lm(emmean ~ Dim1, data=common_W_W_rare)
summary(mod_3PCA)

# mod_2PCA<- lm(emmean ~ Dim1, data=common_W_W_rare)
summary(mod_2PCA)

```


#boxplots
##essai
```{r}

long_data_all %>%
  filter(Treatment == "Control_dest") %>%   # change : un treatment à la fois 
  ggplot(aes(x = value, fill = statut)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("firebrick1", "grey80", "#adf452")) +
  facet_grid(Rarity ~ Dim, scales = "free_y") +
  theme_minimal(base_size = 12) +
  labs(title = "(T) - distribution par statut")

# Quantitatif : skewness et variance par groupe, pour repérer les formes très différentes
a<- long_data_all %>%
  group_by(Dim, Treatment, Rarity, statut) %>%
  summarise(n = n(),
            variance = var(value, na.rm = TRUE),
            skewness = skewness(value, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(abs(skewness)))



#The standard effect size for a Kruskal-Wallis test is eta-squared (\(\eta _{H}^{2}\)) based on the H-statistic, which measures the proportion of variance explained in the ranks of the dependent variable: https://pmc.ncbi.nlm.nih.gov/articles/PMC12701665/ --> convey the magnitude of an observed effect and are highly sensitive to sample size

long_data_all <- pca_all_treatments %>%
  pivot_longer(cols = c("Dim1", "Dim2", "Dim3", "Dim4", "Dim5"),
               names_to = "Dim", values_to = "value")

long_data_all$statut <- factor(long_data_all$statut, levels = c('loser', 'stable', 'winner'))

kw_summary <- long_data_all %>%
  group_by(Dim, Treatment, Rarity) %>%
  kruskal_effsize(value ~ statut) %>%
  ungroup()

kw_pvals <- long_data_all %>%
  group_by(Dim, Treatment, Rarity) %>%
  kruskal_test(value ~ statut) %>%
  ungroup() %>%
  dplyr::select(Dim, Treatment, Rarity, p)

dunn_wl <- long_data_all %>%
  group_by(Dim, Treatment, Rarity) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup() %>%
  filter(group1 == "loser", group2 == "winner") %>% 
  mutate(direction = case_when(
    p.adj >= 0.05 ~ "ns",
    statistic > 0 ~ "winner",
    statistic < 0 ~ "loser"
  )) %>%
  dplyr::select(Dim, Treatment, Rarity, statistic, direction, dunn_p = p.adj)

plot_data <- kw_summary %>%
  left_join(kw_pvals, by = c("Dim", "Treatment", "Rarity")) %>%
  left_join(dunn_wl, by = c("Dim", "Treatment", "Rarity"))

##Add all species data
long_data_all <- pca_all_treatments %>%
  pivot_longer(cols = c("Dim1", "Dim2", "Dim3", "Dim4", "Dim5"),
               names_to = "Dim", values_to = "value")

long_data_all$statut <- factor(long_data_all$statut, levels = c('loser', 'stable', 'winner'))

kw_summary2 <- long_data_all %>%
  group_by(Dim, Treatment) %>%
  kruskal_effsize(value ~ statut) %>%
  ungroup()

kw_pvals2 <- long_data_all %>%
  group_by(Dim, Treatment) %>%
  kruskal_test(value ~ statut) %>%
  ungroup() %>%
  dplyr::select(Dim, Treatment, p)

dunn_wl2 <- long_data_all %>%
  group_by(Dim, Treatment) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup() %>%
  filter(group1 == "loser", group2 == "winner") %>% 
  mutate(direction = case_when(
    p.adj >= 0.05 ~ "ns",
    statistic > 0 ~ "winner",
    statistic < 0 ~ "loser"
  )) %>%
  dplyr::select(Dim, Treatment,statistic, direction, dunn_p = p.adj)

plot_data2 <- kw_summary2 %>%
  left_join(kw_pvals2, by = c("Dim", "Treatment")) %>%
  left_join(dunn_wl2, by = c("Dim", "Treatment"))%>%
  mutate(Rarity ="All species")

#join both
plot_data3 <- rbind(plot_data, plot_data2) %>%
  mutate(effsize_sign = case_when(
    statistic > 0 ~ effsize*1,
    statistic < 0 ~ effsize*(-1)))

#reorder & recode
plot_data3 <- plot_data3 %>%
  mutate(Rarity  = factor(Rarity,
                       levels = c("average", "Fcommon", "Frare", "All species")), 
       Treatment =  dplyr::recode(Treatment,
     "Control_dest" ="Control destination",
    "Control_origin" = "Control origin",
    "warm" = "Warm"
  ))

#plot
(plot_kw <- ggplot(plot_data3) +
  aes(x = effsize_sign, y = Rarity, color = direction) +
  geom_segment(aes(x = 0, xend = effsize_sign), size = 1) +
  geom_point(size = 5) +
  scale_Wlor_manual(values = c("winner" = '#67eb72', 
                                 "loser" = "firebrick1",
                                 "ns" = "grey70")) +
  theme_minimal(base_size = 12) +
  facet_grid(Dim ~ Treatment) +
  labs(x = "Kruskal-Wallis effect size",
       y = NULL,
       color = NULL) +
  theme(strip.text = element_text(size = 12),
        legend.position = "bottom"))


#for powerpoint only ALL species
x_range <- range(plot_data3$effsize_sign, na.rm = TRUE)

(slide1 <- ggplot(plot_data3) +
  aes(x = effsize_sign, y = Rarity) +
  geom_segment(
    data = plot_data3 %>% filter(Rarity != "All species"),
    aes(x = 0, xend = effsize_sign),
    colour = "grey90", size = 1, alpha =0
  ) +
  geom_point(
    data = plot_data3 %>% filter(Rarity != "All species"),
    colour = "grey90", size = 5, alpha=0
  ) +
  geom_segment(
    data = plot_data3 %>% filter(Rarity == "All species"),
    aes(x = 0, xend = effsize_sign, colour = direction),
    size = 1
  ) +
  geom_point(
    data = plot_data3 %>% filter(Rarity == "All species"),
    aes(colour = direction),
    size = 5
  ) +
  scale_Wlour_manual(values = c(
    "winner" = "#67eb72",
    "loser"  = "firebrick1",
    "ns"     = "grey70"
  )) +
  scale_x_Wntinuous(limits = x_range) +
  facet_grid(Dim ~ Treatment) +
  labs(x = "Kruskal-Wallis effect size",
       y = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(strip.text      = element_text(size = 12),
        legend.position = "bottom"))

(slide2 <- ggplot(plot_data3) +
  aes(x = effsize_sign, y = Rarity) +
  geom_segment(
    data = plot_data3 %>% filter(Treatment !="Warm"),
    aes(x = 0, xend = effsize_sign),
    colour = "grey90", size = 1, alpha =0
  ) +
  geom_point(
    data = plot_data3 %>% filter(Treatment !="Warm"),
    colour = "grey90", size = 5, alpha=0
  ) +
  geom_segment(
    data = plot_data3 %>% filter(Treatment !="Warm"),
    aes(x = 0, xend = effsize_sign, colour = direction),
    size = 1
  ) +
  geom_point(
    data = plot_data3 %>% filter(Rarity == "All species"),
    aes(colour = direction),
    size = 5
  ) +
  scale_Wlour_manual(values = c(
    "winner" = "#67eb72",
    "loser"  = "firebrick1",
    "ns"     = "grey70"
  )) +
  scale_x_Wntinuous(limits = x_range) +
  facet_grid(Dim ~ Treatment) +
  labs(x = "Kruskal-Wallis effect size",
       y = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(strip.text      = element_text(size = 12),
        legend.position = "bottom"))


```


##all species- WARM
```{r}
pca_all_treatments 
dim(pca_all_treatments)
names(pca_all_treatments)

(tab <- table(pca_all_treatments$Rarity))

(plot_all <- ggplot(data = pca_all_treatments %>% filter(Treatment =="warm"), 
       aes(x = statut, y=Dim1, fill= statut))+
  geom_boxplot()+
  scale_fill_manual(values=c('#adf452','grey','firebrick1'))+
  theme_minimal()+
    facet_wrap(~Treatment))

long_data <- pivot_longer(pca_all_treatments %>% filter(Treatment=="warm"), cols=c("Dim1", "Dim2"), # "Dim3", "Dim4", "Dim5"), 
                          names_to="Dim", values_to="value")


long_data$statut <- factor(long_data$statut, levels = c('loser','stable', 'winner'))

dunn_res_all <- long_data %>%
  group_by(Dim) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup()

dunn_res_all <- dunn_res_all %>%
  group_by(Dim) %>%
  add_xy_position(x = "statut") %>%
  ungroup()

(plot <- ggplot(data=long_data, 
                aes(x=statut, y=value, fill= statut))+
    geom_boxplot()+
    scale_fill_manual(values= c("firebrick1","grey80", "#adf452"))+
    theme_minimal(base_size = 20)+
     labs(title="Warm - all species")+
    facet_wrap(~Dim)+
  ylim(-6,8))

plot <- plot + 
  stat_pvalue_manual(dunn_res_all, label = "p.adj.signif", size = 5, p.digits = 3, tip.length = 0.1, hide.ns = T, color="black", bracket.size = 1, bracket.nudge.y = c(0.5, 2))


ns_dims <- dunn_res_all %>%
  group_by(Dim) %>%
  summarise(all_ns = all(p.adj.signif == "ns")) %>%
  filter(all_ns) %>%
  pull(Dim)

ns_labels <- data.frame(
  Dim = ns_dims,
  label = "ns"
)

plot +
  geom_text(data = ns_labels, 
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.2, vjust = 1.5, 
            size = 5, color = "grey40",
            inherit.aes = FALSE)

```

##all species - CO
```{r}

(plot_all <- ggplot(data = pca_all_treatments %>% filter(Treatment =="Control_origin"), 
       aes(x = statut, y=Dim1, fill= statut))+
  geom_boxplot()+
  scale_fill_manual(values=c('#adf452','grey','firebrick1'))+
  theme_minimal()+
    facet_wrap(~Treatment))

long_data <- pivot_longer(pca_all_treatments %>% filter(Treatment=="Control_origin"), cols=c("Dim1", "Dim2"), # "Dim3", "Dim4", "Dim5"), 
                          names_to="Dim", values_to="value")


long_data$statut <- factor(long_data$statut, levels = c('loser','stable', 'winner'))

dunn_res_all <- long_data %>%
  group_by(Dim) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup()

dunn_res_all <- dunn_res_all %>%
  group_by(Dim) %>%
  add_xy_position(x = "statut") %>%
  ungroup()

(plot <- ggplot(data=long_data, 
                aes(x=statut, y=value, fill= statut))+
    geom_boxplot()+
    scale_fill_manual(values= c("firebrick1","grey80", "#adf452"))+
    theme_minimal(base_size = 20)+
     labs(title="Control origin - all species")+
    facet_wrap(~Dim)+
  ylim(-6,8))

plot <- plot + 
  stat_pvalue_manual(dunn_res_all, label = "p.adj.signif", size = 5, p.digits = 3, tip.length = 0.1, hide.ns = T, color="black", bracket.size = 1)


ns_dims <- dunn_res_all %>%
  group_by(Dim) %>%
  summarise(all_ns = all(p.adj.signif == "ns")) %>%
  filter(all_ns) %>%
  pull(Dim)

ns_labels <- data.frame(
  Dim = ns_dims,
  label = "ns"
)

plot +
  geom_text(data = ns_labels, 
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.2, vjust = 1.5, 
            size = 5, color = "grey40",
            inherit.aes = FALSE)



```

##all species - CD
```{r}
(plot_all <- ggplot(data = pca_all_treatments %>% filter(Treatment =="Control_dest"), 
       aes(x = statut, y=Dim1, fill= statut))+
  geom_boxplot()+
  scale_fill_manual(values=c('#adf452','grey','firebrick1'))+
  theme_minimal()+
    facet_wrap(~Treatment))

long_data <- pivot_longer(pca_all_treatments %>% filter(Treatment=="Control_dest"), cols=c("Dim1", "Dim2"), # "Dim3", "Dim4", "Dim5"), 
                          names_to="Dim", values_to="value")


long_data$statut <- factor(long_data$statut, levels = c('loser','stable', 'winner'))

dunn_res_all <- long_data %>%
  group_by(Dim) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup()

dunn_res_all <- dunn_res_all %>%
  group_by(Dim) %>%
  add_xy_position(x = "statut") %>%
  ungroup()

(plot <- ggplot(data=long_data, 
                aes(x=statut, y=value, fill= statut))+
    geom_boxplot()+
    scale_fill_manual(values= c("firebrick1","grey80", "#adf452"))+
    theme_minimal(base_size = 20)+
     labs(title="Control dest - all species")+
    facet_wrap(~Dim)+
  ylim(-6,8))

plot <- plot + 
  stat_pvalue_manual(dunn_res_all, label = "p.adj.signif", size = 5, p.digits = 3, tip.length = 0.1, hide.ns = T, color="black", bracket.size = 1)


ns_dims <- dunn_res_all %>%
  group_by(Dim) %>%
  summarise(all_ns = all(p.adj.signif == "ns")) %>%
  filter(all_ns) %>%
  pull(Dim)

ns_labels <- data.frame(
  Dim = ns_dims,
  label = "ns"
)

plot +
  geom_text(data = ns_labels, 
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.2, vjust = 1.5, 
            size = 5, color = "grey40",
            inherit.aes = FALSE)
```



##functionally rare species - WARM
```{r}
#let's process per rarity category
#first functionally rare species
FR_species <- pca_all_treatments%>% filter(Rarity =="Frare") 
table(FR_species$statut, FR_species$Treatment)

#functionally common 
FC_species <- pca_all_treatments%>% filter(Rarity =="Fcommon")
table(FC_species$statut, FC_species$Treatment)

#average
AV_species <- pca_all_treatments%>% filter(Rarity =="average")
table(AV_species$statut, AV_species$Treatment)

#FR in warm only 
FR_species_warm <- pca_all_treatments%>% filter(Rarity =="Frare") %>%filter(Treatment =="warm") 
table(FR_species_warm$statut)

kruskal_test(data = FR_species_warm, Dim1 ~statut)
kruskal_test(data = FR_species_warm, Dim2 ~statut)
kruskal_test(data = FR_species_warm, Dim3 ~statut)
kruskal_test(data = FR_species_warm, Dim4 ~statut)
kruskal_test(data = FR_species_warm, Dim5 ~statut)

# pvals <- c(Dim1=5.32e-6, Dim2=0.381, Dim3=0.0628, Dim4=0.981, Dim5=0.699)
# p.adjust(pvals, method="fdr")

dunn_res1 <- dunn_test(data=FR_species_warm, Dim1~statut, p.adjust.method = "BH") # Benjamini-Hochberg procedure, also known as the False Discovery Rate (FDR)
dunn_res1 <- dunn_res1 %>% add_xy_position(x="statut")

#add for the others 
dunn_res2 <- dunn_test(data=FR_species_warm, Dim2~statut, p.adjust.method = "BH")
dunn_res2 <- dunn_res2 %>% add_xy_position(x="statut")
dunn_res3 <- dunn_test(data=FR_species_warm, Dim3~statut, p.adjust.method = "BH")
dunn_res3 <- dunn_res3 %>% add_xy_position(x="statut")
dunn_res4 <- dunn_test(data=FR_species_warm, Dim4~statut, p.adjust.method = "BH")
dunn_res4 <- dunn_res4 %>% add_xy_position(x="statut")
dunn_res5 <- dunn_test(data=FR_species_warm, Dim5~statut, p.adjust.method = "BH")
dunn_res5 <- dunn_res5 %>% add_xy_position(x="statut")

dunn_res_all <- rbind(dunn_res1, dunn_res2, dunn_res3, dunn_res4, dunn_res5)

(plot_dim1 <- ggplot(data = FR_species_warm, 
       aes(x = statut, y=Dim1, fill=statut))+
  geom_boxplot()+
   scale_fill_manual(values=c('#adf452','grey','firebrick1'))+
  theme_minimal()) #for all Dim? 

#format long
long_data <- pivot_longer(FR_species_warm, cols=c("Dim1", "Dim2"),
                          #"Dim3", "Dim4", "Dim5"), 
                          names_to="Dim", values_to="value")


long_data$statut <- factor(long_data$statut, levels = c('loser','stable', 'winner'))

dunn_res_all <- long_data %>%
  group_by(Dim) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup()

dunn_res_all <- dunn_res_all %>%
  group_by(Dim) %>%
  add_xy_position(x = "statut") %>%
  ungroup()

(plot <- ggplot(data=long_data, 
                aes(x=statut, y=value, fill= statut))+
    geom_boxplot()+
    scale_fill_manual(values= c("firebrick1","grey80", "#adf452"))+
    theme_minimal(base_size = 20)+
    theme(axis.text.x = element_blank())+
     labs(title="Warm - Functionally rare species")+
    facet_wrap(~Dim)+
  ylim(-6,8))

plot <- plot + 
  stat_pvalue_manual(dunn_res_all, label = "p.adj.signif", size = 5, p.digits = 3, tip.length = 0.1, hide.ns = T, color="black", bracket.size = 1, bracket.nudge.y = c(0.2, 1.5, 0.5))


ns_dims <- dunn_res_all %>%
  group_by(Dim) %>%
  summarise(all_ns = all(p.adj.signif == "ns")) %>%
  filter(all_ns) %>%
  pull(Dim)

ns_labels <- data.frame(
  Dim = ns_dims,
  label = "ns"
)

plot +
  geom_text(data = ns_labels, 
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.2, vjust = 1.5, 
            size = 5, color = "grey40",
            inherit.aes = FALSE)
```

##functionally rare species - CO
```{r}
pca_all_treatments 
dim(pca_all_treatments)
names(pca_all_treatments)

#let's process per rarity category
#first functionally rare species
FR_species <- pca_all_treatments%>% filter(Rarity =="Frare") 
table(FR_species$statut, FR_species$Treatment)

#FR in CO only 
FR_species_W <- pca_all_treatments%>% filter(Rarity =="Frare") %>%filter(Treatment =="Control_origin") 
table(FR_species_W$statut)

kruskal_test(data = FR_species_W, Dim1 ~statut)
kruskal_test(data = FR_species_W, Dim2 ~statut)
kruskal_test(data = FR_species_W, Dim3 ~statut)
kruskal_test(data = FR_species_W, Dim4 ~statut)
kruskal_test(data = FR_species_W, Dim5 ~statut)

pvals <- c(Dim1= 0.0464, Dim2= 0.683, Dim3=0.626 , Dim4= 0.0726, Dim5= 0.671)
p.adjust(pvals, method="BH")

#format long
long_data_W <- pivot_longer(FR_species_W, cols=c("Dim1", "Dim2"), 
                             #"Dim3", "Dim4", "Dim5"), 
                             names_to="Dim", values_to="value")

long_data_W$statut <- factor(long_data_W$statut, levels = c('loser','stable', 'winner'))

dunn_res_all_W <- long_data_W %>%
  group_by(Dim) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup()

dunn_res_all_W <- dunn_res_all_W %>%
  group_by(Dim) %>%
  add_xy_position(x = "statut") %>%
  ungroup()

(plot_W <- ggplot(data=long_data_W, 
            aes(x=statut, y=value, fill= statut))+
    geom_boxplot()+
     scale_fill_manual(values= c("firebrick1","grey80", "#adf452"))+
    theme_minimal(base_size = 20)+
    theme(axis.text.x = element_blank())+
    facet_wrap(~Dim)+
  labs(title="Control origin - Functionally rare species")+
  ylim(-6,8))

(plot_W <- plot_W + 
  stat_pvalue_manual(dunn_res_all_W, label = "p.adj.signif", size = 5, p.digits = 3, tip.length = 0.1, hide.ns = T, color="black", bracket.size = 1))


ns_dims <- dunn_res_all_W %>%
  group_by(Dim) %>%
  summarise(all_ns = all(p.adj.signif == "ns")) %>%
  filter(all_ns) %>%
  pull(Dim)

ns_labels <- data.frame(
  Dim = ns_dims,
  label = "ns"
)

plot_W +
  geom_text(data = ns_labels, 
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.2, vjust = 1.5, 
            size = 5, color = "grey40",
            inherit.aes = FALSE)
```
##functionally rare species - CD
```{r}
pca_all_treatments 
dim(pca_all_treatments)
names(pca_all_treatments)

#let's process per rarity category
#first functionally rare species
FR_species <- pca_all_treatments%>% filter(Rarity =="Frare") 
table(FR_species$statut, FR_species$Treatment)

#FR in CO only 
FR_species_CD <- pca_all_treatments%>% filter(Rarity =="Frare") %>%filter(Treatment =="Control_dest") 
table(FR_species_CD$statut)

kruskal_test(data = FR_species_CD, Dim1 ~statut)
kruskal_test(data = FR_species_CD, Dim2 ~statut)
kruskal_test(data = FR_species_CD, Dim3 ~statut)
kruskal_test(data = FR_species_CD, Dim4 ~statut)
kruskal_test(data = FR_species_CD, Dim5 ~statut)

pvals <- c(Dim1= 0.0464, Dim2= 0.683, Dim3=0.626 , Dim4= 0.0726, Dim5= 0.671)
p.adjust(pvals, method="BH")

#format long
long_data_CD <- pivot_longer(FR_species_CD, cols=c("Dim1", "Dim2"), # "Dim3", "Dim4", "Dim5"), 
                             names_to="Dim", values_to="value")

long_data_CD$statut <- factor(long_data_CD$statut, levels = c('loser','stable', 'winner'))

dunn_res_all_CD <- long_data_CD %>%
  group_by(Dim) %>%
  dunn_test(value ~ statut, p.adjust.method = "BH") %>%
  ungroup()

dunn_res_all_CD <- dunn_res_all_CD %>%
  group_by(Dim) %>%
  add_xy_position(x = "statut") %>%
  ungroup()

(plot_CD <- ggplot(data=long_data_CD, 
            aes(x=statut, y=value, fill= statut))+
    geom_boxplot()+
     scale_fill_manual(values= c("firebrick1","grey80", "#adf452"))+
    theme_minimal(base_size = 20)+
    theme(axis.text.x = element_blank())+
    facet_wrap(~Dim)+
  labs(title="Control destination - Functionally rare species")+
  ylim(-6,8))

(plot_CD <- plot_CD + 
  stat_pvalue_manual(dunn_res_all_CD, label = "p.adj.signif", size = 5, p.digits = 3, tip.length = 0.1, hide.ns = T, color="black", bracket.size = 1))


ns_dims <- dunn_res_all_CD %>%
  group_by(Dim) %>%
  summarise(all_ns = all(p.adj.signif == "ns")) %>%
  filter(all_ns) %>%
  pull(Dim)

ns_labels <- data.frame(
  Dim = ns_dims,
  label = "ns"
)

plot_CD +
  geom_text(data = ns_labels, 
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.2, vjust = 1.5, 
            size = 5, color = "grey40",
            inherit.aes = FALSE)
```






##essai script disctictiveness orientéee

*Directional distinctiveness* 
Idée : Di est toujours positif mais la position sur PC1
 donc indique si l'espèce est "distinctive vers le grand/gras" (Dim.1 élevé = grande espèce) ou (Dim.1 faible = petite espèce) --> Di_directional = Di_global × sign(VH_axis)
*positive* = distinctivement grande, 
*negative* = distinctivement petite
  
```{r}
# ESPÈCES Frare présentes en CO ET en Warm, les 105
Frare_W_W <- intersect(Frare_W, Species_W)
length(Frare_W_W)

#Coordonnées PCA après rotation Varimax ────────────────────
scores <- common_W_W_rare %>% dplyr::select(Species, Dim1, Dim2)

df <-  common_W_W_rare %>%
  mutate(
    Di_directional = global_di * sign(Dim1))

head(df)
df %>% count(statut)
summary(df$Di_directional)

# TESTS — emmean ~ Di_directional par rapport à juste di_global
mod_di_dir <- lm(emmean ~ Di_directional, data = df)
summary(mod_di_dir)

mod_di_glob <- lm(emmean ~ global_di, data = df)
summary(mod_di_glob)

#fig emmean ~ Di_directional

p1 <- ggplot(df,
             aes(x = Di_directional, y = emmean,
                 colour = statut)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "black", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey40") +
  scale_Wlour_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey60",
    "loser"  = "#f452ad"
  )) +
  annotate("text",
           x    = max(df$Di_directional, na.rm = TRUE) * 0.7,
           y    = max(df$emmean, na.rm = TRUE) * 0.9,
           label = paste0("R² = ",
                          round(summary(mod_di_dir)$r.squared, 3),
                          "\np = ",
                          signif(summary(mod_di_dir)$coefficients[2, 4], 2)),
           size = 3.5) +
  labs(x      = "Directional distinctiveness (Di_global × sign(Dim1))",
       y      = "r (in warm)",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p1

# Di_directional basé sur l'axe SLA (acquisitif vs conservateur)
# Positif = distinctivement acquisitif (SLA élevé)
# Négatif = distinctivement conservateur (SM élevé)

df <- df %>%
  mutate(
    Di_SLA_directional = global_di * sign(Dim2),
    Di_VH_directional  = global_di * sign(Dim1)
  )

# Tester les deux
mod_sla <- lm(emmean ~ Di_SLA_directional, data = df)
mod_vh  <- lm(emmean ~ Di_VH_directional,  data = df)

summary(mod_sla)
summary(mod_vh)



df_data <- df %>%
  mutate(label_sp = ifelse(
    abs(Di_VH_directional) > quantile(abs(Di_VH_directional),
                                      0.8, na.rm = TRUE) |
    abs(emmean) > quantile(abs(emmean),
                             0.8, na.rm = TRUE),
    Species, ""
  ))


###avec abundance moyenne en CO
data_plot$Species<- str_replace(data_plot$Species, "[_]", " ")

abundance_mean <- data_plot%>%
    filter(Species %in% Frare_W_W)%>%
  filter(Treatment =="Control_origin")%>% #or control_origin? 
  group_by(Species)%>%
  summarise(
    mean_abund = mean(Abs_Cover_initial_adj)
  )

df<- df%>%
  left_join(abundance_mean, by="Species")

mod_di_N <- lm(mean_abund ~ Di_directional, data = df)
summary(mod_di_N)

mod_di_N_glob <- lm(mean_abund ~ global_di, data = df)
summary(mod_di_N_glob)

p2_abundance <- ggplot(df,
             aes(x = Di_directional, y = mean_abund,
                 colour = statut)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "black", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey40") +
  scale_Wlour_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey60",
    "loser"  = "#f452ad"
  )) +
  annotate("text",
           x    = 3,
           y    = 40,
           label = paste0("R² = ",
                          round(summary(mod_di_N)$r.squared, 3),
                          "\np = ",
                          signif(summary(mod_di_N)$coefficients[2, 4], 2)),
           size = 3.5) +
  labs(x      = "Directional distinctiveness (Di_global × sign(Dim1))",
       y      = "mean initial abundance in CO",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p2_abundance

#plot both
p1 |p2_abundance


```

#growth form
```{r}
data_GF<- read_csv("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/TRY/try_clean_transplant.csv") %>%
  dplyr::select(Family, Genus, Species, GF)%>%
  filter(!is.na(Family))

#or Diaz data
library(readxl)

data_GF <- read_excel("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/copula/data/TRY_Diaz2016.xlsx")%>%
  dplyr::select(Family, Genus, `Species name standardized against TPL`, `Growth Form`)%>%
  rename(Species = `Species name standardized against TPL`) 

head(data_GF)

  essai_total <- pca_all_treatments%>%
  left_join(data_GF, by="Species")%>%
    filter(Rarity=="Frare")%>%
    filter(Treatment=="warm")

names(essai_total)


essai_total_clean <- essai_total %>%
  filter(!is.na(`Growth Form`)) %>%
  mutate(
    statut      = factor(statut,
                         levels = c("winner", "stable", "loser")),
    growth_form = factor(`Growth Form`,
                         levels = c("climber",
                                    "herbaceous graminoid",
                                    "herbaceous non-graminoid",
                                    "herbaceous non-graminoid/shrub",
                                    "shrub",
                                    "fern"))
  )


count_gf <- essai_total_clean %>%
  count(statut, growth_form) %>%
  group_by(growth_form) %>%
  mutate(
    n_total = sum(n),
    pct     = n / n_total * 100
  ) %>%
  ungroup()

print(count_gf)

#Chi² Growth Form × statut
chisq_gf <- chisq.test(
  xtabs(~ growth_form + statut,
        data = essai_total_clean)
)
print(chisq_gf)
round(chisq_gf$residuals, 2)

fig_gf <- count_gf %>%
  ggplot(aes(x = pct, y = reorder(growth_form, pct),
             fill = statut)) +
  geom_Wl(position = "stack") +
  geom_text(aes(label = ifelse(pct > 8,
                               paste0(round(pct, 0), "%"), "")),
            position = position_stack(vjust = 0.5),
            colour = "white", fontface = "bold", size = 3) +
  geom_text(
    data = count_gf %>% distinct(growth_form, n_total),
    aes(x = 102, y = growth_form,
        label = paste0("n=", n_total)),
    inherit.aes = FALSE,
    size = 3, hjust = 0, colour = "grey40"
  ) +
  scale_x_Wntinuous(limits = c(0, 115),
                     labels = function(x)
                       ifelse(x <= 100, paste0(x, "%"), "")) +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  labs(
    x        = "% species",
    y        = NULL,
    fill     = NULL,
    title    = "Growth form of functionally rare species present under Warm",
    subtitle = paste0("Chi² p = ",
                      signif(chisq_gf$p.value, 2))
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

fig_gf


ggplot(essai_total_clean) +
  aes(x = growth_form, y = global_di, fill = Rarity) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal()
```

#C3C4

enfait they are all C3 species!
```{r}
wright <- read_excel("C:/Users/boisseaux/Dropbox/Mon PC (Jaboty20)/Postdoc_CEFE/copula/data/Wright_2004_glopnet_data.xls", col_names = F)

wright <- wright[-c(1,2,3,4,5,6, 7, 8, 9, 10),]

colnames(wright) <- wright[1,]

wright <- wright[-1,]

names(wright)

wright<- wright %>%
  dplyr::select(Species, GF, BIOME, `Decid/E'green`,  `Needle/Broad lf`,  `C3C4`, `N2-fixer`)


Frare_c3c4<- pca_warm_on_global2 %>%
  left_join(wright, by="Species")

#add c3c4 data from Alec Baird resolving Nature paper

c3c4 <- read_excel("raw data/c3c4.xlsx") 

names(c3c4)

Frare_c3c4_add<- Frare_c3c4 %>%
  left_join(c3c4, by="Species")


#add c3c4 from https://www.frontiersin.org/journals/ecology-and-evolution/articles/10.3389/fevo.2021.760118/full
c3c4_new <- read_excel("raw data/Copie de Supplementary material S3_c3c4.xlsx", sheet= "Sheet 1") %>%dplyr::select(Genus, Metabolism)

Frare_c3c4_add <- Frare_c3c4_add%>%
  separate(Species, into = c("Genus", "Species"), sep = " ")

Frare_c3c4_add2<- Frare_c3c4_add %>%
  left_join(c3c4_new, by="Genus")%>%
  mutate(Metabolism2 = coalesce(Metabolism, C3C4),
        Species2 = paste0(Genus, " ", Species))%>%
  dplyr::select(-Species)%>%
  rename(Species= Species2)%>%
  left_join(data_scaled%>% dplyr::select(Family, Species), by= "Species")
  
Frare_c3c4_add3 <- Frare_c3c4_add2%>%
  dplyr::select(Family, Genus, Species, Metabolism2)%>%
  filter(Family =="Cyperaceae")
  
Frare_c3c4_add3

```



#UICN

"We used the taxize package v.0.7.899 to retrieve up-to-date IUCN
status for mammals and birds79. For easier interpretation, we grouped species in
three categories: Critically Endangered (CR), Endangered (EN), and Vulnerable
(VU) species as “Threatened” (TH); Least Concern (LC) and Near Threatened
(NT) species as “Least Concern” (LC); Data Deficient (DF) and Not Evaluated
species as “Not Evaluated” (NE). We performed a multiple comparison Kruskal
Wallis rank-sum test to compare indices across IUCN categories.
```{r}
library(taxize)
get_iucn("Branta canadensis")
get_iucn("Branta bernicla")
get_iucn("Panthera uncia")

as.iucn(essai_total$Species, check = TRUE, key = NULL)


##from a downloaded version of UICN plant species' lsit on 09/09/2026
library(readr)
UICN <- read_csv("raw data/UICN/simple_summary.csv")

names(UICN)

UICN <- UICN %>%
  dplyr::select(scientificName, redlistCategory, populationTrend, scopes) %>%
  rename(Species = scientificName)

Frare_uicn<- pca_all_treatments %>%
  filter(Rarity == "Frare")%>%
  filter(Treatment=="warm")%>%
  left_join(UICN, by="Species")

```




##distinctivness per region
```{r}
#distincitveness per region ___________________________________________________
pres_matrix_region <- data_plot %>%
  mutate(Species = gsub("_", " ", Species)) %>%
  group_by(Region, Species) %>%
  summarise(
    cover = sum(Abs_Cover_initial_adj + Abs_Cover_final_adj, na.rm = TRUE),#not sure
    .groups = "drop"
  ) %>%
  mutate(
    Presence = if_else(cover > 0, 1, 0))%>%
  complete(Region, Species, fill= list(Presence =  0, cover=0))%>%
  dplyr::select(-cover)%>%
  pivot_wider(names_from = Species, values_from = Presence)

rownames(pres_matrix_region) <- pres_matrix_region$Region
pres_matrix_region <- pres_matrix_region[,-1]

distinctiveness_result_region <- funrar::distinctiveness(
  pres_matrix = as.matrix(pres_matrix_region),
  dist_matrix = Distance,
  relative = F)

rownames(distinctiveness_result_region) <- pres_matrix_region$Region

#Format long 
di_long_region <- as.data.frame(distinctiveness_result_region) %>%
  rownames_to_Wlumn("Region") %>%
  pivot_longer(-Region, names_to = "Species", values_to = "Di")%>%
  rename(Di_region = Di)%>%
  filter(!is.na(Di_region))


# quantiles pour définir rare / intermédiaire / commun
q_region <- quantile(di_long_region$Di_region, probs = c(0.25, 0.75))

di_long_region <- di_long_region %>%
  mutate(
    Rarity_region= case_when(
      Di_region >= q_region[2] ~ "Frare",
      Di_region <=q_region[1] ~"Fcommon",
      T ~ "average"))

n_rare <- di_long_region%>%
  filter(Rarity_region == "Frare")%>%
  summarise(
    n_rare = n()) 

n_Wmmon <- di_long_region%>%
  filter(Rarity_region == "Fcommon")%>%
  summarise(
    n_Wmmon = n()) 

##merge
di_long <- di_long %>% 
  rename(Di_global = Di,
         Rarity_global=Rarity)

rarity_all <- di_long_region %>%
  left_join(di_long, by="Species")

#OLD code
# Run global PCA on standardized traits on the 642 species
#pca_global_result <- PCA(data_global_scaled[4:8], scale.unit=T, graph=T)
# 
# # Variance explained by each axis
# fviz_eig(pca_global_result, addlabels = TRUE, ylim = c(0, 100),
#          main = "Variance explained per axis")
# 
# # Extract species coordinates in PCA space
# eig_values <- pca_global_result$eig
# significant_axes <- 1:ncol(pca_global_result$ind$coord)
# coords <- pca_global_result$ind$coord[, significant_axes, drop = FALSE]
# coord <- data.frame(Species = data_global_scaled$Species, coords)
# 
# # Visualize species and trait loadings
# fviz_pca_biplot(pca_global_result,
#                 axes = c(1, 2),
#                 geom.ind = "point",
#                 col.ind = "cadetblue3",
#                 alpha.ind = 0.6,
#                 col.var = "darkblue",
#                 arrowsize = 1.2,
#                 label = "var"
# ) +
#   theme_minimal()
# 
# 
# 
# pca_demo <- coord %>%
#   left_join(data_global_scaled, by = "Species") 
# 
# pca_demo <- data.frame(
#   Species   = data_global_scaled$Species,
#   Dim.1     = pca_global_result$ind$coord[, 1],
#   Dim.2     = pca_global_result$ind$coord[, 2]
# )
# 
# var_Words <- as.data.frame(pca_global_result$var$coord[, 1:2])
# var_Words$varname <- rownames(var_Words)
# 
# scale_factor <- max(abs(pca_demo[, c("Dim.1", "Dim.2")])) /
#                 max(abs(var_Words[, c("Dim.1", "Dim.2")])) * 0.6
# 
# #calculate distinctiveness _________________________________________________
# 
# rownames(coord) <- coord$Species
# coord$Species <- NULL
# species_names    <- rownames(coord)
# 
# 
# Distance <- compute_dist_matrix(coord,
#                                 metric = "euclidean",
#                                 center = FALSE,
#                                 scale  = FALSE)
# rownames(Distance) <- colnames(Distance) <- species_names
# 
# dim(Distance)
# 
# #Matrice de présence globale so all 1 - globalement les espèces sont toutes présentes = 1
# pres_matrix <- matrix(1, nrow=1, ncol = 642)
# rownames(pres_matrix) <- "Species"
# colnames(pres_matrix) <- species_names
# 
# 
# distinctiveness_result <- funrar::distinctiveness(
#   pres_matrix = pres_matrix,
#   dist_matrix = Distance,
#   relative = F)
# 
# #or with this function
# di_test <- funrar::distinctiveness_global(dist_obj = Distance, di_name = "global_di") #which is exactly the same thing
# 
# 
# #Format long 
# di_long <- as.data.frame(distinctiveness_result) %>%
#   rownames_to_Wlumn("Scale") %>%
#   pivot_longer(-Scale, names_to = "Species", values_to = "Di") %>%
#   select(-Scale)
# 
# max_di <- max(di_long$Di) # 0.614503
# min_di <- min(di_long$Di) #0.1688295
# 
# # quantiles pour définir rare / intermédiaire / commun
# q <- quantile(di_long$Di, probs = c(0.25, 0.75))
# 
# di_long <- di_long %>%
#   mutate(
#     Rarity= case_when(
#       Di >= q[2] ~ "Frare",
#       Di <=q[1] ~"Fcommon",
#       T ~ "average"))
# 
# n_rare <- di_long%>%
#   filter(Rarity == "Frare")%>%
#   summarise(
#     n_rare = n()) 
# 
# n_Wmmon <- di_long%>%
#   filter(Rarity == "Fcommon")%>%
#   summarise(
#     n_Wmmon = n()) 
# 
# coord <- data.frame(Species = data_global_scaled$Species, coords)
# rownames(coord) <- coord$Species
# rarity_df <- coord %>% left_join(di_long, by="Species")
# 
# #merge
# stopifnot(nrow(rarity_df) == nrow(pca_global_result$ind$coord))
# stopifnot(!any(is.na(rarity_df$Di))) 
# 
# fviz_pca_biplot(pca_global_result,
#                  axes = c(1, 2),
#                  geom.ind = "point",
#                  col.ind = rarity_df$Di,
#                  gradient.cols = c( "#ffffbf","#2c7bb6", "red"),  # low -> high Di
#                  alpha.ind = 0.8,
#                  col.var = "grey70",
#                  arrowsize = 1.2,
#                  label = "var") +
#   theme_minimal() +
#   labs(color = "Global distinctiveness")
# 
# ##Regional distinctiveness 
# #with pres_matrix of presence /absence in regions of the species 
# 
# 
# #distincitveness per region ___________________________________________________
# pres_matrix_region <- data_plot %>%
#   mutate(Species = gsub("_", " ", Species)) %>%
#   group_by(Region, Species) %>%
#   summarise(
#     cover = sum(Abs_Cover_initial_adj + Abs_Cover_final_adj, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   mutate(
#     Presence = if_else(cover > 0, 1, 0))%>%
#   complete(Region, Species, fill= list(Presence =  0, cover=0))%>%
#   select(-cover)%>%
#   pivot_wider(names_from = Species, values_from = Presence)
# 
# rownames(pres_matrix_region) <- pres_matrix_region$Region
# pres_matrix_region <- pres_matrix_region[,-1]
# 
# distinctiveness_result_region <- funrar::distinctiveness(
#   pres_matrix = as.matrix(pres_matrix_region),
#   dist_matrix = Distance,
#   relative = F)
# 
# rownames(distinctiveness_result_region) <- pres_matrix_region$Region
# 
# #Format long 
# di_long_region <- as.data.frame(distinctiveness_result_region) %>%
#   rownames_to_Wlumn("Region") %>%
#   pivot_longer(-Region, names_to = "Species", values_to = "Di")%>%
#   rename(Di_region = Di)%>%
#   filter(!is.na(Di_region))
# 
# 
# # quantiles pour définir rare / intermédiaire / commun
# q_region <- quantile(di_long_region$Di_region, probs = c(0.25, 0.75))
# 
# di_long_region <- di_long_region %>%
#   mutate(
#     Rarity_region= case_when(
#       Di_region >= q_region[2] ~ "Frare",
#       Di_region <=q_region[1] ~"Fcommon",
#       T ~ "average"))
# 
# n_rare <- di_long_region%>%
#   filter(Rarity_region == "Frare")%>%
#   summarise(
#     n_rare = n()) 
# 
# n_Wmmon <- di_long_region%>%
#   filter(Rarity_region == "Fcommon")%>%
#   summarise(
#     n_Wmmon = n()) 
# 
# ##merge
# di_long <- di_long %>% 
#   rename(Di_global = Di,
#          Rarity_global=Rarity)
# 
# rarity_all <- di_long_region %>%
#   left_join(di_long, by="Species")
```


#other tests



##boxplots
```{r}

statut_by_sp_trt <- data_status %>%
  distinct(Species, Treatment, statut) %>%
  filter(!is.na(statut))%>%
  mutate(Treatment = factor(case_when(
    Treatment == "control_origin" ~ "control\n origin",
    Treatment == "control_dest"   ~ "control\n dest",
    Treatment == "warm"           ~ "Warm"
  ), levels = c("control_origin", "control_dest", "warm")))
  

# Joindre avec les régions depuis data_plot
count_region_trt <- data_plot %>%
  distinct(Species, Treatment, Region)  %>%
  left_join(statut_by_sp_trt,
            by = c("Species", "Treatment")) %>%
  filter(!is.na(statut)) %>%
  count(Region, Treatment, statut) %>%
  group_by(Region, Treatment) %>%
  mutate(
    n_total = sum(n),
    pct     = n / n_total * 100
  ) %>%
  ungroup()

# Vérifier
count_region_trt %>%
  filter(statut == "winner") %>%
  arrange(desc(pct)) %>%
  head(10)

#winner and looser and stable per region 

all_Wmbinations <- expand.grid(
  Region    = unique(count_region_trt$Region),
  Treatment = c("Control_origin", "Control_dest", "warm"),
  statut    = c("winner", "stable", "loser"),
  stringsAsFactors = FALSE
)

# Joindre avec les données existantes
count_region_trt_Wmplete <- all_Wmbinations %>%
  left_join(
    count_region_trt %>%
      filter(statut %in% c("winner", "stable", "loser")) %>%
      dplyr::select(Region, Treatment, statut, n, pct),
    by = c("Region", "Treatment", "statut")
  ) %>%
  mutate(
    n   = ifelse(is.na(n),   0, n),
    pct = ifelse(is.na(pct), 0, pct),
    statut    = factor(statut,
                       levels = c("winner", "stable", "loser")),
    Treatment = factor(Treatment,
                       levels = c("Control_origin",
                                  "Control_dest",
                                  "warm"))
  )

count_region_trt_Wmplete %>%
  filter(Region == "US_Montana") %>%
  arrange(Treatment, statut)


fig_1A <- count_region_trt_Wmplete %>%
  ggplot(aes(x    = pct,
             y    = Region,
             fill = statut)) +
  geom_Wl(position = "stack") +
  scale_x_Wntinuous(limits = c(0, 101),
                     labels = function(x) paste0(x, "%")) +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  facet_wrap(~ Treatment, ncol = 3,
             labeller = labeller(Treatment = c(
               "Control_origin" = "Control origin",
               "Control_dest"   = "Control dest",
               "warm"           = "Warm"
             ))) +
  labs(
    x     = "% species",
    y     = NULL,
    fill  = NULL,
    title = "Winners, stable and losers per region"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

fig_1A


### ── Nombre de winners/stable/losers par région × traitement ───
count_region_n <- data_plot %>%
  distinct(Species, Treatment, Region) %>%
  mutate(Treatment = factor(case_when(
    Treatment == "Control_origin" ~ "Control_origin",
    Treatment == "Control_dest"   ~ "Control_dest",
    Treatment == "warm"           ~ "warm"
  ), levels = c("Control_origin", "Control_dest", "warm"))) %>%
  left_join(
    data_status %>%
      mutate(Species = gsub(" ", "_", Species)) %>%
      distinct(Species, Treatment, statut),
    by = c("Species", "Treatment")
  ) %>%
  count(Region, Treatment, statut) %>%
  mutate(statut = factor(statut,
                         levels = c("winner", "stable", "loser")))


# ── Nombre par traitement seulement ───────────────────────────
count_trt_n <- count_region_n %>%
  group_by(Treatment, statut) %>%
  summarise(n = sum(n), .groups = "drop")


count_region_n %>%
  ggplot(aes(x = n, y = Region, fill = statut)) +
  geom_Wl(position = "stack") +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  facet_wrap(~ Treatment, ncol = 3,
             labeller = labeller(Treatment = c(
               "Control_origin" = "Control origin",
               "Control_dest"   = "Control dest",
               "warm"           = "Warm"
             ))) +
  labs(
    x     = "Number of species",
    y     = NULL,
    fill  = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

count_trt_n %>%
  ggplot(aes(x = Treatment, y = n, fill = statut)) +
  geom_Wl(position = "stack") +
  geom_text(aes(label = n),
            position  = position_stack(vjust = 0.5),
            colour    = "black",
            fontface  = "bold",
            size      = 3.5) +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  scale_x_discrete(labels = c(
    "Control_origin" = "Control\norigin",
    "Control_dest"   = "Control\ndest",
    "warm"           = "Warm"
  )) +
  labs(
    x     = NULL,
    y     = "Number of species",
    fill  = NULL) +
  theme_minimal(base_size = 20) +
  theme(legend.position = "bottom")

prop_W <- count_trt_n %>%
  filter(Treatment =="Control_origin")%>%
  mutate(
    n_total = sum(n),
    propp = n/n_total
  )

prop_cd <- count_trt_n %>%
  filter(Treatment =="Control_dest")%>%
  mutate(
    n_total = sum(n),
    propp = n/n_total
  )


prop_w <- count_trt_n %>%
  filter(Treatment =="warm")%>%
  mutate(
    n_total = sum(n),
    propp = n/n_total
  )

prop <- rbind(prop_W, prop_cd, prop_w)
prop

prop %>%
  ggplot(aes(x = Treatment, y = propp, fill = statut)) +
  geom_Wl(position = "stack") +
  geom_text(aes(label = n),
            position  = position_stack(vjust = 0.5),
            colour    = "black",
            fontface  = "bold",
            size      = 3.5) +
  scale_fill_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  scale_x_discrete(labels = c(
    "Control_origin" = "Control\norigin",
    "Control_dest"   = "Control\ndest",
    "warm"           = "Warm"
  )) +
  labs(
    x     = NULL,
    y     = "Prop of species",
    fill  = NULL) +
  theme_minimal(base_size = 20) +
  theme(legend.position = "bottom")
```


```{r}
kruskal_warm <- kruskal.test(global_di ~ statut,
  data = pca_all_treatments %>%
    filter(Treatment == "warm"))
kruskal_W <- kruskal.test(global_di ~ statut,
  data = pca_all_treatments %>%
    filter(Treatment == "Control_origin"))
kruskal_CD <- kruskal.test(global_di ~ statut,
  data = pca_all_treatments %>%
    filter(Treatment == "Control_dest"))

kruskal_warm
kruskal_W
kruskal_CD

dunn_W <- pca_all_treatments %>%
  filter(Treatment == "Control_origin") %>%
  dunn_test(global_di ~ statut, p.adjust.method = "BH")
dunn_W

data_W <- pca_all_treatments %>% filter(Treatment == "Control_origin")

dunn_W_pos <- dunn_W %>%
  add_xy_position(data = data_W, x = "statut") %>%
  mutate(Treatment = "Control_origin")   # tags it to the right facet

# boxplot Di ~ statut × traitement ───────────────
y_max_di <- max(pca_all_treatments$global_di, na.rm = TRUE)

fig_di_statut <- pca_all_treatments %>%
  ggplot(aes(x = statut, y = global_di, fill = statut)) +
  geom_boxplot(alpha = 0.5, outlier.shape = 16, outlier.size = 0.8) +
  geom_jitter(aes(colour = statut), width = 0.1, alpha = 0.3, size = 0.8) +
  scale_fill_manual(values = c("winner" = "#adf452", "stable" = "grey70", "loser" = "#f452ad")) +
  scale_Wlour_manual(values = c("winner" = "#adf452", "stable" = "grey70", "loser" = "#f452ad")) +
  facet_wrap(~ Treatment, ncol = 3,
             labeller = labeller(Treatment = c(
               "Control_origin" = "Control origin",
               "Control_dest"   = "Control dest",
               "warm"           = "Warm"
             ))) +
  stat_pvalue_manual(
    dunn_W_pos,
    label = "p.adj.signif",
    tip.length = 0.01,
    hide.ns = TRUE
  ) +
  labs(x = NULL, y = "Functional distinctiveness (Di global)", fill = NULL,
       subtitle = "Kruskal-Wallis, Dunn post-hoc (BH-adjusted) CO only") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

fig_di_statut

#fig emmean ~ Di_global ─────────────────────
fig_scatter_di <- pca_all_treatments %>%
  ggplot(aes(x =global_di, y = emmean, colour = statut)) +
  geom_point(size = 1.5, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "black", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey40") +
  scale_Wlour_manual(values = c(
    "winner" = "#adf452",
    "stable" = "grey70",
    "loser"  = "#f452ad"
  )) +
  facet_wrap(~ Treatment, ncol = 3,
             labeller = labeller(Treatment = c(
               "Control_origin" = "Control origin",
               "Control_dest"   = "Control dest",
               "warm"           = "Warm"
             ))) +
  labs(x      = "Functional distinctiveness (Di global)",
       y      = "r",
       colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

fig_scatter_di

```





```{r}
#For unbalanced design, that is, unequal numbers of subjects in each subgroup, the recommended methods are:
# the type II ANOVA when there is no significant interaction, which can be done in R with Anova(mod, type = "II") or Anova(mod, type = 2), where mod is the name of your saved model, and
# the type III ANOVA when there is a significant interaction,

aov<- aov(Dim1~statut*Treatment, data= pca_all_treatments %>%filter(Rarity=="Frare"))
aov
summary(aov)
Anova(aov, type=2)

#check assumptions for the model
aov.res <- pca_all_treatments %>%filter(Rarity=="Frare")
aov.res$fit = fitted(aov)
aov.res$resid = resid(aov)

ggplot(data = aov.res, aes(x= fit, y = resid, color = statut))+
  geom_point()+
  facet_wrap(~Treatment)

#normality of residuals
ggplot(data=aov.res, aes(sample= resid))+ stat_qq()+ stat_qq_line()
(shapiro.test(resid(aov)))

(car::leveneTest(aov)) #We do not reject the null hypothesis that the variances are equal

#library(stats)#library(aov.resstats)
TukeyHSD(aov, which = "Treatment")
```



























