---
title: "Marion analysis"
output: html_document
date: "2026-02-09"
editor_options: 
  chunk_output_type: console
---

https://ourcodingclub.github.io/tutorials/mixed-models/

#Library
```{r}
library(tidyverse)   
library(readxl)
library(ggpubr)
library(funrar)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(TNRS) #c'est une application, pas une base de données
library(emmeans)
```

#Data

#Clean taxonomy
```{r}

# Load original dataset containing species names
load("./raw data/TransPlantNetwork.RData")

# ============================================
# 2. EXTRACT AND PREPARE SPECIES NAME LIST
# ============================================

# Extract unique species codes from dataset
species_codes_raw <- unique(dat$SpeciesName)
species_codes_df <- data.frame(code = species_codes_raw) 


# Join with taxa table to match species codes with their names
species_codes_df <- species_codes_df %>%
  left_join(spcodes %>% select(code, taxa), by = "code") %>%
  distinct()

dim(species_codes_df)

species_codes_df <- species_codes_df %>% filter(is.na(code) | code != "Equ pal/pra") #removed code equi.pal/pra as it could mean either Equisetum pratense or Equisetum palustre. 

# Replace missing taxa values by using the original code as fallback
species_codes_clean <- species_codes_df %>%
  mutate(taxa = ifelse(is.na(taxa), code, taxa))

# Create TNRS input dataframe with unique row numbers
tnrs_input <- data.frame(Row_Number = 1:nrow(species_codes_clean),
                         SpeciesName = species_codes_clean$taxa)

# ========================================
# 3. TNRS CORRESPONDENCE
# ========================================

tnrs_results <- TNRS(tnrs_input)

# ========================================
# 4. CLEANING TNRS RESULTS AND FIXING ERRORS
# ========================================
# Get the best resolved name: species or fallback to accepted name
tnrs_results <- tnrs_results %>%
  mutate(sp_final = ifelse(Accepted_species == "", 
                           ifelse(Accepted_name == "", NA, Accepted_name), #If Accepted_species is empty, use Accepted_name instead. if both empty, then NA
                           Accepted_species))

# Manual corrections for problematic species
problem_ids <- c(129, 168, 275, 441, 442, 
                 446, 447, 448, 551, 556, 
                 579, 584, 589, 591, 622, 
                 625, 636, 637, 652, 735, 
                 776, 803, 864, 875, 876, 
                 877, 878, 879, 1092, 1123, 
                 1136, 1138, 1305, 1306, 1325) #marion adding more


## TNRS Marion checked - 35 totals
tnrs_marion <- tnrs_results %>% select(ID, Name_submitted, Overall_score, Accepted_name, Accepted_species)
View(tnrs_marion)
tnrs_marion <- tnrs_marion %>%filter(Overall_score < 1)
write.table(tnrs_marion, "./intermediate data/tnrs_marion_below_tnrs_score_1.csv", row.names = F, sep = ";")

#ID129 = cannot be Daniella : Dianthus meducensis ?
#ID168 = Dianthus lanceolatus
#ID275 = Arctous alpina
#ID441 = Ranunculus acris ###already flagged
#ID442; 447 = Leontodon hispidus
#ID446; 448 = Anthyllis vulneraria
#ID551 = Poaceae ###already flagged
#ID556 = Gentiana alpina
#ID579 = Silene nutans
#ID584 = Carlina
#ID589 = Phyteuma scaposum 
#ID591 = Polygala
#ID622 = Sinapis alba ###already flagged
#ID625 = Poaceae ###already flagged
#ID636 = NA
#ID637 = Poaceae ###already flagged
#ID652 = Brassicaceae
#ID735 = Hieracium froelichianum
#ID776 = Noccaea montana
#ID803 = Androsace septentrionalis
#ID864 = NA ###already flagged
#ID875; 846; 877; 878; 879 = NA
#ID1092 = Cerastium fontanum
#ID1123 = Apicaceae
#ID1136 = NA
#ID1138 = NA
#ID1305 = Poa alpigena
#ID1306 = Poa alpigena
#ID1325 = Selaginella selaginoides


# Save old values before manual correction
manual_fixes <- tnrs_results %>%
  filter(ID %in% problem_ids) %>%
  select(ID, Name_submitted, sp_final, Accepted_family) %>%
  rename(sp_final_old = sp_final,
         Accepted_family_old = Accepted_family)

# Apply manual corrections
tnrs_results$sp_final[tnrs_results$ID %in% c(551, 625, 637, 622, 864)] <- NA
tnrs_results$Accepted_family[tnrs_results$ID %in% c(551, 625, 637, 622, 864)] <- NA

tnrs_results$sp_final[tnrs_results$ID == 441] <- "Ranunculus acris"
tnrs_results$Accepted_family[tnrs_results$ID == 441] <- "Ranunculaceae"

#Marion additional manual corrections
tnrs_results$sp_final[tnrs_results$ID %in% c(636, 875, 876, 877, 878, 879, 1136, 1138)] <- NA
tnrs_results$Accepted_family[tnrs_results$ID %in% c(636, 875, 876, 877, 878, 879, 1136, 1138)] <- NA

tnrs_results$sp_final[tnrs_results$ID == 129] <- "Dianthus meducensis"
tnrs_results$Accepted_family[tnrs_results$ID == 129] <- "Caryophyllaceae"

tnrs_results$sp_final[tnrs_results$ID == 168] <- "Dianthus lanceolatus"
tnrs_results$Accepted_family[tnrs_results$ID == 168] <- "Caryophyllaceae"

tnrs_results$sp_final[tnrs_results$ID == 275] <- "Arctous alpina"
tnrs_results$Accepted_family[tnrs_results$ID == 275] <- "Ericaceae"

tnrs_results$sp_final[tnrs_results$ID == 442] <- "Leontodon hispidus"
tnrs_results$Accepted_family[tnrs_results$ID == 442] <- "Asteraceae"

tnrs_results$sp_final[tnrs_results$ID == 447] <- "Leontodon hispidus"
tnrs_results$Accepted_family[tnrs_results$ID == 447] <- "Asteraceae"

tnrs_results$sp_final[tnrs_results$ID == 446] <- "Anthyllis vulneraria"
tnrs_results$Accepted_family[tnrs_results$ID == 446] <- "Fabaceae"

tnrs_results$sp_final[tnrs_results$ID == 448] <- "Anthyllis vulneraria"
tnrs_results$Accepted_family[tnrs_results$ID == 448] <- "Fabaceae"

tnrs_results$sp_final[tnrs_results$ID == 551] <- "Poaceae"
tnrs_results$Accepted_family[tnrs_results$ID == 551] <- "Poaceae" #already flagged but not replaced

tnrs_results$sp_final[tnrs_results$ID == 556] <- "Gentiana alpina"
tnrs_results$Accepted_family[tnrs_results$ID == 556] <- "Gentianaceae"

tnrs_results$sp_final[tnrs_results$ID == 579] <- "Silene nutans"
tnrs_results$Accepted_family[tnrs_results$ID == 579] <- "Caryophyllaceae"

tnrs_results$sp_final[tnrs_results$ID == 584] <- "Carlina"
tnrs_results$Accepted_family[tnrs_results$ID == 584] <- "Asteraceae"

tnrs_results$sp_final[tnrs_results$ID == 589] <- "Phyteuma scaposum"
tnrs_results$Accepted_family[tnrs_results$ID == 589] <- "Campanulaceae"

tnrs_results$sp_final[tnrs_results$ID == 591] <- "Polygala"
tnrs_results$Accepted_family[tnrs_results$ID == 591] <- "Polygalaceae"

tnrs_results$sp_final[tnrs_results$ID == 622] <- "Sinapis alba"
tnrs_results$Accepted_family[tnrs_results$ID == 622] <- "Brassicaceae"

tnrs_results$sp_final[tnrs_results$ID == 625] <- "Poaceae"
tnrs_results$Accepted_family[tnrs_results$ID == 625] <- "Poaceae"

tnrs_results$sp_final[tnrs_results$ID == 637] <- "Poaceae"
tnrs_results$Accepted_family[tnrs_results$ID == 637] <- "Poaceae"

tnrs_results$sp_final[tnrs_results$ID == 652] <- "Brassicaceae"
tnrs_results$Accepted_family[tnrs_results$ID == 652] <- "Brassicaceae"

tnrs_results$sp_final[tnrs_results$ID == 735] <- "Hieracium froelichianum"
tnrs_results$Accepted_family[tnrs_results$ID == 735] <- "Asteraceae"

tnrs_results$sp_final[tnrs_results$ID == 776] <- "Noccaea montana"
tnrs_results$Accepted_family[tnrs_results$ID == 776] <- "Brassicaceae"

tnrs_results$sp_final[tnrs_results$ID == 803] <- "Androsace septentrionalis"
tnrs_results$Accepted_family[tnrs_results$ID == 803] <- "Primulaceae"

tnrs_results$sp_final[tnrs_results$ID == 1092] <- "Cerastium fontanum"
tnrs_results$Accepted_family[tnrs_results$ID == 1092] <- "Caryophyllaceae"

tnrs_results$sp_final[tnrs_results$ID == 1123] <- "Apicaceae"
tnrs_results$Accepted_family[tnrs_results$ID == 1123] <- "Apicaceae"

tnrs_results$sp_final[tnrs_results$ID == 1305] <- "Poa alpigena"
tnrs_results$Accepted_family[tnrs_results$ID == 1305] <- "Poaceae"

tnrs_results$sp_final[tnrs_results$ID == 1306] <- "Poa alpigena"
tnrs_results$Accepted_family[tnrs_results$ID == 1306] <- "Poaceae"

tnrs_results$sp_final[tnrs_results$ID == 1325] <- "Selaginella selaginoides"
tnrs_results$Accepted_family[tnrs_results$ID == 1325] <- "Selaginellaceae"

# Save corrected values into manual fix table
manual_fixes$sp_final_new <- tnrs_results$sp_final[match(manual_fixes$ID, tnrs_results$ID)]
manual_fixes$Accepted_family_new <- tnrs_results$Accepted_family[match(manual_fixes$ID, tnrs_results$ID)]

# ==============================================
# 5. FORMAT CLEANED RESULTS INTO FINAL TABLE
# ==============================================

tnrs_summary <- tnrs_results %>%
  select(Name_submitted, sp_final, Accepted_family) %>%
  rename(Species = sp_final,
         Family = Accepted_family) %>%
  mutate(
    Family = ifelse(is.na(Family) | Family == "", Species, Family),
    Species = ifelse(Family == Species, NA, Species), #remove species' names that are actually family name / if family equal to species then species is NA
    Genus = ifelse(is.na(Species), NA, word(Species, 1)), #if species is na then genus is na / or extract first word
    Species = ifelse(str_count(Species, "\\S+") == 1, NA, Species), #remove if in species there is a single word
    name_species = case_when( #new column that is either filled with family/genus/species name to ensure a taxo name
      !is.na(Species) ~ Species,
      !is.na(Genus) ~ Genus,
      TRUE ~ Family
    )
  ) %>%
  select(Name_submitted, name_species, Species, Genus, Family)


# ==========================================
# 6. PLOT QUALITY OF TAXONOMIC IDENTIFICATION
# ==========================================

tnrs_quality <- tnrs_summary %>%
  mutate(niveau_id = case_when(
    !is.na(Species) ~ "Species",
    !is.na(Genus) ~ "Genus",
    !is.na(Family) ~ "Family",
    TRUE ~ "Unidentified"
  ))

identification_summary <- tnrs_quality %>%
  count(niveau_id)


#if does not work, faire dev.off()
ggplot(identification_summary, aes(x = "", y = n, fill = niveau_id)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Identification Level of Submitted Names",
       fill = "Taxonomic Level") +
  theme_void()

#remove when only identified at genus or family level
tnrs_quality_final <- tnrs_quality %>%
  filter(niveau_id =="Species") %>%
  select(-niveau_id)


# Standardize species names (replace spaces with underscores)
species_corrections <- tnrs_quality_final %>%
  mutate(name_species = gsub(" ", "_", name_species))

# Rename the column in species_corrections
species_corrections <- species_corrections %>%
  dplyr::rename(SpeciesName = Name_submitted)

#remove trees
trees = c("Acer", "Betula","Juniperus","Larix", "Quercus",
          "Picea","Pinus","Sorbus","Salix")
species_corrections_final <- species_corrections %>%
  filter(!(Genus %in% trees))

setdiff(species_corrections$Species, species_corrections_final$Species)
 
[1] "Acer pseudoplatanus" #erable sycomore -tree

[2] "Betula pubescens"   #le boulot -tree
[3] "Betula nana"   #dwarf tree 

[4] "Juniperus communis" #Genévrier commun - tree

[5] "Larix decidua" #Mélèze d'Europe - tree

[6] "Quercus robur"   #Chêne pédonculé - tree

[7] "Picea abies" #epicea commun -tree

[8] "Pinus sylvestris"   #pin sylvestre -tree

[9] "Sorbus aucuparia"   #Sorbier des oiseleurs - tree

[10] "Salix herbacea"  #c'est un arbre nain mais arbre néanmoins car produit du xylème secondaire

[11] "Salix reticulata" #has xylem 2 -tree  https://www.researchgate.net/publication/260159392_Dendrochronological_dating_of_geomorphic_processes_in_the_High_Arctic
[12] "Salix retusa"  #appears to be tree as well
[13] "Salix polaris" #has xylem 2 - tree https://www.researchgate.net/publication/260159392_Dendrochronological_dating_of_geomorphic_processes_in_the_High_Arctic
[14] "Salix lindleyana"  #looks like a tree as it has wood too
[15] "Salix serpyllifolia" #has wood - tree


#print a list of unique species names
species_unique_names <- unique(species_corrections_final$Species) %>% sort() 
write.csv(species_unique_names, "./intermediate data/unique_species_names.csv", row.names = FALSE) #899

#print a list of unique genus names
genus_unique_names <- unique(species_corrections_final$Genus) %>% sort() 
write.csv(genus_unique_names, "./intermediate data/unique_genus_names.csv", row.names = FALSE) #349

#print a list of unique family names
family_unique_names <- unique(species_corrections_final$Family) %>% sort()
write.csv(family_unique_names, "./intermediate data/unique_family_names.csv", row.names = FALSE) #67

# Export final TNRS correspondence table
write.csv(species_corrections_final, "./intermediate data/TNRS_Correspondance_marion.csv", row.names = FALSE)

# Apply corrections to main dataset (dat) - to load from the file RData - TransPlantNetwork
#but before dat has for some sites just a species code, which blocks the left_join afterwards, so we take out the part of dat (dat_subset) which has speciesName written in code, we change them and assemble dat as dat_final
species_code_df_unique <- species_codes_df %>%
  filter(!is.na(taxa)) %>%
  group_by(code) %>%
  slice(1) %>%   # garde juste la première occurrence par code
  ungroup()

dat <- dat %>%
  mutate(rowid = row_number())

code_pattern <- species_code_df_unique$code #species names written in code like Poa.pra

# Subset dat only for which spcecies Name ressembles codes
dat_subset <- dat %>%
  filter(SpeciesName %in% code_pattern)


#replace for those written in code the correspodnance
dat_subset2 <- dat_subset %>%
  left_join(
    species_code_df_unique %>% select(code, taxa),
    by = c("SpeciesName" = "code")
  ) %>%
  select(-SpeciesName)%>%
  mutate(SpeciesName = taxa) %>%
  select(-taxa)


dat_final <- bind_rows(
  dat %>% filter(!SpeciesName %in% code_pattern), 
  dat_subset2                                     
) %>%
  arrange(row_number())  # optionnel pour garder l'ordre original

dim(dat)
dim(dat_subset)
dim(dat_subset2)
dim(dat_final)

data_clean <- dat_final %>%
  left_join(species_corrections_final %>% select(SpeciesName, name_species),
            by = "SpeciesName")%>%
  mutate(SpeciesName = name_species) %>%
  select(-name_species) %>%
  filter(!is.na(SpeciesName))  #if a species is successfully corrected, then SpeciesName is NOT NA. If a species was not matched then NA. 

table(is.na(data_clean$SpeciesName))
 
# Export data_clean clean taxonomy
write.csv(data_clean, "./intermediate data/dat_clean_taxo.csv", row.names = FALSE)
```


#Metadata information

Information from Billur Bektas Ecography paper.

- 20 Different Mountain Regions of the Northern Hemisphere 
- Pays: France, Italie, Autriche, Allemagne, Suisse / Norvège, Suède / Chine / USA / Inde (IN_Kashmir)
- 22 elevation gradients
- Year 2007 - 2022
- 40 experiments (12 gradients were composed of more than two sites )
- 1 experiment =  a set of warmed, origin and destination control turfs within each gradient
- Each experiment has 3–10 replicates of transplant and control turfs, total of 668 turfs 
- Turf size 0.02 and 1 m2 (mean size of 0.3 m2 across the 40 experiments)
- https://billurbektas.shinyapps.io/transplantshiny-1/

```{r}
# Export data_clean clean taxonomy
data_clean <- read.csv("./intermediate data/dat_clean_taxo.csv")

unique(data_clean$Region) #22 regions 
 [1] "NO_Ulvhaugen"        "CH_Lavey"            "CH_Calanda"         
 [4] "CH_Calanda2"         "US_Colorado"         "US_Montana"         
 [7] "CN_Damxung"          "IN_Kashmir"          "CN_Gongga"          
[10] "CN_Heibei"           "DE_Grainau"          "DE_Susalps"         
[13] "DE_TransAlps"        "FR_AlpeHuez"         "SE_Abisko"          
[16] "FR_Lautaret"         "IT_MatschMazia1"     "IT_MatschMazia2"    
[19] "NO_Lavisdalen"       "NO_Gudmedalen"       "NO_Skjellingahaugen"
[22] "US_Arizona"   

sort(unique(data_clean$Year)) #from 2007 to 2022 

 [1] 2007 2008 2009 2010 2011 2012 2013 2014 2015 2016 2017 2018 2019 2020 2021 2022

unique(data_clean$originSiteID) #43 origin Site ID

[1] "Ulvhaugen"        "Alrust"           "Fauske"           "Hogsete"         
 [5] "Vikesland"        "Lavisdalen"       "Gudmedalen"       "Rambera"         
 [9] "Arhelleren"       "Skjellingahaugen" "Veskre"           "Ovstedal"        
[13] "CRE"              "RIO"              "MAR"              "PRA"             
[17] "Cal"              "Pea"              "Nes"              "mo"              
[21] "um"               "pf"               "Low"              "High"            
[25] "MC"               "PP"               "LOW"              "HIGH"            
[29] "A"                "H"                "L"                "M"               
[33] "3200"             "3400"             "3800"             "BT"              
[37] "EB"               "FE"               "GW"               "FP"              
[41] "SP"               "Mid"              "G"               

A<- data_clean %>% filter(originBlockID != "NA") 
unique(A$originBlockID)#94 block ID origin

 [1] "Fau5"  "1"     "2"     "3"     "4"     "5"     "A1"    "A2"   
 [9] "A3"    "A4"    "A5"    "A6"    "A7"    "H1"    "H2"    "H3"   
[17] "H4"    "H5"    "H6"    "H7"    "L1"    "L2"    "L3"    "L4"   
[25] "L5"    "L6"    "L7"    "M1"    "M2"    "M3"    "M4"    "M5"   
[33] "M6"    "M7"    "Ulv3"  "Ulv4"  "Ulv1"  "Ulv5"  "Alr1"  "Alr2" 
[41] "Alr3"  "Alr4"  "Alr5"  "Fau1"  "Fau2"  "Fau3"  "Fau4"  "Ulv2" 
[49] "Hog1"  "Hog2"  "Hog3"  "Hog4"  "Hog5"  "Vik1"  "Vik2"  "Vik3" 
[57] "Vik4"  "Vik5"  "Lav1"  "Lav2"  "Lav3"  "Lav4"  "Lav5"  "Gud3" 
[65] "Gud4"  "Ram4"  "Ram5"  "Ram6"  "Arh1"  "Arh2"  "Arh3"  "Arh4" 
[73] "Arh5"  "Gud11" "Gud5"  "Gud12" "Gud13" "Ram9"  "Ram8"  "Skj1" 
[81] "Skj2"  "Skj5"  "Ves1"  "Ves2"  "Ves3"  "Ves4"  "Ves5"  "Ovs1" 
[89] "Ovs2"  "Ovs3"  "Ovs4"  "Ovs5"  "Skj11" "Skj12"

unique(data_clean$destSiteID) #43 destSiteIF
[1] "Ulvhaugen"        "Alrust"           "Fauske"           "Hogsete"         
 [5] "Vikesland"        "Lavisdalen"       "Rambera"          "Gudmedalen"      
 [9] "Arhelleren"       "Skjellingahaugen" "Veskre"           "Ovstedal"        
[13] "CRE"              "MAR"              "PRA"              "RIO"             
[17] "Cal"              "Nes"              "Pea"              "um"              
[21] "pf"               "mo"               "Low"              "High"            
[25] "PP"               "MC"               "LOW"              "HIGH"            
[29] "M"                "A"                "L"                "H"               
[33] "3200"             "3400"             "3800"             "BT"              
[37] "EB"               "GW"               "FE"               "SP"              
[41] "FP"               "Mid"              "G"  


B<- data_clean %>% filter(destBlockID != "NA") 
unique(B$destBlockID)#98
[1] "Fau5"  "1"     "2"     "3"     "4"     "5"     "6"     "7"    
 [9] "8"     "9"     "10"    "M1"    "A1"    "M2"    "A2"    "M3"   
[17] "A3"    "M4"    "A4"    "M5"    "A5"    "M6"    "A6"    "M7"   
[25] "A7"    "L1"    "H1"    "L2"    "H2"    "L3"    "H3"    "L4"   
[33] "H4"    "L5"    "H5"    "L6"    "H6"    "L7"    "H7"    "Ulv3" 
[41] "Alr3"  "Alr4"  "Ulv4"  "Alr1"  "Ulv5"  "Alr5"  "Fau1"  "Alr2" 
[49] "Fau2"  "Fau3"  "Fau4"  "Ulv1"  "Ulv2"  "Hog1"  "Vik1"  "Vik2" 
[57] "Hog2"  "Vik3"  "Hog3"  "Hog4"  "Vik4"  "Vik5"  "Hog5"  "Lav1" 
[65] "Lav2"  "Lav3"  "Lav4"  "Lav5"  "Ram5"  "Gud4"  "Ram4"  "Arh4" 
[73] "Arh1"  "Ram6"  "Arh2"  "Arh3"  "Arh5"  "Ram8"  "Gud11" "Gud5" 
[81] "Gud12" "Ram9"  "Gud13" "Skj1"  "Ves4"  "Ves2"  "Skj2"  "Ves5" 
[89] "Skj5"  "Ovs4"  "Ves1"  "Ovs1"  "Ves3"  "Ovs2"  "Ovs3"  "Ovs5" 
[97] "Skj11" "Skj12"

unique(data_clean$destPlotID) #727 different
unique(data_clean$Treatment) #suelement LocalControl and Warm

unique(data_clean$turfID) #157 turf ID

length(unique(data_clean$UniqueID)) #3215

length(unique(data_clean$SpeciesName)) #899 different species

dim(data_clean) # 56 442 * 15 #dans Billur et al. 2024 - Ecography 60 635 observation / 967 identifed specues.

```

#Treatment and controls

Create a clean variable with Treatment as 3 levels - Warm / Control_Origin / Control_Destination
Check with elevation that experiment goes from top to bottom hill

Pour l'instant on a levels(factor(data_clean$Treatment))
[1] "LocalControl" "Warm"  

Creation de la variable Region.OriginSiteID et Region.DestSiteID car sinon a on peut avoir les mêmes noms de sites ex. "low" dans plusieurs regions, mais ce ne sont pas les mêmes.

```{r}
# Identify warm sites (both origin and destination) 
warm_sites <- data_clean %>%
  filter(Treatment == "Warm") %>% #there are 2 treatments: warm and LocalControl (origin control / destination control)
  group_by(Region)%>%
  distinct(originSiteID, destSiteID) #on ne peut differencier un site LocalControl s'il est origin ou destination que par rapport au "warm"

warm_sites$Region.OriginSiteID <- interaction(warm_sites$Region, warm_sites$originSiteID)
warm_sites$Region.destSiteID <- interaction(warm_sites$Region, warm_sites$destSiteID)

site_elevation_ref <- data_clean %>%
    filter(Treatment == "LocalControl") %>%       # Use only control plots
    group_by(Region, originSiteID) %>%                    # Each origin site
   summarise(Elevation = mean(Elevation, na.rm = TRUE), .groups = "drop") %>%
   arrange(desc(Elevation)) 

site_elevation_ref$Region.OriginSiteID<- interaction(site_elevation_ref$Region, site_elevation_ref$originSiteID)

site_elevation_origin <- data_clean %>%
  filter(Treatment== "LocalControl")%>%
  group_by(Region, originSiteID) %>%
  summarise(Elevation_origin = mean(Elevation, na.rm = TRUE), .groups = "drop")

site_elevation_origin$Region.OriginSiteID<- interaction(site_elevation_origin$Region, site_elevation_origin$originSiteID)

site_elevation_dest <- data_clean %>%
 filter(Treatment== "LocalControl")%>%
 group_by(Region, destSiteID) %>% #de toute facon 
  summarise(Elevation_dest = mean(Elevation, na.rm = TRUE), .groups = "drop")

site_elevation_dest$Region.destSiteID<- interaction(site_elevation_dest$Region, site_elevation_dest$destSiteID)

# Extraire les paires uniques Warm
warm_pairs <- data_clean %>%
  filter(Treatment == "Warm") %>%
  group_by(Region) %>%
  distinct(originSiteID, destSiteID)

warm_pairs$Region.OriginSiteID <- interaction(warm_pairs$Region, warm_pairs$originSiteID)
warm_pairs$Region.destSiteID <- interaction(warm_pairs$Region, warm_pairs$destSiteID)

warm_pairs <- warm_pairs %>%
  select(Region.OriginSiteID, Region.destSiteID)

# Joindre les élévations
warm_pairs <- warm_pairs %>%
  left_join(site_elevation_origin %>% select(Elevation_origin, Region.OriginSiteID), by = "Region.OriginSiteID") %>%
  left_join(site_elevation_dest %>% select(Elevation_dest, Region.destSiteID), by = "Region.destSiteID") %>%
  arrange(Elevation_origin)  

warm_pairs <- warm_pairs %>%
  mutate(downhill = Elevation_origin > Elevation_dest,
         diff = Elevation_origin - Elevation_dest)

print(warm_pairs, n=44)

# Export elevation data
write.csv(warm_pairs, "./intermediate data/delevation_warm_pairs.csv", row.names = FALSE)

# Reclassify control treatments 
data_clean$Region.OriginSiteID <- interaction(data_clean$Region, data_clean$originSiteID)
data_clean$Region.destSiteID <- interaction(data_clean$Region, data_clean$destSiteID)


data_clean <- data_clean %>%
  mutate(Treatment = case_when(
    Treatment == "Warm" ~ "Warm",
    Treatment == "LocalControl" & Region.OriginSiteID %in% warm_sites$Region.OriginSiteID ~ "Control_Origin",
    Treatment == "LocalControl" & Region.destSiteID %in% warm_sites$Region.destSiteID ~ "Control_Destination",
    TRUE ~ Treatment
  )) 

#controlling whether we do have the three levels
levels(factor(data_clean$Treatment))

# Export data_clean clean taxonomy
write.csv(data_clean, "./intermediate data/dat_clean_taxo_3treatments.csv", row.names = FALSE)
```

[1] "Control_Destination" "Control_Origin"      "Warm"   


#Number of years surveyed

## -Were all plots within a region surveyed in the same years?

*Diagnostic chekcs*

```{r}
data_clean <- read.csv("./intermediate data/dat_clean_taxo_3treatments.csv")

#look at whether all plots in the same region had been surveyed equally. 
#resultats : no ! in one region, some plots were not measured as often as others.

plot_year_list <- data_clean %>%
  distinct(Region, destPlotID, Year) %>%
  group_by(Region, destPlotID) %>%
  summarise(years = list(sort(Year)), .groups = "drop")

region_reference <- plot_year_list %>%
  group_by(Region) %>%
  summarise(region_years = list(years[[1]]), .groups = "drop")

plots_mismatch <- plot_year_list %>%
  left_join(region_reference, by = "Region") %>%
  filter(!purrr::map2_lgl(years, region_years, identical))


```

## -Keep plots with a least 3 years of monitoring

Since there are inconsistencies in the monitoring of plots, we decide to standardize by at least 3 years of monitoring.

```{r}
#okay donc decider de ne garder que les plots avec 3 années de suivis
plot_years <- data_clean %>%
  distinct(Region, destPlotID, Year) %>%
  group_by(Region, destPlotID) %>%
  summarise(n_years = n(), .groups = "drop") #sur les 756 destPlotIDs

plots_keep <- plot_years %>%
  filter(n_years >= 3) %>%#keep 692
  filter(Region != "US_Arizona") #671, as in Billur's ecography paper remove 

data_clean_years_filtered <- data_clean %>%
  semi_join(plots_keep, by = c("Region", "destPlotID"))

table(plot_years$n_years) #le nombre d’années de suivi par plot

```

before with *Lula's* script
2   3   4   5   6   7   8   9 
29  70  62 178 241 101  58  17 #sum =756

*marion*
 2   3   4   5   6   8 
 64  96 222 217 140  17 

Keep only plots with ≥ 3 years

## -survey summary data

```{r}
#restructure data to visaully see when plots wre surveyed

survey_dates <- data_clean_years_filtered %>%
  select(Region.OriginSiteID, Region.destSiteID, destPlotID, Year) %>%
  distinct() %>%
  mutate(value = "surveyed") %>%
  pivot_wider(names_from = Year, values_from = value, values_fill = NA)

year_columns <- sort(setdiff(names(survey_dates), c("Region.destSiteID", "Region.OriginSiteID", "destPlotID")))

year_columns

survey_dates <- survey_dates %>%
  mutate(n_years_surveyed = rowSums(across(all_of(year_columns)) == "surveyed", na.rm = TRUE))  # Add number of survey years to table 

# Extract first and last survey years 
survey_long <- survey_dates %>%
  select(Region.OriginSiteID, Region.destSiteID, all_of(year_columns), destPlotID) %>%
  pivot_longer(
    cols = all_of(year_columns),
    names_to = "Year",
    values_to = "surveyed"
  ) %>%
  mutate(Year = as.numeric(Year)) %>%
  filter(!is.na(surveyed))   # ne garder que les années avec données

# number of years surveyed, beg and end
survey_summary <- survey_long %>%
  group_by(Region.OriginSiteID, Region.destSiteID, destPlotID) %>%
  summarise(
    n_years_surveyed = n(),
    first_year = min(Year),
    last_year = max(Year),
    .groups = "drop"
  )

#keeping data only for those years surveyes at leasst 3 years
data_clean_filtered <- data_clean %>%
  semi_join(survey_summary, by = c("Region.OriginSiteID", "Region.destSiteID", "destPlotID"))

colnames(data_clean_filtered)

unique(data_clean_filtered$Region) #only *19 sites* surveyes at least *3* years (no US_Arizona, US_colorado, IN_Kashmir) #same as in Bektas et al., 2024 ecography
```

[1] "Region"              "Year"                "originSiteID"       
 [4] "originBlockID"       "destSiteID"          "destBlockID"        
 [7] "destPlotID"          "Treatment"           "turfID"             
[10] "UniqueID"            "SpeciesName"         "Cover"              
[13] "Rel_Cover"           "Elevation"           "rowid"              
[16] "Region.OriginSiteID" "Region.destSiteID"  



#Get metadata tables 

```{r}
#using climdata from
load("./raw data/TransPlantNetwork.RData")

# Gradient-level metadata
explist = list(CH_Calanda, CH_Calanda2, CH_Lavey, #Switzerland
               CN_Damxung, CN_Gongga, CN_Heibei, #China
               DE_Grainau, DE_Susalps, DE_TransAlps, #Germany
               FR_AlpeHuez, FR_Lautaret, #France
               IN_Kashmir, #India
               IT_MatschMazia1, IT_MatschMazia2, #Italy
               NO_Gudmedalen, NO_Lavisdalen, NO_Skjellingahaugen, NO_Ulvhaugen, #Norway
               SE_Abisko, #Sweden
               US_Arizona, US_Colorado, US_Montana #United States
) 

# Get the metadata per experiment
metadata = map_dfr(explist, ~ .x$meta)
metadata =
  metadata %>%
  mutate(YearEstablished = ifelse(Gradient == "FR_Lautaret",2016,YearEstablished))%>% # Correction to Lautaret experiment
  mutate(Elevation = ifelse(Gradient == "CH_Calanda" & Elevation == 2800, 2600, Elevation))%>% # Correction to Calanda experiment
  rename(Region = Gradient,
         PlotSize = PlotSize_m2)

# Save outputs
write.csv(metadata, "./intermediate data/metadata.csv", row.names = FALSE)
```


#Absolute and Relative Cover matrices

Cover matrices for each year
```{r}

## Re-calculate the relative cover as in BIllur's github : https://github.com/billurbektas/TransPlant_convergence_analysis/blob/main/R/02_organize.R
#This means that Cover in dat is the absolute cover. Total_Cover = somme de toutes les Cover dans le plot (toutes les espèces). Rel_Cover = fraction du cover totale occupée par cette ligne (ici ligne devient l'espèce) dans le plot. Donc Rel_Cover = abondance relative de l’espèce dans le plot.

data_cover <- data_clean_filtered %>%  
  select(-Elevation)%>%
  group_by(Region, Year, originSiteID, originBlockID, destSiteID, destBlockID, destPlotID, Treatment, turfID, UniqueID)%>%
  mutate(Total_Cover = sum(Cover, na.rm = TRUE), Rel_Cover = Cover/Total_Cover)%>%
  ungroup() %>%
   filter(!is.na(Rel_Cover)) %>%#4 NAs in DE_Susalps.
 filter(!is.na(SpeciesName))%>%
   filter(!(Region=="DE_Susalps" & Year==2016)) %>% # Not enough data (from Billur)
  filter(!(Region=="CH_Calanda2" & Year==2020)) # OriginControl not counted (from Billur)
 
metadata <- read.csv("./intermediate data/metadata.csv")


data_cover <- data_cover%>%
    left_join(metadata %>% dplyr::select(Region, destSiteID, YearEstablished, YearMin), by = c("Region","destSiteID"))
  
  ##Billur does also this, but I am not sure, il n'y a jamais de Year_0 negatif and I do not understand how Year=0 is a pre-transplantation survey: 
# %>%
#    mutate(Year_0 = Year-YearEstablished)%>%
#  filter(Year_0 != 0)%>% # We do not take pre-transplantation vegetation surveys
#   dplyr::select(-(YearEstablished:Year_0))



```

#Ultimate clean from Billur
```{r}
# Unresolved botanical issues -> we assign genus names
# CN_Damxung: Stipa capillacea, Stipa capillata -> Stipa

#UPDATE from TNRS tool: both are accepted

# CN_Damxung: Polygonaceae -> drives the pattern too much and all of a sudden disappears to be taken out. 
# DE_Susalps (GW-FE): Festuca pratensis (in destination control plots)? Elymus repens (in origin control plots)? -> drives the pattern too much and all of a sudden disappears to be taken out.
# DE_Susalps (GW-BT): Festuca pratensis (in destination control plots)? Elymus repens (in origin control plots)? -> drives the pattern too much and all of a sudden disappears to be taken out.

data_cover2 <- data_cover%>%
  #mutate(SpeciesName = ifelse(Region == "CN_Damxung" & SpeciesName %in% c("Stipa capillacea","Stipa capillata"), "Stipa", SpeciesName))%>%
  mutate(Rel_Cover = ifelse(Region == "CN_Damxung" & SpeciesName == "Polygonaceae", 0,Rel_Cover))%>%
  mutate(Rel_Cover = ifelse(Region == "DE_Susalps" & originSiteID == "FE" & destSiteID == "FE" &  SpeciesName == "Festuca pratensis", 0, Rel_Cover))%>%
  mutate(Rel_Cover = ifelse(Region == "DE_Susalps" & originSiteID == "BT" & destSiteID == "BT" & SpeciesName == "Festuca pratensis", 0, Rel_Cover))%>%
  mutate(Rel_Cover = ifelse(Region == "DE_Susalps" & originSiteID == "GW" & destSiteID == "GW" & SpeciesName == "Elymus repens", 0, Rel_Cover))%>%
  filter(Rel_Cover != 0)

# I take out species that occurred in communities only once over the whole experimental duration.
## This is because we cannot distinguish if there is an botanical identification error or true presence.
### However, I do not change the relative cover as the species contributed to the total cover. 
data_cover3 <- data_cover2 %>%
  group_by(Region, originSiteID, destSiteID, Treatment, SpeciesName) %>%
  mutate(n_occ = n()) %>%
  filter(n_occ > 1) %>%
  ungroup() %>%
  select(-n_occ) #removing 106 of DE_Susalps


# save outputs
write.csv(data_cover3, "data_cover.csv", row.names = FALSE)

# Nest the data per experiment
dfx = data_cover3 %>%
  dplyr::select(Region, originSiteID, destSiteID, Treatment) %>% 
  distinct() %>% 
  filter(Treatment == "Warm") %>% 
  dplyr::select(-Treatment) %>%
  mutate(comm = pmap(.l = list(R = Region, O = originSiteID, D = destSiteID), .f = function(R, O, D){
    bind_rows(
      originControls = data_cover2 %>% filter(Region == R & originSiteID == O & destSiteID == O & Treatment == "LocalControl"),
      destControls = data_cover2%>% filter(Region == R &  originSiteID == D & destSiteID == D & Treatment == "LocalControl"),
      warmed =  data_cover2 %>% filter(Region == R & originSiteID == O & destSiteID == D & Treatment == "Warm"),
      .id = "ODT")
  })) 


```

#species-site matrix
```{r}

#relative abundance 

site_species <- data_cover3 %>%
  group_by(Region, originSiteID, destSiteID, destPlotID, Year, Treatment, SpeciesName) %>%
  summarise(Rel_Cover = sum(Rel_Cover, na.rm = TRUE), .groups = "drop")


site_species_rel_abund <- site_species %>%
  pivot_wider(
    names_from = SpeciesName,
    values_from = Rel_Cover,
    values_fill = 0 )

#### absolute abundance

site_species <- data_cover3 %>%
  group_by(Region, originSiteID, destSiteID, destPlotID, Year, Treatment, SpeciesName) %>%
  summarise(Cover = sum(Cover, na.rm = TRUE), .groups = "drop")

site_species_abs_abund <- site_species %>%
  pivot_wider(
    names_from = SpeciesName,
    values_from = Cover,
    values_fill = 0 )


# save outputs - global outputs
write.csv(site_species_rel_abund, "./intermediate data/site_species_rel_abund.csv", row.names = FALSE)
write.csv(site_species_abs_abund, "./intermediate data/site_species_abs_abund.csv", row.names = FALSE)
```


#Final - initial relative abundance matrix

```{r}
#relative abundance per treatement
site_species_warm <- read.csv("./intermediate data/site_species_rel_abund.csv", sep = ",") %>%
  filter(Treatment =="Warm")

site_species_CO <- read.csv("./intermediate data/site_species_rel_abund.csv", sep = ",") %>%
  filter(Treatment =="Control_Origin")

site_species_CD <- read.csv("./intermediate data/site_species_rel_abund.csv", sep = ",") %>%
  filter(Treatment =="Control_Destination")

#Delta Cover for each treatement 
###get matrix for initial year of survey per Plot ID
site_species_warm_initial <- site_species_warm %>%
 group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
#  filter((max(Year) - min(Year)) >= 3) %>% #mais normalement on a déja fait ce filtre plus haut
  filter(Year == min(Year)) %>%
  ungroup() %>%
  rename(Year_initial = Year)

site_species_CO_initial <- site_species_CO %>%
 group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
 # filter((max(Year) - min(Year)) >= 3) %>% 
  filter(Year == min(Year)) %>%
  ungroup() %>%
  rename(Year_initial = Year)

site_species_CD_initial <- site_species_CD %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
#  filter((max(Year) - min(Year)) >= 3) %>% 
  filter(Year == min(Year)) %>%
  ungroup() %>%
  rename(Year_initial = Year)

###get matrix for last year of survey per plot ID
site_species_warm_final <- site_species_warm %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
 # filter((max(Year) - min(Year)) >= 3) %>%
  filter(Year == max(Year)) %>%
  ungroup() %>%
  rename(Year_final = Year)

site_species_CO_final <- site_species_CO %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
 # filter((max(Year) - min(Year)) >= 3) %>%
  filter(Year == max(Year)) %>%
  ungroup() %>%
  rename(Year_final = Year)

site_species_CD_final <- site_species_CD %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
 # filter((max(Year) - min(Year)) >= 3) %>%
  filter(Year == max(Year)) %>%
  ungroup() %>%
  rename(Year_final = Year)

# Convert from wide to long format for initial and final datasets
#WARM
site_species_warm_initial_long <- site_species_warm_initial %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Rel_Cover_initial") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Rel_Cover_initial, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

site_species_warm_final_long <- site_species_warm_final %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Rel_Cover_final") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Rel_Cover_final, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

#Control origin
site_species_CO_initial_long <- site_species_CO_initial %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Rel_Cover_initial") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Rel_Cover_initial, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

site_species_CO_final_long <- site_species_CO_final %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Rel_Cover_final") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Rel_Cover_final, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

#Control destination
site_species_CD_initial_long <- site_species_CD_initial %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Rel_Cover_initial") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Rel_Cover_initial, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

site_species_CD_final_long <- site_species_CD_final %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Rel_Cover_final") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Rel_Cover_final, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

# Merge initial and final datasets
site_species_warm_end <- bind_cols(
  site_species_warm_initial_long,
  site_species_warm_final_long %>% dplyr::select(Year_final, Rel_Cover_final)
) %>%relocate(Year_final, .after=Year_initial)

site_species_CO_end <- bind_cols(
  site_species_CO_initial_long,
  site_species_CO_final_long %>% dplyr::select(Year_final, Rel_Cover_final)
) %>%relocate(Year_final, .after=Year_initial)

site_species_CD_end <- bind_cols(
  site_species_CD_initial_long,
  site_species_CD_final_long %>% dplyr::select(Year_final, Rel_Cover_final)
) %>%relocate(Year_final, .after=Year_initial)


# Remove 0/0 cases (species absent both times)
# Replace 0 with small value if species appears later (to enable log calculations)
site_species_warm_end_clean <- site_species_warm_end %>%
  filter(!(Rel_Cover_initial == 0 & Rel_Cover_final == 0)) %>%
  mutate(Rel_Cover_initial = ifelse(Rel_Cover_initial == 0 & Rel_Cover_final != 0, 0.00120, Rel_Cover_initial))

site_species_CO_end_clean <- site_species_CO_end %>%
  filter(!(Rel_Cover_initial == 0 & Rel_Cover_final == 0)) %>%
  mutate(Rel_Cover_initial = ifelse(Rel_Cover_initial == 0 & Rel_Cover_final != 0, 0.00120, Rel_Cover_initial))

site_species_CD_end_clean <- site_species_CD_end %>%
  filter(!(Rel_Cover_initial == 0 & Rel_Cover_final == 0)) %>%
  mutate(Rel_Cover_initial = ifelse(Rel_Cover_initial == 0 & Rel_Cover_final != 0, 0.00120, Rel_Cover_initial))

#save outputs
write.csv(site_species_warm_end_clean, "./intermediate data/site_species_warm_end_clean_rel_abund_cover_initial_final.csv", row.names = FALSE)
write.csv(site_species_CO_end_clean, "./intermediate data/site_species_CO_end_clean_rel_abund_cover_initial_final.csv", row.names = FALSE)
write.csv(site_species_CD_end_clean, "./intermediate data/site_species_CD_end_clean_rel_abund_cover_initial_final.csv", row.names = FALSE)
```

#Final - initial Absolute abundance

```{r}
site_species_abs_abund <- read.csv("./intermediate data/site_species_abs_abund.csv")

#ABSOLUTE abundance per treatement
site_species_warm <- site_species_abs_abund %>%
  filter(Treatment =="Warm")

site_species_CO <- site_species_abs_abund %>%
  filter(Treatment =="Control_Origin")

site_species_CD <- site_species_abs_abund%>%
  filter(Treatment =="Control_Destination")

#Delta Cover for each treatement 
###get matrix for initial year of survey per Plot ID
site_species_warm_initial <- site_species_warm %>%
 group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
  filter(Year == min(Year)) %>%
  ungroup() %>%
  rename(Year_initial = Year)

site_species_CO_initial <- site_species_CO %>%
 group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
  filter(Year == min(Year)) %>%
  ungroup() %>%
  rename(Year_initial = Year)

site_species_CD_initial <- site_species_CD %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
  filter(Year == min(Year)) %>%
  ungroup() %>%
  rename(Year_initial = Year)

###get matrix for last year of survey per plot ID
site_species_warm_final <- site_species_warm %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
  filter(Year == max(Year)) %>%
  ungroup() %>%
  rename(Year_final = Year)

site_species_CO_final <- site_species_CO %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
  filter(Year == max(Year)) %>%
  ungroup() %>%
  rename(Year_final = Year)

site_species_CD_final <- site_species_CD %>%
  group_by(Region, originSiteID, destSiteID, destPlotID) %>% 
  filter(Year == max(Year)) %>%
  ungroup() %>%
  rename(Year_final = Year)

# Convert from wide to long format for initial and final datasets
#WARM
site_species_warm_initial_long <- site_species_warm_initial %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Abs_Cover_initial") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Abs_Cover_initial, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

site_species_warm_final_long <- site_species_warm_final %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Abs_Cover_final") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Abs_Cover_final, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

#Control origin
site_species_CO_initial_long <- site_species_CO_initial %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Abs_Cover_initial") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Abs_Cover_initial, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

site_species_CO_final_long <- site_species_CO_final %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Abs_Cover_final") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Abs_Cover_final, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

#Control destination
site_species_CD_initial_long <- site_species_CD_initial %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Abs_Cover_initial") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Abs_Cover_initial, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

site_species_CD_final_long <- site_species_CD_final %>%
  pivot_longer(cols = 7:ncol(.),
               names_to = "Species",
               values_to = "Abs_Cover_final") %>%
  distinct(Region, originSiteID, destSiteID, destPlotID, Species, Treatment, Abs_Cover_final, .keep_all = TRUE) %>%
  mutate(across(c(Region, originSiteID, destSiteID, destPlotID, Species, Treatment), as.character))

# Merge initial and final datasets
site_species_warm_end <- site_species_warm_initial_long %>%
  left_join(site_species_warm_final_long %>% 
              dplyr::select(Region, originSiteID, destSiteID, 
                    destPlotID, Species, Treatment,
                    Year_final, Abs_Cover_final),
    by = c("Region", "originSiteID", "destSiteID", 
           "destPlotID", "Species", "Treatment")
  )

site_species_CO_end <- site_species_CO_initial_long %>%
  left_join(site_species_CO_final_long %>% 
               dplyr::select(Region, originSiteID, destSiteID, 
                    destPlotID, Species, Treatment,
                    Year_final, Abs_Cover_final),
    by = c("Region", "originSiteID", "destSiteID", 
           "destPlotID", "Species", "Treatment")
  )

site_species_CD_end <- site_species_CD_initial_long %>%
  left_join(site_species_CD_final_long %>% 
               dplyr::select(Region, originSiteID, destSiteID, 
                    destPlotID, Species, Treatment,
                    Year_final, Abs_Cover_final),
    by = c("Region", "originSiteID", "destSiteID", 
           "destPlotID", "Species", "Treatment")
  )

# Remove 0/0 cases (species absent both times)
data_full_warm <- site_species_warm_end %>%
  filter(!(Abs_Cover_initial == 0 & Abs_Cover_final == 0)) %>%  # retire seulement les 0/0
  mutate(
    dynamique_statut = case_when(
      Abs_Cover_initial == 0   & Abs_Cover_final > 0  ~ "Colonizer",
      Abs_Cover_final   == 0   & Abs_Cover_initial > 0 ~ "Extinct",
      Abs_Cover_initial > 0    & Abs_Cover_final > 0   ~ "Persistent",
      TRUE ~ NA_character_
    )
  )

data_full_CO <- site_species_CO_end %>%
  filter(!(Abs_Cover_initial == 0 & Abs_Cover_final == 0)) %>%
  mutate(
    dynamique_statut = case_when(
      Abs_Cover_initial == 0   & Abs_Cover_final > 0  ~ "Colonizer",
      Abs_Cover_final   == 0   & Abs_Cover_initial > 0 ~ "Extinct",
      Abs_Cover_initial > 0    & Abs_Cover_final > 0   ~ "Persistent",
      TRUE ~ NA_character_
    )
  )

data_full_CD <- site_species_CD_end %>%
  filter(!(Abs_Cover_initial == 0 & Abs_Cover_final == 0)) %>%
  mutate(
    dynamique_statut = case_when(
      Abs_Cover_initial == 0   & Abs_Cover_final > 0  ~ "Colonizer",
      Abs_Cover_final   == 0   & Abs_Cover_initial > 0 ~ "Extinct",
      Abs_Cover_initial > 0    & Abs_Cover_final > 0   ~ "Persistent",
      TRUE ~ NA_character_
    )
  )

# Combiner les trois traitements
data_full <- bind_rows(
  data_full_warm %>% mutate(Treatment = "warm"),
  data_full_CO   %>% mutate(Treatment = "Control_origin"),
  data_full_CD   %>% mutate(Treatment = "Control_dest")
) %>%
  filter(!is.na(dynamique_statut))

# Vérifier les effectifs
data_full %>% count(dynamique_statut, Treatment)

# ###############################################
# site_species_warm_end_clean <- site_species_warm_end %>%
#   filter(!(Abs_Cover_initial == 0 & Abs_Cover_final == 0)) %>%
#   filter(!(Abs_Cover_initial == 0 & Abs_Cover_final != 0)) #remove immigration
# 
# site_species_CO_end_clean <- site_species_CO_end %>%
#   filter(!(Abs_Cover_initial == 0 & Abs_Cover_final == 0)) %>%
#    filter(!(Abs_Cover_initial == 0 & Abs_Cover_final != 0)) 
# 
# site_species_CD_end_clean <- site_species_CD_end %>%
#   filter(!(Abs_Cover_initial == 0 & Abs_Cover_final == 0)) %>%
#   filter(!(Abs_Cover_initial == 0 & Abs_Cover_final != 0)) 

#save outputs
# write.csv(site_species_warm_end_clean, "./intermediate data/site_species_warm_end_clean_abs_abund_cover_initial_final.csv", row.names = FALSE)
# write.csv(site_species_CO_end_clean, "./intermediate data/site_species_CO_end_clean_abs_abund_cover_initial_final.csv", row.names = FALSE)
# write.csv(site_species_CD_end_clean, "./intermediate data/site_species_CD_end_clean_abs_abund_cover_initial_final.csv", row.names = FALSE)
# 

#save outputs
write.csv(data_full, "./intermediate data/site_species_data_full_end_clean_abs_abund_cover_initial_final.csv", row.names = FALSE)
```


